---
title: "Area-level SAE model"
subtitle: "A Bayesian spatio-temporal extension of the Fay--Herriot model"
---

<!-- ::: {.callout-note title="Theory in brief"}
The area-level model treats each direct estimate $\hat{\mu}^{w}_{jt}$ as a noisy observation of a true value $\mu_{jt}$:

$$
\begin{aligned}
\hat{\mu}^{w}_{jt} &= \mu_{jt} + e_{jt}, \qquad
e_{jt} \sim \mathcal{N}(0, \psi_{jt}), \\
&\quad \mu_{jt} = \alpha + x_{jt}^{T}\beta + u_j + v_t.
\end{aligned}
$$

Here, $\psi_{jt}$ is the direct-estimation sampling variance, $x_{jt}$ contains area-level predictors, $u_j$ is an area effect, and $v_t$ is a time effect. The fitted model produces the area-time estimate by combining the direct estimate, its sampling variance, and the model prediction:

$$
\hat{\mu}_{jt}
=
\widehat{\mathbb{E}}(\mu_{jt} \mid \hat{\mu}^{w}_{jt}, \psi_{jt}, x_{jt}).
$$

::: -->

```{r}
#| label: fh-setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(INLA)
library(tibble)
library(purrr)
library(scales)

base_path <- here::here("data")

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = MSOA21CD) |>
  filter(str_starts(area_id, "W")) |>
  arrange(area_id)

domain_frame <- read_csv(
  file.path(base_path, "clean", "domain_msoayear.csv"),
  show_col_types = FALSE
)

spatial_graph_file <- file.path(base_path, "clean", "msoa_queen.adj")
```

```{r}
#| label: fh-plot-config
#| echo: false
variance_label <- label_number(accuracy = 0.0001)
missing_fill <- "grey85"
diverging_low <- "#2166ac"
diverging_high <- "#b2182b"
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")

map_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey30"),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

chart_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)

map_annotation_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

Area-level SAE uses direct estimates as the modelling input. In the Fay--Herriot area-level model, each direct estimate for an *MSOA-year* cell is treated as a noisy observation of the underlying frequent-cycling rate, with its sampling variance determining how much weight the model gives to that estimate. The model borrows strength across MSOAs and survey years through spatial and temporal random effects, while area-level predictors, obtained from aggregated benchmarked sources such as [ONS UK Census 2021](https://www.ons.gov.uk/census), help explain systematic differences in the frequent cycling estimand across areas.

## Prepare modelling data

We start from generated direct estimates and raw area predictors, which will be standardised for model fitting.

```{r}
#| label: fh-load-direct

direct_bike <- read_csv(
  file.path(base_path, "freqcyc_estimates_direct.csv"),
  show_col_types = FALSE
) |>
  select(area_id, msoa_name, time_id, direct_estimate, variance, n)

head(direct_bike)
```

```{r}
#| label: fh-load-predictors

area_predictors <- read_csv(
  file.path(base_path, "clean", "areapred_freqcyc.csv"),
  show_col_types = FALSE
) |>
  select(area_id, age_adult_mean, caraccess_adult_share, pop_density) |>
  mutate(
    across(
      c(age_adult_mean, caraccess_adult_share, pop_density),
      ~ as.numeric(scale(.x)),
      .names = "z_{.col}"
    )
  )

head(area_predictors)
```

The missing or zero sampling variances of direct estimates need to be handled, because this controls how much weight each direct estimate receives in the area-level model.

- For area-time cells without a direct estimate or variance, we assign the maximum observed positive sampling variance, signalling high uncertainty. Their estimates will be determined by the model predictions.
- Because the outcome is binary (`1` for cycling frequently, `0` otherwise), very small samples can produce a variance of `0` when all sampled respondents give the same response. We also assign these cells the maximum observed positive sampling variance.

```{r}
#| label: fh-variance
#| code-fold: false

max_variance <- max(direct_bike$variance[direct_bike$variance > 0], na.rm = TRUE)

direct_bike <- direct_bike |>
  mutate(
    adj_variance = case_when(
      is.na(direct_estimate) ~ max_variance,
      is.na(variance) ~ max_variance,
      variance <= 1e-8 ~ max_variance,
      TRUE ~ variance
    )
  )
```

Finally, we assemble the modelling data for `INLA` by combining the direct estimates with the area predictors and the imported numeric indices for the temporal and spatial effects.

```{r}
#| label: fh-model-data
#| code-fold: false

area_index <- domain_frame |>
  distinct(area_id, area_inla_id)

time_index <- domain_frame |>
  distinct(time_id, time_inla_id)

model_data <- direct_bike |>
  left_join(area_predictors, by = "area_id") |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  arrange(time_inla_id, area_inla_id)

head(model_data)
```

## Fit and evaluate models

Each model uses the direct estimate as the response. The sampling variance from the direct estimation is passed to `INLA` as known observation-level uncertainty, so cells with noisy direct estimates receive less weight. We define these model specifications and fit each one to the direct estimates using `INLA`.

| Model | Name | Specifications |
|---|---|---|
| M1 | +TempSpatial | Temporal and spatial effect |
| M2 | +AvgAge | M1 plus mean adult age |
| M3 | +PopdensCaraccess | M2 plus population density and share of adults with car access |

The choice of spatial and temporal models should reflect assumptions about how residual variation is structured across space and time.

| Temporal model | Description |
|---|---|
| `rw1` | First-order random walk: smooths neighbouring time periods. |
| `rw2` | Second-order random walk: smooths changes in the trend over longer time series. |
| `ar1` | Autoregressive effect: assumes correlation decays with time distance, so closer years are more similar than distant years. |

| Spatial model | Description |
|---|---|
| `besag` | Smooths estimates across neighbouring areas, capturing spatially structured residual variation. |
| `bym2` | Updated BYM model with a mixing parameter, `phi`, which indicates how much residual variation is spatially structured. |

With only three survey years, `rw1` is suitable because it smooths neighbouring waves. Spatially, `bym2` is appropriate because it allows the model to borrow strength from neighbouring MSOAs while still capturing unstructured local variability.

```{r}
#| label: fh-specs
#| code-fold: false

m1_formula <- direct_estimate ~
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  ) +
  f(time_inla_id, model = "rw1", constr = TRUE)

m2_formula <- update(m1_formula, . ~ . + z_age_adult_mean)

m3_formula <- update(m2_formula, . ~ . +
  z_pop_density +
  z_caraccess_adult_share
)

model_formulas <- list(
  "M1 +TempSpatial" = m1_formula,
  "M2 +AvgAge" = m2_formula,
  "M3 +PopdensCaraccess" = m3_formula
)
```

```{r}
#| label: fh-fit
#| code-fold: false

fit_fh_model <- function(formula) {
  inla(
    formula = formula,
    family = "gaussian",
    data = model_data,
    scale = 1 / model_data$adj_variance,
    control.family = list(
      hyper = list(
        prec = list(initial = 0, fixed = TRUE)
      )
    ),
    control.predictor = list(compute = TRUE),
    control.compute = list(waic = TRUE, config = TRUE)
  )
}

fh_fits <- lapply(model_formulas, fit_fh_model)
```

:::{.callout-note}
When fitting area-level SAE models with `INLA`, these arguments are needed:

- `control.predictor = list(compute = TRUE)` returns fitted values for every row. In an area-level model, these are the final estimates for each area-time row.
- `scale` passes the inverse sampling variance to `INLA`, so less precise direct estimates get smoothed more.
- `control.family` fixes the Gaussian residual precision because the direct-estimate variances already provide the uncertainty.
- `control.compute = list(waic = TRUE, config = TRUE)` requests WAIC for comparing model fit and stores the configuration needed for posterior sampling.
:::

Model fit is compared with `WAIC` (Widely Applicable Information Criterion), a global predictive-fit measure returned by `INLA`. Lower values indicate better expected prediction after penalising model complexity.

```{r}
#| label: fh-waic
model_waic <- imap_dfr(
  fh_fits,
  \(fit, model_name) {
    tibble(
      model = model_name,
      waic = fit$waic$waic,
      eff_params = fit$waic$p.eff
    )
  }
) |>
  arrange(waic)
model_waic

# Best model selection
fh_best_model <- model_waic |>
  slice(1) |>
  pull(model)
fh_selected_fit <- fh_fits[[fh_best_model]]
cat("Best fitted model spec:", fh_best_model)
```

Based on this comparison, M2, which includes average MSOA age, performs best. This suggests that differences in the age profile of MSOAs explain much of the variation in the direct estimates, and that adding car access or population density provides little additional improvement in model fit.

## Estimation with posterior draws

Posterior draws from the selected `INLA` fit give area-level SAE estimates for each MSOA-year cell. The fitted-value draws are summarised into the posterior mean, variance, and 95% credible interval.

:::{.callout-note}
`INLA` stores the fitted values inside each posterior draw's latent vector. The fitted-value rows are named `Predictor...`.
:::

```{r}
#| label: fh-draws
#| code-fold: false

set.seed(100)
posterior_draws <- inla.posterior.sample(
  n = 300,
  result = fh_selected_fit
)
```

```{r}
#| label: fh-predict
#| code-fold: false

prediction_draws <- sapply(posterior_draws, \(draw) {
  prediction_rows <- startsWith(rownames(draw$latent), "Predictor")
  as.numeric(draw$latent[prediction_rows, 1])
})
```

The matrix `prediction_draws` has one MSOA-year estimate per row and one posterior draw per column. Posterior means, variances, and 95% credible intervals are computed directly from this draw matrix.

```{r}
#| label: fh-results
#| code-fold: false

draw_quantile <- \(p) {
  apply(prediction_draws, 1, quantile, probs = p, names = FALSE)
}

final_fh_results <- model_data |>
  select(area_id, time_id) |>
  bind_cols(
    tibble(
      fh_estimate = rowMeans(prediction_draws),
      fh_variance = apply(prediction_draws, 1, var),
      fh_lower = draw_quantile(0.025),
      fh_upper = draw_quantile(0.975)
    )
  ) |>
  mutate(
    fh_ci_width = fh_upper - fh_lower
  )

head(final_fh_results)
```

```{r}
#| label: fh-map
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

final_fh_map_data <- msoa_boundaries |>
  left_join(final_fh_results, by = "area_id")
fh_estimate_limits <- range(final_fh_results$fh_estimate, na.rm = TRUE)
fh_ci_width_limits <- c(0, max(final_fh_results$fh_ci_width, na.rm = TRUE))
fh_wave_years <- sort(unique(final_fh_results$time_id))

plot_fh_estimate <- function(wave) {
  final_fh_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = fh_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = fh_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_fh_ci_width <- function(wave) {
  final_fh_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = fh_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = fh_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

fh_estimate_maps <- lapply(fh_wave_years, plot_fh_estimate)
fh_uncertainty_maps <- lapply(fh_wave_years, plot_fh_ci_width)

fh_estimate_maps[[1]] <- fh_estimate_maps[[1]] + ylab("Final estimate")
fh_uncertainty_maps[[1]] <- fh_uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(fh_wave_years) - 1)) {
  fh_estimate_maps[[i]] <- fh_estimate_maps[[i]] +
    theme(legend.position = "none")
  fh_uncertainty_maps[[i]] <- fh_uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(fh_estimate_maps, fh_uncertainty_maps), ncol = length(fh_wave_years)) +
  plot_annotation(
    title = "Share of adults who cycle for transport frequently",
    subtitle = paste0("Selected model (", fh_best_model, ") and 95% CI width"),
    theme = map_annotation_theme
  )
```

The area-level SAE estimates are less jagged across neighbouring areas and show clearer broad regional structure than the direct estimates. In sparsely sampled MSOA-year cells, the FH estimate depends heavily on area-level predictors and random effects because the direct estimate contributes limited local information. The model therefore cannot use within-area respondent variation beyond what is already captured in the direct estimate. [**Unit-level models**](unit-sae-summ.qmd) address this by modelling individual outcomes with respondent-level predictors.

## Comparative analysis

```{r}
#| label: fh-collate

summarise_fh_draws <- function(fit, model_name, n_draws = 250) {
  posterior_draws <- inla.posterior.sample(
    n = n_draws,
    result = fit
  )

  prediction_draws <- sapply(posterior_draws, \(draw) {
    prediction_rows <- startsWith(rownames(draw$latent), "Predictor")
    as.numeric(draw$latent[prediction_rows, 1])
  })

  draw_quantile <- \(p) {
    apply(prediction_draws, 1, quantile, probs = p, names = FALSE)
  }

  bind_cols(
    model_data |>
      select(area_id, time_id, msoa_name, direct_estimate, variance, n),
    tibble(
      fh_estimate = rowMeans(prediction_draws),
      fh_variance = apply(prediction_draws, 1, var),
      fh_lower = draw_quantile(0.025),
      fh_upper = draw_quantile(0.975)
    )
  ) |>
    transmute(
      model = model_name,
      area_id,
      time_id,
      msoa_name,
      direct_estimate,
      variance,
      n,
      fh_estimate,
      fh_variance,
      fh_lower,
      fh_upper,
      fh_ci_width = fh_upper - fh_lower
    )
}

all_fh_results <- imap_dfr(
  fh_fits,
  \(fit, model_name) summarise_fh_draws(fit, model_name)
) |>
  mutate(model = factor(model, levels = names(model_formulas))) |>
  pivot_wider(
    id_cols = c(area_id, time_id, msoa_name, direct_estimate, variance, n),
    names_from = model,
    values_from = c(fh_estimate, fh_variance, fh_lower, fh_upper, fh_ci_width),
    names_sep = "__"
  )

head(all_fh_results)
```

We can compare how the direct estimate for each MSOA is handled by the baseline area-level model (M1) with only the temporal-spatial random effects and no predictors to understand area-level smoothing. Area-time cells with high sampling variance contain less reliable local information, so M1 pulls them more strongly toward the temporal-spatial model prediction. 

```{r}
#| label: fh-baseline-data

direct_to_m1 <- all_fh_results |>
  transmute(
    area_id,
    time_id,
    direct_variance = variance,
    variance_change = .data[["fh_variance__M1 +TempSpatial"]] - variance
  )

wave_years <- sort(unique(all_fh_results$time_id))
```

```{r}
#| label: fh-baseline-plot
#| echo: false
#| fig-width: 16
#| fig-height: 12
#| out-width: "100%"

variance_change_limit <- max(abs(direct_to_m1$variance_change), na.rm = TRUE) * .5
direct_variance_limit <- max(direct_to_m1$direct_variance, na.rm = TRUE)

direct_to_m1_map_data <- msoa_boundaries |>
  left_join(direct_to_m1, by = "area_id")

plot_m1_variance_change <- function(wave) {
  direct_to_m1_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = variance_change), colour = "transparent") +
    scale_fill_gradient2(
      low = diverging_low,
      mid = "white",
      high = diverging_high,
      midpoint = 0,
      limits = c(-variance_change_limit, variance_change_limit),
      oob = squish,
      labels = variance_label,
      na.value = missing_fill,
      name = "Change"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_direct_variance <- function(wave) {
  direct_to_m1_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = direct_variance), colour = "transparent") +
    scale_fill_distiller(
      palette = "Reds",
      direction = 1,
      limits = c(0, direct_variance_limit),
      labels = variance_label,
      na.value = missing_fill,
      name = "Variance"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

variance_change_maps <- lapply(wave_years, plot_m1_variance_change)
direct_variance_maps <- lapply(wave_years, plot_direct_variance)

variance_change_maps[[1]] <- variance_change_maps[[1]] +
  ylab("Variance change")
direct_variance_maps[[1]] <- direct_variance_maps[[1]] +
  ylab("Direct variance")

for (i in seq_len(length(wave_years) - 1)) {
  variance_change_maps[[i]] <- variance_change_maps[[i]] +
    theme(legend.position = "none")
  direct_variance_maps[[i]] <- direct_variance_maps[[i]] +
    theme(legend.position = "none")
}

direct_to_m1_ranked <- direct_to_m1 |>
  drop_na(variance_change) |>
  group_by(time_id) |>
  arrange(direct_variance, .by_group = TRUE) |>
  mutate(variance_rank = row_number()) |>
  ungroup() |>
  select(
    time_id,
    variance_rank,
    `Variance change` = variance_change,
    `Variance (direct)` = direct_variance
  ) |>
  pivot_longer(
    cols = -c(time_id, variance_rank),
    names_to = "measure",
    values_to = "value"
  )

direct_to_m1_ranked_plot <- ggplot(
  direct_to_m1_ranked,
  aes(
    x = variance_rank,
    y = value,
    colour = measure
  )
) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line(alpha = 0.8) +
  facet_wrap(vars(time_id), nrow = 1) +
  scale_y_continuous(labels = variance_label) +
  scale_colour_manual(values = c(
    "Variance change" = "purple",
    "Variance (direct)" = "red"
  )) +
  labs(
    title = "",
    x = "MSOAs sorted by variance",
    y = "Variance",
    colour = NULL
  ) +
  theme_minimal() +
  chart_text_theme

(wrap_plots(c(variance_change_maps, direct_variance_maps), ncol = 3) /
  direct_to_m1_ranked_plot) +
  plot_layout(heights = c(2, 1)) +
  plot_annotation(
    title = "Direct estimates compared with area-level model (M1)",
    subtitle = "M1 stabilises noisy direct estimates with temporal-spatial smoothing",
    theme = map_annotation_theme
  )
```

The `INLA` fit also produces posterior estimates for cells with missing direct estimates or zero/missing variance, allowing sparse or unsampled area-time cells to remain in the output with explicit uncertainty.

```{r}
#| label: fh-synthetic-est

cat("Missing direct estimates / variances")
all_fh_results |>
  filter(is.na(direct_estimate) | is.na(variance) )|>
  transmute(
    `MSOA` = area_id,
    `Wave year` = time_id,
    `Direct estimate` = share_label(direct_estimate),
    `Sampling variance` = round(variance, 6),
    `M1 estimate` = share_label(.data[["fh_estimate__M1 +TempSpatial"]]),
    `M1 variance` = round(
      .data[["fh_variance__M1 +TempSpatial"]],
      6
    )
  )

cat("Zero variances")
all_fh_results |>
  filter(variance < 1e-8) |>
  transmute(
    `MSOA` = area_id,
    `Wave year` = time_id,
    `Direct estimate` = share_label(direct_estimate),
    `Sampling variance` = round(variance, 6),
    `M1 estimate` = share_label(.data[["fh_estimate__M1 +TempSpatial"]]),
    `M1 variance` = round(
      .data[["fh_variance__M1 +TempSpatial"]],
      6
    )
  ) |>
  head(5)
```

For *spatial and temporal random effects*, `INLA` also estimates hyperparameters that show residual variation and how that remaining variation is structured across time and space.

- `tau` is the precision of a random effect, or the inverse of its variance. Higher precision means less residual variation in that component. A higher `tau` for the spatial term indicates smaller area differences between neighbouring MSOAs. Temporal term's `tau` in comparison is moderate.

- `phi` is the BYM2 mixing parameter for the area random effect. It describes how much of that residual area effect is spatially structured. A lower `phi` here (less than `0.5`) means the remaining area differences are mostly local and less strongly explained by adjacency. 

```{r}
#| label: fh-hyper

fh_fits[["M1 +TempSpatial"]]$summary.hyperpar |>
  as_tibble(rownames = "parameter") |>
  filter(str_detect(parameter, "time_inla_id|area_inla_id|Phi")) |>
  select(parameter, mean, sd, `0.025quant`, `0.5quant`, `0.975quant`) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```
The temporal term captures moderate wave-to-wave variation, keeping differences between survey years explicit instead of pooling the three waves into a single period, so its inclusion is integral. The spatial term, on the other hand, appears to add little to model fit. This may reflect sparse or missing direct estimates in many areas, making the spatial structure not especially informative to the model.

So far, model choice has favoured a parsimonious specification, but area-level predictor inclusion can also be assessed by looking at how estimates change locally and by using domain knowledge about the outcome. Thsi is because adding predictors changes the model prediction for each area-time cell, and unstable or missing direct estimates are further pulled towards these updated predictions. These small local changes may not produce a large improvement in global fit, but they can still show substantive plausibility.

For example, we compare M1 with M3, where M3 adds average age, population density, and car access as predictors.

```{r}
#| label: fh-predictor-change
predictor_change <- all_fh_results |>
  transmute(
    area_id,
    estimate_change =
      .data[["fh_estimate__M3 +PopdensCaraccess"]] -
      .data[["fh_estimate__M1 +TempSpatial"]]
  ) |>
  distinct(area_id, .keep_all = TRUE)

model_data |>
  distinct(area_id, .keep_all = TRUE) |>
  left_join(predictor_change, by = "area_id") |>
  mutate(
    lower_cutoff = quantile(estimate_change, 0.1, na.rm = TRUE),
    upper_cutoff = quantile(estimate_change, 0.9, na.rm = TRUE),
    change_group = case_when(
      estimate_change >= upper_cutoff ~ "Raised most when adding 3 predictors",
      estimate_change <= lower_cutoff ~ "Lowered most when adding 3 predictors"
    )
  ) |>
  drop_na(change_group) |>
  summarise(
    `Mean adult age (z)` = mean(z_age_adult_mean, na.rm = TRUE),
    `Population density (z)` = mean(z_pop_density, na.rm = TRUE),
    `Car-access share (z)` = mean(z_caraccess_adult_share, na.rm = TRUE),
    .by = change_group
  )
```
```{r}
#| label: fh-map-change
#| echo: false
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

predictor_change_estimate_limit <- quantile(
  abs(predictor_change$estimate_change),
  0.95,
  na.rm = TRUE,
  names = FALSE
)

msoa_boundaries |>
  left_join(predictor_change, by = "area_id") |>
  ggplot() +
  geom_sf(aes(fill = estimate_change), colour = "transparent") +
  scale_fill_gradient2(
    low = diverging_low,
    mid = "white",
    high = diverging_high,
    midpoint = 0,
    limits = c(-predictor_change_estimate_limit, predictor_change_estimate_limit),
    oob = squish,
    labels = pp_label,
    na.value = missing_fill,
    name = "Change"
  ) +
  labs(
    title = "Estimate change after adding predictors",
    subtitle = "M3 minus M1"
  ) +
  theme_void() +
  map_text_theme
```
```{r}
#| label: fh-m3-fixed
fh_fits[["M3 +PopdensCaraccess"]]$summary.fixed |>
  as_tibble(rownames = "term") |>
  transmute(
    term,
    mean,
    lower = `0.025quant`,
    upper = `0.975quant`
  )
```

The M3 fixed-effect summary suggests that MSOAs with older populations, lower population density, and a higher share of adults with car access tend to have lower rates of frequent cycling. This supports considering these predictors, but with the caveat that their contribution should be judged alongside model fit diagnostics and local estimate changes.

```{r}
#| label: fh-export
#| code-fold: true
write_csv(final_fh_results, file.path(base_path, "freqcyc_estimates_fh.csv"))
```

```{r}
#| label: fh-area-effects-export
#| include: false
#| echo: false
#| message: false
#| warning: false

besag_model_data <- model_data |>
  mutate(
    area_struct_id = area_inla_id,
    area_unstruct_id = area_inla_id
  )

fh_besag_formula <- direct_estimate ~
  f(
    area_struct_id,
    model = "besag",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  ) +
  f(area_unstruct_id, model = "iid") +
  f(time_inla_id, model = "rw1", constr = TRUE)

if (fh_best_model %in% c("M2 +AvgAge", "M3 +PopdensCaraccess")) {
  fh_besag_formula <- update(fh_besag_formula, . ~ . + z_age_adult_mean)
}

if (fh_best_model == "M3 +PopdensCaraccess") {
  fh_besag_formula <- update(fh_besag_formula, . ~ . +
    z_pop_density +
    z_caraccess_adult_share
  )
}

fh_besag_fit <- inla(
  formula = fh_besag_formula,
  family = "gaussian",
  data = besag_model_data,
  scale = 1 / besag_model_data$adj_variance,
  control.family = list(
    hyper = list(
      prec = list(initial = 0, fixed = TRUE)
    )
  ),
)

fh_iid <- bind_rows(
  fh_besag_fit$summary.random$area_struct_id |>
    as_tibble() |>
    transmute(
      area_inla_id = ID,
      component = "Structured",
      effect = mean,
      lower = `0.025quant`,
      upper = `0.975quant`
    ),
  fh_besag_fit$summary.random$area_unstruct_id |>
    as_tibble() |>
    transmute(
      area_inla_id = ID,
      component = "Unstructured",
      effect = mean,
      lower = `0.025quant`,
      upper = `0.975quant`
    )
) |>
  left_join(area_index, by = "area_inla_id") |>
  mutate(method = "FH") |>
  select(method, component, area_id, area_inla_id, effect, lower, upper)

write_csv(fh_iid, file.path(base_path, "freqcyc_iid_fh.csv"))
```
