---
title: "Unit-Level SAE Modelling (Battese--Harter--Fuller)"
subtitle: "A Bayesian spatio-temporal extension"
---

::: {.callout-note title="Theory in brief"}
The unit-level route models respondent outcomes directly. For the area-time target $\mu_{jt}$, a simple mixed-model form is:

$$
Y_{ijt} = \alpha + x_{ijt}^{T}\beta + u_j + v_t + \epsilon_{ijt},
$$

where $x_{ijt}$ contains unit-level predictors, $u_j$ is an area effect, and $v_t$ is a time effect. The area effect can be replaced by a structured spatial effect, and a temporal effect $v_t$ can be added to represent wave-to-wave dependence. 
To estimate the target for an area-time cell, the fitted model is evaluated using the corresponding population-level auxiliary summaries $\bar{x}_{jt}$:

$$
\hat{\mu}_{jt}^{\mathrm{BHF}}
=
\widehat{\mathbb{E}}(Y \mid j,t,\bar{x}_{jt}).
$$
:::


```{r}
#| label: bhf-setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(spdep)
library(INLA)
library(tibble)
library(purrr)
library(scales)

base_path <- here::here("data")
domain_frame <- read_csv(file.path(base_path, "clean", "domain_msoayear.csv"), show_col_types = FALSE)
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = all_of("MSOA21CD")) |>
  filter(str_starts(area_id, "W"))
```

```{r}
#| label: bhf-plot-config
#| echo: false
variance_label <- label_number(accuracy = 0.0001)
missing_fill <- "grey85"

map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

Unit-level SAE models respondent records directly instead of modelling pre-aggregated direct estimates. For our estimand, the Battese--Harter--Fuller candidate models estimate the binary frequent-cycling indicator from survey records using temporal and spatial random effects, with unit-level `age`, `sex`, and `car_access` added across model specifications.

The fitted model is then used to predict the target *MSOA-year* estimates using MSOA-level summaries of age, sex, and car access, obtained from ONS UK Census 2021. This avoids relying directly on unstable area-level direct estimates, but introduces new assumptions about how adequately area-level summaries represent the population composition.

## Prepare model and aggregation data

Before preparing the model inputs, we create consistent numeric indices for each area and survey year so that `INLA` can attach random effects to the correct area and period, as well as a Queen-contiguity adjacency graph `.adj` as input, similar to what we did in the previous FH model estimation.

```{r}
#| label: bhf-index
#| code-fold: false

area_levels <- sort(unique(domain_frame[["area_id"]]))
time_levels <- sort(unique(domain_frame[["time_id"]]))

area_index <- tibble(
  area_id = area_levels,
  area_inla_id = seq_along(area_levels)
)

time_index <- tibble(
  time_id = time_levels,
  time_inla_id = seq_along(time_levels)
)
```
```{r}
#| label: bhf-graph
#| code-fold: false

msoa_boundaries_sorted <- msoa_boundaries |>
  inner_join(area_index, by = "area_id") |>
  arrange(area_inla_id)

msoa_neighbours <- poly2nb(
  msoa_boundaries_sorted,
  queen = TRUE,
  row.names = msoa_boundaries_sorted$area_id
)

spatial_graph_file <- tempfile(fileext = ".adj")

nb2INLA(
  file = spatial_graph_file,
  nb = msoa_neighbours
)
```

We now prepare the two datasets needed for the unit-level `INLA` workflow:

- `model_data` contains the sampled respondent records used to fit the model. 
- `aggregation_data` contains one row for each area-time cell we want to estimate.

### Respondent model data

The respondent model data has one row per sampled person. For this example we use:

- `indicator` as the binary response, our target;
- `age` as a unit-level predictor;
- `female` as a binary unit-level predictor;
- `car_access` as a unit-level predictor;
- `time_id` as the survey wave year identifier;
- `area_id` as the area identifier;

```{r}
#| label: bhf-model-data

cycling_indicator <- read_csv(file.path(base_path, "clean", "surveyind_freqbike.csv"), show_col_types = FALSE) |>
  select(unit_id, time_id, area_id, indicator, age, sex, car_access) |>
  mutate(female = as.integer(sex == 2))

model_data <- cycling_indicator |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  drop_na(indicator, area_id, time_id, age, female, car_access) |>
  select(
    unit_id,
    area_id,
    time_id,
    indicator,
    age,
    female,
    car_access,
    time_inla_id,
    area_inla_id
  ) |>
  arrange(time_id, area_inla_id)

head(model_data)
```

### Area aggregation data

The aggregation input contains the complete area-time grid we want to estimate, and the area-level mean of the unit-level predictor. Since unit-level `car_access` is binary, the area mean is simply the share of households with access to at least one car or van.

```{r}
#| label: bhf-agg-data

area_predictors <- read_csv(file.path(base_path, "clean", "areapred_freqbike.csv"), show_col_types = FALSE) |>
  transmute(
    area_id,
    age = mean_age_adult,
    female = raw_prob_female,
    car_access = raw_car_access_hhshare
  )

aggregation_data <- domain_frame |>
  left_join(area_predictors, by = "area_id") |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  select(
    area_id,
    msoa_name,
    time_id,
    age,
    female,
    car_access,
    time_inla_id,
    area_inla_id
  ) |>
  arrange(time_id, area_inla_id)

head(aggregation_data)
```

::: {.callout-note}
These area means come from Census 2021 and are used as a time-stable MSOA predictor across the three survey waves. This is reasonable only if auxiliary variables are unlikely to change quickly.
:::

## Fit and evaluate models

We fit linear mixed models to respondent-level records, then use them to predict area-time means. Because the response variable is binary, the fitted values are interpreted as predicted probabilities.

We specify three unit-level models to evaluate at the next step:

| Model | Name | Specifications |
|---|---|---|
| M1 | +TempSpatial | Temporal RW1 plus BYM2 spatial MSOA effect  |
| M2 | +Demographic | M1 plus respondent age and sex |
| M3 | +CarAccess | M2 plus respondent-level car access |

```{r}
#| label: bhf-specs
#| code-fold: false

m1_formula <- indicator ~
  1 +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m2_formula <- update(m1_formula, . ~ . +
  age +
  female
)

m3_formula <- update(m2_formula, . ~ . +
  car_access
)

model_formulas <- list(
  "M1 +TempSpatial" = m1_formula,
  "M2 +Demographic" = m2_formula,
  "M3 +CarAccess" = m3_formula
)

```

Conceptually, the workflow has two steps: fit the respondent-level model, then evaluate it for each area-time. We fit the candidate models using only the sampled respondent rows. The area-time aggregation rows are used later, after model selection, when posterior draws are evaluated at the MSOA-year predictor summaries.

```{r}
#| label: bhf-fit
#| code-fold: false

fit_bhf_model <- function(formula) {
  inla(
    formula = formula,
    family = "gaussian",
    data = model_data,
    control.predictor = list(compute = TRUE),
    control.compute = list(waic = TRUE, config = TRUE)
  )
}

bhf_fits <- lapply(model_formulas, fit_bhf_model)
```

:::{.callout-note}
When fitting unit-level BHF models with `INLA`, these arguments are needed:

- `control.predictor = list(compute = TRUE)` returns fitted values for the respondent rows used in the likelihood.
- `control.compute = list(waic = TRUE, config = TRUE)` requests WAIC for comparing model fit and stores the configuration needed for posterior sampling.
:::

To compare the fitted models' overall performance, we again use `WAIC` as a first check, a global predictive-fit measure returned by `INLA`. Lower values indicate better expected prediction after penalising model complexity.

```{r}
#| label: bhf-waic

model_waic <- imap_dfr(
  bhf_fits,
  \(fit, model_name) {
    tibble(
      model = model_name,
      waic = fit$waic$waic,
      eff_params = fit$waic$p.eff
    )
  }
) |>
  arrange(waic)

bhf_best_model <- model_waic |>
  slice(1) |>
  pull(model)

bhf_selected_fit <- bhf_fits[[bhf_best_model]]

model_waic
```

The table is sorted by WAIC. The first row is used for the final BHF estimates. Here, the car-access extension has the strongest WAIC support among the BHF specifications.

A look into the fitted coefficients for the fixed effect terms sheds more light about the selected model's predictive structure.

```{r}
#| label: bhf-fixed

bhf_fits[[bhf_best_model]]$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  transmute(
    term,
    mean = pp_label(mean),
    lower = pp_label(.data[["0.025quant"]]),
    upper = pp_label(.data[["0.975quant"]])
  )
```

Finally, the **spatial and temporal random-effect** hyperparameters `tau` and `phi` help show what the temporal and spatial terms are doing in the selected model. 

```{r}
#| label: bhf-hyper

bhf_fits[[bhf_best_model]]$summary.hyperpar |>
  as_tibble(rownames = "parameter") |>
  filter(str_detect(parameter, "time_inla_id|area_inla_id|Phi")) |>
  select(parameter, mean, sd, `0.025quant`, `0.5quant`, `0.975quant`) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

The random-effect table reports the posterior summaries for the temporal and spatial components in the selected BHF specification. Higher precision means less residual variation in that component, while the BYM2 `phi` indicates how much of the area effect is spatially structured.

## Estimation and aggregation with posterior draws

The selected `INLA` fit gives posterior draws for the fixed effects and the temporal and spatial random effects. For each draw, we evaluate the fitted model at the MSOA-year aggregation rows, using area-level summaries for age, female share, and car access. The final estimate and uncertainty are summaries of those posterior predictions.

```{r}
#| label: bhf-results
#| code-fold: false

set.seed(100)

bhf_posterior_draws <- inla.posterior.sample(
  n = 300,
  result = bhf_selected_fit
)

latent_names <- rownames(bhf_posterior_draws[[1]]$latent)
fixed_terms <- rownames(bhf_selected_fit$summary.fixed)

fixed_effect_terms <- tibble(
  term = fixed_terms,
  latent_term = paste0(term, ":1")
)

area_effect_terms <- tibble(
  area_inla_id = area_index$area_inla_id,
  term = paste0("area_inla_id:", area_inla_id)
)

time_effect_terms <- tibble(
  time_inla_id = time_index$time_inla_id,
  term = paste0("time_inla_id:", time_inla_id)
)

stopifnot(all(fixed_effect_terms$latent_term %in% latent_names))
stopifnot(all(area_effect_terms$term %in% latent_names))
stopifnot(all(time_effect_terms$term %in% latent_names))

fixed_draws <- sapply(bhf_posterior_draws, \(draw) {
  values <- as.numeric(draw$latent[fixed_effect_terms$latent_term, 1])
  names(values) <- fixed_effect_terms$term
  values
})

area_draws <- sapply(bhf_posterior_draws, \(draw) {
  values <- as.numeric(draw$latent[area_effect_terms$term, 1])
  names(values) <- area_effect_terms$area_inla_id
  values
})

time_draws <- sapply(bhf_posterior_draws, \(draw) {
  values <- as.numeric(draw$latent[time_effect_terms$term, 1])
  names(values) <- time_effect_terms$time_inla_id
  values
})

bhf_draw_matrix <- matrix(
  NA_real_,
  nrow = nrow(aggregation_data),
  ncol = length(bhf_posterior_draws)
)

for (draw_id in seq_along(bhf_posterior_draws)) {
  fixed <- fixed_draws[, draw_id]
  prediction <- fixed["(Intercept)"] +
    area_draws[as.character(aggregation_data$area_inla_id), draw_id] +
    time_draws[as.character(aggregation_data$time_inla_id), draw_id]

  if ("age" %in% names(fixed)) {
    prediction <- prediction + fixed["age"] * aggregation_data$age
  }
  if ("female" %in% names(fixed)) {
    prediction <- prediction + fixed["female"] * aggregation_data$female
  }
  if ("car_access" %in% names(fixed)) {
    prediction <- prediction + fixed["car_access"] * aggregation_data$car_access
  }

  bhf_draw_matrix[, draw_id] <- pmin(pmax(prediction, 0), 1)
}

final_bhf_results <- aggregation_data |>
  select(area_id, msoa_name, time_id) |>
  bind_cols(
    tibble(
      bhf_estimate = rowMeans(bhf_draw_matrix),
      bhf_variance = apply(bhf_draw_matrix, 1, var),
      bhf_lower = apply(bhf_draw_matrix, 1, quantile, probs = 0.025),
      bhf_upper = apply(bhf_draw_matrix, 1, quantile, probs = 0.975)
    )
  ) |>
  mutate(
    bhf_ci_width = bhf_upper - bhf_lower
  )

head(final_bhf_results)
```

```{r}
#| label: bhf-map
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

final_bhf_map_data <- msoa_boundaries |>
  left_join(final_bhf_results, by = "area_id")
bhf_estimate_limits <- range(final_bhf_results$bhf_estimate, na.rm = TRUE)
bhf_ci_width_limits <- c(0, max(final_bhf_results$bhf_ci_width, na.rm = TRUE))
bhf_wave_years <- sort(unique(final_bhf_results$time_id))

plot_bhf_estimate <- function(wave) {
  final_bhf_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = bhf_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = bhf_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_bhf_ci_width <- function(wave) {
  final_bhf_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = bhf_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = bhf_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(bhf_wave_years, plot_bhf_estimate)
uncertainty_maps <- lapply(bhf_wave_years, plot_bhf_ci_width)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(bhf_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = length(bhf_wave_years)) +
  plot_annotation(
    title = "Share of adults who cycle for transport frequently",
    subtitle = paste0("Selected BHF model (", bhf_best_model, ") and 95% CI width"),
    theme = map_annotation_theme
  )
```

## Comparative analysis

Compared with the mapped direct estimates, the modelled estimates show a more plausible spatial pattern: more urban areas in south Wales have higher rates of frequent cycling among adults, while the sporadic jaggedness seen in low-sample regions is smoothed. The scatterplot below shows that the BHF estimates are in fact much smoother than the direct estimates and closer to the Fay--Herriot estimates, as expected from a model-based approach that borrows strength across area-time cells.

```{r}
#| label: bhf-load-comp

direct_estimates <- read_csv(file.path(base_path, "freqbike_estimates_direct.csv"), show_col_types = FALSE) |>
  select(area_id, time_id, n, direct_estimate, variance)

fh_final_estimates <- read_csv(file.path(base_path, "freqbike_estimates_fh.csv"), show_col_types = FALSE) |>
  transmute(
    area_id,
    time_id,
    fh_estimate
  )

bhf_estimates <- final_bhf_results |>
  left_join(fh_final_estimates, by = c("area_id", "time_id")) |>
  left_join(direct_estimates, by = c("area_id", "time_id")) |>
  relocate(direct_estimate, n, .after = time_id)
```
```{r}
#| label: bhf-plot-comp
#| echo: false
#| fig-height: 6
#| fig-width: 12

bhf_scatter_comparison <- bhf_estimates |>
  select(
    area_id,
    time_id,
    n,
    bhf_estimate,
    direct_estimate,
    fh_estimate
  ) |>
  pivot_longer(
    cols = c(direct_estimate, fh_estimate),
    names_to = "comparison",
    values_to = "comparison_estimate"
  ) |>
  mutate(
    comparison = recode(
      comparison,
      direct_estimate = "Direct estimates",
      fh_estimate = "Fay-Herriot estimates"
    )
  ) |>
  drop_na(comparison_estimate)

bhf_scatter_x_limits <- c(0, max(bhf_scatter_comparison$comparison_estimate, na.rm = TRUE))
bhf_scatter_y_limits <- c(0, max(bhf_scatter_comparison$bhf_estimate, na.rm = TRUE))

bhf_scatter_comparison |>
  ggplot(aes(x = comparison_estimate, y = bhf_estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(aes(size = n), alpha = 0.55, colour = "#1f4e5f") +
  facet_wrap(vars(comparison), nrow = 1) +
  scale_x_continuous(labels = share_label) +
  scale_y_continuous(labels = share_label) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  coord_cartesian(
    xlim = bhf_scatter_x_limits,
    ylim = bhf_scatter_y_limits
  ) +
  labs(
    title = "Estimate comparison",
    subtitle = "Dashed line shows equality; points are area-time cells",
    x = "Comparison estimate",
    y = paste0("BHF estimate (", bhf_best_model, ")")
  ) +
  theme_minimal(base_size = 12) +
  chart_text_theme +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

However, the BHF and Fay--Herriot estimates for each small area can still differ because the two models use the survey information differently. The BHF model is fitted to unweighted respondent records, while the Fay--Herriot model starts from weighted direct estimates and their sampling variances to model area-year estimates. As a result, unstable direct estimates can affect the Fay--Herriot results, whereas the BHF model may be inaccurate if the complex survey design and weights are ignored. In general, for scenarios where rich survey microdata is present, a unit-level approach is preferred.

That said, although the model is fitted with unit-level predictors, the aggregation step relies on strong assumptions about each area’s population composition. Representing age, sex, and car access with area-level summaries, such as mean age, proportion female, and share of households with car access, can hide important compositional differences. Two areas may have similar summaries but very different age-sex-car structures, meaning the model could produce similar estimates even where cycling behaviour differs across specific combinations of these predictors.

This limitation can be addressed by using a poststratification table that contains population counts for the joint distribution of the relevant predictors. This is the logic of **Multilevel Regression with Poststratification (MRP)**, predict the outcome for each poststratification cell, then aggregate those predictions using the corresponding population counts so the final estimates better reflect the target population composition.

```{r}
#| label: bhf-export
#| code-fold: true

write_csv(final_bhf_results, file.path(base_path, "freqbike_estimates_bhf.csv"))
```
