---
title: "Unit-Level SAE Modelling (Battese--Harter--Fuller)"
subtitle: "A Bayesian spatio-temporal extension"
---

::: {.callout-note title="Theory in brief"}
The unit-level route models respondent outcomes directly. For the area-time target $\mu_{jt}$, a simple mixed-model form is:

$$
Y_{ijt} = \alpha + x_{ijt}^{T}\beta + u_j + v_t + \epsilon_{ijt},
$$

where $x_{ijt}$ contains unit-level predictors, $u_j$ is an area effect, and $v_t$ is a time effect. The area effect can be replaced by a structured spatial effect, and a temporal effect $v_t$ can be added to represent wave-to-wave dependence. 
To estimate the target for an area-time cell, the fitted model is evaluated using the corresponding population-level auxiliary summaries $\bar{x}_{jt}$:

$$
\hat{\mu}_{jt}^{\mathrm{BHF}}
=
\widehat{\mathbb{E}}(Y \mid j,t,\bar{x}_{jt}).
$$
:::


```{r}
#| label: bhf-setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(INLA)
library(tibble)
library(purrr)
library(scales)

base_path <- here::here("data")
domain_frame <- read_csv(file.path(base_path, "clean", "domain_msoayear.csv"), show_col_types = FALSE)
spatial_graph_file <- file.path(base_path, "clean", "msoa_queen.adj")
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = all_of("MSOA21CD")) |>
  filter(str_starts(area_id, "W"))
```

```{r}
#| label: bhf-plot-config
#| echo: false
variance_label <- label_number(accuracy = 0.0001)
missing_fill <- "grey85"

map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

Unit-level SAE models respondent records directly instead of starting from pre-aggregated direct estimates, which helps when direct estimates are too sparse or unstable to provide reliable modelling inputs. Here, the binary frequent-cycling indicator is modelled from survey records using temporal and spatial random effects, with unit-level `age`, `sex`, and `car_access` added across model specifications.

The approach follows the Battese--Harter--Fuller nested-error regression model, where the individual outcome is modelled using predictors present in the survey microdata. It is then evaluated using MSOA-level summaries of the predictors extracted from [ONS UK Census 2021](https://www.ons.gov.uk/census) summaries to produce the final *MSOA-year* estimates. This reduces reliance on unstable or missing direct estimates, but it introduces a different assumption that the area-level summaries of `age`, `sex`, and `car_access` must adequately represent the target population composition in each *MSOA-year* cell.

## Prepare model and aggregation data

There are two inputs when performing unit-level SAE:

- `model_data` contains the sampled respondent records used to fit the model.
- `aggregation_data` contains one row for each area-time cell we want to estimate.


```{r}
#| label: bhf-model-data

survey_microdata <- read_csv(
  file.path(base_path, "clean", "surveyind_freqbike.csv"),
  show_col_types = FALSE
)

model_data <- survey_microdata |>
  transmute(
    unit_id,
    area_id,
    time_id,
    indicator,
    age,
    female = as.integer(sex == 2),
    car_access
  )

head(model_data)
```


```{r}
#| label: bhf-agg-data

area_predictors <- read_csv(file.path(base_path, "clean", "areapred_freqbike.csv"), show_col_types = FALSE) |>
  transmute(
    area_id,
    age = mean_age_adult,
    female = raw_prob_female,
    car_access = raw_car_access_hhshare
  )

aggregation_data <- domain_frame |>
  left_join(area_predictors, by = "area_id") |>
  select(
    area_id,
    msoa_name,
    time_id,
    age,
    female,
    car_access
  ) |>
  arrange(time_id, area_id)

head(aggregation_data)
```

::: {.callout-note}
The aggregation data contains the complete area-time grid we want to estimate, and the area-level mean of the unit-level predictors. These area means come from Census 2021 and are used as a time-stable MSOA predictor across the three survey waves. This is reasonable only if auxiliary variables are unlikely to change quickly.
:::

Before fitting the model, we set `INLA` time and area indices to both all the datasets we have prepared.

```{r}
#| label: bhf-add-ids
#| code-fold: false

inla_domain_index <- domain_frame |>
  select(area_id, time_id, area_inla_id, time_inla_id)

area_ids <- sort(unique(inla_domain_index$area_inla_id))
time_ids <- sort(unique(inla_domain_index$time_inla_id))

model_data <- model_data |>
  left_join(inla_domain_index, by = c("area_id", "time_id")) |>
  drop_na(indicator, age, female, car_access, area_inla_id, time_inla_id) |>
  arrange(time_inla_id, area_inla_id)

aggregation_data <- aggregation_data |>
  left_join(inla_domain_index, by = c("area_id", "time_id")) |>
  drop_na(age, female, car_access, area_inla_id, time_inla_id) |>
  arrange(time_inla_id, area_inla_id)
```

## Fit and evaluate models

We fit binomial unit-level models to respondent records, then use them to predict area-time means. We specify three unit-level models to evaluate at the next step:

| Model | Name | Specifications |
|---|---|---|
| M1 | +TempSpatial | Temporal RW1 plus BYM2 spatial MSOA effect  |
| M2 | +Demographic | M1 plus respondent age and sex |
| M3 | +CarAccess | M2 plus respondent-level car access |

```{r}
#| label: bhf-specs
#| code-fold: false

m1_formula <- indicator ~
  1 +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m2_formula <- update(m1_formula, . ~ . +
  age +
  female
)

m3_formula <- update(m2_formula, . ~ . +
  car_access
)

model_formulas <- list(
  "M1 +TempSpatial" = m1_formula,
  "M2 +Demographic" = m2_formula,
  "M3 +CarAccess" = m3_formula
)

```

```{r}
#| label: bhf-fit
#| code-fold: false

fit_bhf_model <- function(formula) {
  inla(
    formula = formula,
    family = "binomial",
    Ntrials = 1,
    data = model_data,
    control.predictor = list(compute = TRUE),
    control.compute = list(waic = TRUE, config = TRUE)
  )
}

bhf_fits <- lapply(model_formulas, fit_bhf_model)
```

:::{.callout-note}
When fitting unit-level BHF models with `INLA`, these arguments are needed:

- `control.predictor = list(compute = TRUE)` returns fitted values for the respondent rows used in the likelihood.
- `control.compute = list(waic = TRUE, config = TRUE)` requests WAIC for comparing model fit and stores the configuration needed for posterior sampling.
:::

To compare the fitted models' overall performance, we again use `WAIC` as a first check. Here, the car-access extension has the strongest WAIC support among the model specifications.
```{r}
#| label: bhf-waic

model_waic <- imap_dfr(
  bhf_fits,
  \(fit, model_name) {
    tibble(
      model = model_name,
      waic = fit$waic$waic,
      eff_params = fit$waic$p.eff
    )
  }
) |>
  arrange(waic)

model_waic

bhf_best_model <- model_waic |>
  slice(1) |>
  pull(model)
message("Best fitted model spec: ", bhf_best_model)
bhf_selected_fit <- bhf_fits[[bhf_best_model]]
```
```{r}
#| label: bhf-fixed

bhf_fits[[bhf_best_model]]$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  transmute(
    term,
    mean,
    lower = `0.025quant`,
    upper = `0.975quant`
  )
```

Finally, the **spatial and temporal random-effect** hyperparameters `tau` and `phi` help show what the temporal and spatial terms are doing in the selected model. 

```{r}
#| label: bhf-hyper

bhf_fits[[bhf_best_model]]$summary.hyperpar |>
  as_tibble(rownames = "parameter") |>
  filter(str_detect(parameter, "time_inla_id|area_inla_id|Phi")) |>
  select(parameter, mean, sd, `0.025quant`, `0.5quant`, `0.975quant`) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

The random-effect table reports the posterior summaries for the temporal and spatial components in the selected BHF specification. Higher precision means less residual variation in that component, while the BYM2 `phi` indicates how much of the area effect is spatially structured.

## Estimation and aggregation with posterior draws

The aggregation step uses posterior draws from the selected `INLA` fit to produce BHF estimates for each MSOA-year cell. We first sample from the posterior and extract the fixed effects, spatial area effect, and temporal effect. We then rebuild the fitted model for every MSOA-year aggregation row and summarise the resulting draws.

:::{.callout-note}
Each posterior draw stores the fixed effects, spatial area effect, and temporal effect together in `INLA`'s latent vector. Fixed effects are labelled as `term:1`, while random effects are labelled as `effect:index`.
:::

```{r}
#| label: bhf-draws
#| code-fold: false

# Draw 300 posterior samples from the selected BHF fit.
set.seed(100)
posterior_draws <- inla.posterior.sample(
  n = 300,
  result = bhf_selected_fit
)

# Fixed-effect draw matrix
fixed_terms <- rownames(bhf_selected_fit$summary.fixed)
fixed_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0(fixed_terms, ":1"), 1]),
    fixed_terms
  )
})

# Area-effect draw matrix
area_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("area_inla_id:", area_ids), 1]),
    area_ids
  )
})

# Time-effect draw matrix
time_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("time_inla_id:", time_ids), 1]),
    time_ids
  )
})
```

For each posterior draw, we rebuild the fitted model for every MSOA-year row.

```{r}
#| label: bhf-predict
#| code-fold: false

prediction_draws <- sapply(seq_along(posterior_draws), \(draw_id) {
  log_odds <- fixed_draws["(Intercept)", draw_id] +
    aggregation_data$age * fixed_draws["age", draw_id] +
    aggregation_data$female * fixed_draws["female", draw_id] +
    aggregation_data$car_access * fixed_draws["car_access", draw_id] +
    area_draws[as.character(aggregation_data$area_inla_id), draw_id] +
    time_draws[as.character(aggregation_data$time_inla_id), draw_id]

  plogis(log_odds) # Convert log-odds to probability.
})
```

Finally, we summarise the draws into the final estimate, variance, and 95% credible interval.

```{r}
#| label: bhf-results
#| code-fold: false

final_bhf_results <- aggregation_data |>
  select(area_id, msoa_name, time_id) |>
  bind_cols(
    tibble(
      bhf_estimate = rowMeans(prediction_draws),
      bhf_variance = apply(prediction_draws, 1, var),
      bhf_lower = apply(prediction_draws, 1, quantile, probs = 0.025),
      bhf_upper = apply(prediction_draws, 1, quantile, probs = 0.975)
    )
  ) |>
  mutate(
    bhf_ci_width = bhf_upper - bhf_lower
  )

head(final_bhf_results)
```

```{r}
#| label: bhf-map
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

final_bhf_map_data <- msoa_boundaries |>
  left_join(final_bhf_results, by = "area_id")
bhf_estimate_limits <- range(final_bhf_results$bhf_estimate, na.rm = TRUE)
bhf_ci_width_limits <- c(0, max(final_bhf_results$bhf_ci_width, na.rm = TRUE))
bhf_wave_years <- sort(unique(final_bhf_results$time_id))

plot_bhf_estimate <- function(wave) {
  final_bhf_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = bhf_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = bhf_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_bhf_ci_width <- function(wave) {
  final_bhf_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = bhf_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = bhf_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(bhf_wave_years, plot_bhf_estimate)
uncertainty_maps <- lapply(bhf_wave_years, plot_bhf_ci_width)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(bhf_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = length(bhf_wave_years)) +
  plot_annotation(
    title = "Share of adults who cycle for transport frequently",
    subtitle = paste0("Selected BHF model (", bhf_best_model, ") and 95% CI width"),
    theme = map_annotation_theme
  )
```

Even with unit-level fitting to overcome weak direct estimation, the aggregation step using area summaries relies on strong assumptions about each area's population composition. Representing age, sex, and car access with area-level summaries, such as mean age, proportion female, and share of households with car access, can hide important compositional differences. Two areas may have similar summaries but very different age-sex-car structures, meaning the model could produce similar estimates even where cycling behaviour differs across specific combinations of these predictors.

This limitation can be addressed by using a poststratification table that contains population counts for the joint distribution of the relevant predictors. This is the logic of **Multilevel Regression with Poststratification (MRP)**: predict the outcome for each poststratification cell, then aggregate those predictions using the corresponding population counts so the final estimates better reflect the target population composition.

```{r}
#| label: bhf-export
#| code-fold: true

write_csv(final_bhf_results, file.path(base_path, "freqbike_estimates_bhf.csv"))
```

```{r}
#| label: bhf-area-effects-export
#| include: false
#| echo: false
#| message: false
#| warning: false

besag_model_data <- model_data |>
  mutate(
    area_struct_id = area_inla_id,
    area_unstruct_id = area_inla_id
  )

bhf_besag_formula <- indicator ~
  1 +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_struct_id,
    model = "besag",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  ) +
  f(area_unstruct_id, model = "iid")

if (bhf_best_model %in% c("M2 +Demographic", "M3 +CarAccess")) {
  bhf_besag_formula <- update(bhf_besag_formula, . ~ . +
    age +
    female
  )
}

if (bhf_best_model == "M3 +CarAccess") {
  bhf_besag_formula <- update(bhf_besag_formula, . ~ . + car_access)
}

bhf_besag_fit <- inla(
  formula = bhf_besag_formula,
  family = "binomial",
  Ntrials = 1,
  data = besag_model_data,
  control.compute = list(waic = TRUE)
)

area_effect_lookup <- inla_domain_index |>
  distinct(area_id, area_inla_id)

bhf_area_effects <- bind_rows(
  bhf_besag_fit$summary.random$area_struct_id |>
    as_tibble() |>
    transmute(
      area_inla_id = ID,
      component = "Structured",
      effect = mean,
      lower = `0.025quant`,
      upper = `0.975quant`
    ),
  bhf_besag_fit$summary.random$area_unstruct_id |>
    as_tibble() |>
    transmute(
      area_inla_id = ID,
      component = "Unstructured",
      effect = mean,
      lower = `0.025quant`,
      upper = `0.975quant`
    )
) |>
  left_join(area_effect_lookup, by = "area_inla_id") |>
  mutate(method = "BHF") |>
  select(method, component, area_id, area_inla_id, effect, lower, upper)

write_csv(bhf_area_effects, file.path(base_path, "freqbike_area_effects_bhf.csv"))
```
