---
title: "Spatio-Temporal Battese-Harter-Fuller Unit-Level Model"
subtitle: "Predict small area estimates from survey microdata using temporal and spatial smoothing"
---

```{r}
#| label: setup-bhf
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(spdep)
library(INLA)
library(tibble)
library(scales)

base_path <- here::here("data")
domain_frame <- read_csv(
  file.path(base_path, "clean", "msoa_year_domain_frame.csv"),
  show_col_types = FALSE
)

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(msoa_code = all_of("MSOA21CD")) |>
  filter(str_starts(.data$msoa_code, "W"))
```

```{r}
#| label: plotting-config
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")
variance_label <- label_number(accuracy = 0.0001)
missing_fill <- "grey85"
diverging_low <- "#2166ac"
diverging_high <- "#b2182b"

map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

BHF unit-level SAE starts from respondent records rather than area-level direct estimates. Here we use a Bayesian extension of the Battese--Harter--Fuller model: individual responses borrow strength from respondent-level predictors and area/time random effects, then the fitted relationship is aggregated to MSOA--year estimates using known area-level predictor means.

A reminder that our target estimand is:

> Share of Welsh adults who walk for transport daily by MSOA in 2021, 2023 and 2025

## Define Domain, Indices And Spatial Graph

Before preparing the model inputs, we create a consistent numeric index for each MSOA and survey year in the complete MSOA--year domain. These indices are used by `INLA` to attach random effects to the correct area and period.

```{r}
#| label: define-model-index
#| code-fold: false

area_levels <- sort(unique(domain_frame[["msoa_code"]]))
year_levels <- sort(unique(domain_frame[["wave_year"]]))

area_index <- tibble(
  msoa_code = area_levels,
  area_id = seq_along(area_levels)
)

year_index <- tibble(
  wave_year = year_levels,
  year_id = seq_along(year_levels)
)
```

To include a spatial random effect, we also need an adjacency graph `.adj` as input. We again use Queen contiguity.

```{r}
#| label: build-spatial-graph
#| code-fold: false

msoa_boundaries_sorted <- msoa_boundaries |>
  inner_join(area_index, by = "msoa_code") |>
  arrange(.data$area_id)

msoa_neighbours <- poly2nb(
  msoa_boundaries_sorted,
  queen = TRUE,
  row.names = msoa_boundaries_sorted$msoa_code
)

spatial_graph_file <- tempfile(fileext = ".adj")

nb2INLA(
  file = spatial_graph_file,
  nb = msoa_neighbours
)
```

## Prepare Model And Aggregation Data

BHF modelling has two separate data-preparation tasks:

- prepare the respondent-level rows used to fit the model;
- prepare the area-level rows used to aggregate the fitted relationship to MSOA--year estimates.

### Respondent Model Data

The respondent model data has one row per sampled person. For this constrained example we use:

- `indicator` as the binary response, our target;
- `no_car` as the unit-level auxiliary variable, our predictor;
- `wave_year` as the survey wave year identifier;
- `msoa_code` as the area identifier;

```{r}
#| label: prepare-unit-sample
#| code-fold: false

unit_predictors <- read_csv(
  file.path(base_path, "clean", "unit_predictors_daily_walking.csv"),
  show_col_types = FALSE
)

walking_indicator <- read_csv(
  file.path(base_path, "clean", "ind_weight_freq_walker.csv"),
  show_col_types = FALSE
) |>
  select(
    unit_id,
    wave_year,
    msoa_code,
    indicator
  )

unit_sample <- unit_predictors |>
  left_join(
    walking_indicator,
    by = "unit_id"
  ) |>
  filter(
    !is.na(.data$indicator),
    !is.na(.data$msoa_code),
    !is.na(.data$wave_year),
    !is.na(.data$no_car)
  )

head(unit_sample)
```

### Area Aggregation Data

The aggregation input contains the complete MSOA--year grid we want to estimate, and the MSOA-level mean of the unit-level predictor. Since `no_car` is binary, the area mean is simply the share of households with no car or van (a pre-processed area-level predictor).

```{r}
#| label: prepare-unit-aggregation-data
#| code-fold: false

area_covariates <- read_csv(
  file.path(base_path, "clean", "area_covariates_daily_walking.csv"),
  show_col_types = FALSE
)

prediction_frame <- domain_frame |>
  select(msoa_code, msoa_name, wave_year) |>
  left_join(
    area_covariates |>
      transmute(
        msoa_code,
        no_car = raw_share_no_car_households
      ),
    by = "msoa_code"
  )

head(prediction_frame)
```

::: {.callout-note}
The area mean for `no_car` comes from Census 2021 and is used as a time-stable MSOA predictor across the three survey waves. This is reasonable for auxiliary variables that are unlikely to change quickly.
:::

Finally, we prepare two datasets for `INLA`.

- `model_data` contains the sampled respondent records used to fit the model. `no_car` here represents individual responses.
- `aggregation_data` contains one row for each MSOA--year we want to estimate. `no_car` here represents area means.

```{r}
#| label: prepare-model-data
#| code-fold: false

model_data <- unit_sample |>
  left_join(area_index, by = "msoa_code") |>
  left_join(year_index, by = "wave_year") |>
  transmute(
    unit_id,
    msoa_code,
    wave_year,
    indicator,
    no_car,
    year_id,
    area_spatial_id = area_id
  ) |>
  arrange(.data$wave_year, .data$area_spatial_id)

aggregation_data <- prediction_frame |>
  left_join(area_index, by = "msoa_code") |>
  left_join(year_index, by = "wave_year") |>
  transmute(
    msoa_code,
    msoa_name,
    wave_year,
    no_car,
    year_id,
    area_spatial_id = area_id
  ) |>
  arrange(.data$wave_year, .data$area_spatial_id)

head(model_data)
head(aggregation_data)
```

## Fit And Evaluate Models

We fit linear mixed models to respondent-level records, then use them to predict MSOA--year means. This follows the BHF nested-error idea, extended here with temporal and spatial random effects.

::: {.callout-note}
The response is binary, so this is a linear probability version of the BHF-style model.
:::

The two models are:

| Model | Name | Specifications |
|---|---|---|
| M1 | Temporal + spatial | Respondent outcome with a temporal random walk and BYM2 spatial MSOA effect |
| M2 | +No-car | M1 plus respondent-level no-car access, aggregated using the MSOA Census share of households with no car or van |

```{r}
#| label: define-bhf-models
#| code-fold: false

m1_formula <- indicator ~
  1 +
  f(year_id, model = "rw1", constr = TRUE) +
  f(
    area_spatial_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m2_formula <- indicator ~
  no_car +
  f(year_id, model = "rw1", constr = TRUE) +
  f(
    area_spatial_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

model_formulas <- list(
  "M1 Temporal + Spatial" = m1_formula,
  "M2 +No-car" = m2_formula
)

```

`INLA` does not use a separate `predict(newdata = ...)` step; instead, it returns fitted values for every row in the input data. We therefore add aggregation MSOA--year rows with `indicator = NA` for `INLA` to return fitted values for them.

::: {.callout-note}
In the Fay--Herriot model, this feature allows for MSOA--years without survey data (unsampled) to also receive an estimate.
:::

```{r}
#| label: fit-bhf-models
#| code-fold: false

inla_input <- bind_rows(
  model_data |>
    transmute(
      row_type = "model",
      indicator,
      no_car,
      year_id,
      area_spatial_id
    ),
  aggregation_data |>
    transmute(
      row_type = "aggregation",
      indicator = NA_real_,
      no_car,
      year_id,
      area_spatial_id
    )
)

fit_bhf_model <- function(formula) {
  inla(
    formula = formula,
    family = "gaussian",
    data = inla_input,
    control.predictor = list(compute = TRUE),
    control.compute = list(waic = TRUE, dic = TRUE)
  )
}

bhf_fits <- lapply(model_formulas, fit_bhf_model)
```

WAIC compares the BHF fits to the sampled respondent outcomes. Here, the lower WAIC for M2 shows that adding the no-car predictor improves the temporal-spatial model enough to justify the extra term.

```{r}
#| label: bhf-waic-comparison

bhf_model_comparison <- bind_rows(lapply(names(bhf_fits), function(model_name) {
  tibble(
    model = model_name,
    waic = bhf_fits[[model_name]]$waic$waic,
    effective_parameters = bhf_fits[[model_name]]$waic$p.eff,
    dic = bhf_fits[[model_name]]$dic$dic
  )
})) |>
  arrange(.data$waic)

bhf_model_comparison
```

A look into the fitted coefficients for `no_car` sheds more light about its predictive power. The positive coefficient of `9.1pp` means respondents without access to a car or van are predicted to have a higher probability of walking for transport daily, after accounting for temporal and spatial structure.

```{r}
#| label: bhf-no-car-coefficient

bhf_fits[["M2 +No-car"]]$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  filter(.data$term == "no_car") |>
  transmute(
    term,
    mean = pp_label(.data$mean),
    lower = pp_label(.data[["0.025quant"]]),
    upper = pp_label(.data[["0.975quant"]])
  )
```

Finally, the **spatial and temporal random-effect** hyperparameters help show what the temporal and spatial terms are doing in the selected model. The temporal term has a precision parameter `tau`; the BYM2 spatial term has both `tau` and `phi`.

```{r}
#| label: bhf-random-effect-hyperparameters

bhf_fits[["M2 +No-car"]]$summary.hyperpar |>
  as_tibble(rownames = "parameter") |>
  filter(str_detect(.data$parameter, "year_id|area_spatial_id|Phi")) |>
  select(parameter, mean, sd, `0.025quant`, `0.5quant`, `0.975quant`) |>
  mutate(
    across(where(is.numeric), ~ round(.x, 3))
  )
```
The temporal `tau` is high, suggesting that the survey-wave adjustments are relatively small. By contrast, the spatial `tau` is lower, indicating more remaining variation between MSOAs than between survey waves.

Interestingly, the BYM2 `phi` is high and above `0.5`, suggesting that much of the remaining area-level variation in the BHF model is spatially structured. This may look different from the lower `phi` seen in the Fay--Herriot model, but the two values are not directly comparable. They come from different modelling units and specifications: the BHF model is fitted to individual respondent records, while the Fay--Herriot model is fitted to area-level direct estimates and their sampling variances. Therefore, the higher `phi` here should be read as evidence that, within this simpler unit-level specification, the remaining between-MSOA differences are spatially patterned.

With M2 being the most performant, we now collate the final fitted values for the aggregation rows into one MSOA--year results table. 

```{r}
#| label: collect-bhf-estimates

aggregation_row_ids <- which(inla_input$row_type == "aggregation")

bhf_estimates <- as_tibble(
  bhf_fits[["M2 +No-car"]]$summary.fitted.values[
    aggregation_row_ids,
    ,
    drop = FALSE
  ]
) |>
  transmute(
    bhf_estimate = pmin(pmax(.data$mean, 0), 1),
    bhf_variance = .data$sd^2,
    bhf_lower = pmin(pmax(.data[["0.025quant"]], 0), 1),
    bhf_upper = pmin(pmax(.data[["0.975quant"]], 0), 1)
  ) |>
  bind_cols(
    aggregation_data |>
      select(
        msoa_code,
        msoa_name,
        wave_year,
        no_car
      )
  ) |>
  relocate(msoa_code, msoa_name, wave_year, no_car)

```

```{r}
#| label: final-bhf-estimate-map
#| fig-width: 16
#| fig-height: 12
#| out-width: "100%"

final_bhf_results <- bhf_estimates
final_bhf_map_data <- msoa_boundaries |>
  left_join(final_bhf_results, by = "msoa_code")
bhf_estimate_limits <- range(final_bhf_results$bhf_estimate, na.rm = TRUE)
bhf_variance_limits <- c(0, max(final_bhf_results$bhf_variance, na.rm = TRUE))
bhf_wave_years <- sort(unique(final_bhf_results$wave_year))

plot_bhf_estimate <- function(wave) {
  final_bhf_map_data |>
    filter(.data$wave_year == wave) |>
    ggplot() +
    geom_sf(aes(fill = .data$bhf_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = bhf_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_bhf_variance <- function(wave) {
  final_bhf_map_data |>
    filter(.data$wave_year == wave) |>
    ggplot() +
    geom_sf(aes(fill = .data$bhf_variance), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = bhf_variance_limits,
      labels = variance_label,
      na.value = missing_fill,
      name = "Variance"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(bhf_wave_years, plot_bhf_estimate)
uncertainty_maps <- lapply(bhf_wave_years, plot_bhf_variance)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("Posterior variance")

for (i in seq_len(length(bhf_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = 3) +
  plot_annotation(
    title = "Final BHF Estimates",
    subtitle = "Share of adults who walk for transport daily; M2 with temporal, spatial, and no-car terms",
    theme = map_annotation_theme
  )
```

## Compare With Direct And Fay-Herriot Estimates

We finish by comparing the BHF estimates with the direct and Fay--Herriot estimates. This is less about choosing a winner and more about identifying what the simple BHF specification is missing.

```{r}
direct_estimates <- read_csv(
  file.path(base_path, "freq_walker_estimates_direct.csv"),
  show_col_types = FALSE
) |>
  select(msoa_code, wave_year, direct_estimate, standard_error, variance, n)

fh_final_estimates <- read_csv(
  file.path(base_path, "freq_walker_estimates_area-fh.csv"),
  show_col_types = FALSE
) |>
  transmute(
    msoa_code,
    wave_year,
    fh_estimate,
    fh_variance
  )

bhf_estimates <- bhf_estimates |>
  left_join(direct_estimates, by = c("msoa_code", "wave_year")) |>
  left_join(fh_final_estimates, by = c("msoa_code", "wave_year")) |>
  relocate(direct_estimate, standard_error, variance, n, .after = wave_year)

bhf_estimates |>
  select(
    msoa_code,
    wave_year,
    direct_estimate,
    fh_estimate,
    bhf_estimate
  ) |>
  head()
```
```{r}
#| label: bhf-comparison-scatter
#| fig-height: 6
#| fig-width: 12

bhf_fh_comparison <- bhf_estimates |>
  transmute(
    msoa_code,
    msoa_name,
    wave_year,
    direct_estimate,
    fh_estimate,
    fh_variance,
    bhf_estimate,
    n
  ) |>
  mutate(bhf_minus_fh = .data$bhf_estimate - .data$fh_estimate)

bhf_scatter_comparison <- bhf_fh_comparison |>
  select(
    msoa_code,
    wave_year,
    n,
    bhf_estimate,
    direct_estimate,
    fh_estimate
  ) |>
  pivot_longer(
    cols = c(direct_estimate, fh_estimate),
    names_to = "comparison",
    values_to = "comparison_estimate"
  ) |>
  mutate(
    comparison = recode(
      .data$comparison,
      direct_estimate = "Direct estimates",
      fh_estimate = "Fay-Herriot estimates"
    )
  ) |>
  filter(!is.na(.data$comparison_estimate))

bhf_scatter_limits <- range(
  c(
    bhf_scatter_comparison$comparison_estimate,
    bhf_scatter_comparison$bhf_estimate
  ),
  na.rm = TRUE
)

bhf_scatter_comparison |>
  ggplot(aes(x = .data$comparison_estimate, y = .data$bhf_estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(aes(size = .data$n), alpha = 0.55, colour = "#1f4e5f") +
  facet_wrap(vars(.data$comparison), nrow = 1) +
  scale_x_continuous(labels = share_label) +
  scale_y_continuous(labels = share_label) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  coord_equal(xlim = bhf_scatter_limits, ylim = bhf_scatter_limits) +
  labs(
    title = "BHF Comparison",
    subtitle = "Dashed line shows equality; points are MSOA-years",
    x = "Comparison estimate",
    y = "BHF estimate (M2)"
  ) +
  theme_minimal(base_size = 12) +
  chart_text_theme +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

The BHF estimates are smoother than the direct estimates because the model borrows strength across respondents, years, and neighbouring MSOAs. However, the comparison with Fay--Herriot shows a more systematic issue: the BHF estimates are higher than the FH estimates almost everywhere.

```{r}
#| label: bhf-vs-fh-map
#| fig-width: 16
#| fig-height: 5
#| out-width: "100%"

bhf_fh_map_data <- msoa_boundaries |>
  left_join(bhf_fh_comparison, by = "msoa_code")

bhf_fh_difference_limit <- quantile(
  abs(bhf_fh_comparison$bhf_minus_fh),
  0.95,
  na.rm = TRUE,
  names = FALSE
)
bhf_fh_wave_years <- sort(unique(bhf_fh_comparison$wave_year))

plot_bhf_fh_difference <- function(wave) {
  bhf_fh_map_data |>
    filter(.data$wave_year == wave) |>
    ggplot() +
    geom_sf(aes(fill = .data$bhf_minus_fh), colour = "transparent") +
    scale_fill_gradient2(
      labels = pp_label,
      low = diverging_low,
      mid = "white",
      high = diverging_high,
      midpoint = 0,
      limits = c(-bhf_fh_difference_limit, bhf_fh_difference_limit),
      oob = squish,
      na.value = missing_fill,
      name = "BHF - FH"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

bhf_fh_difference_maps <- lapply(bhf_fh_wave_years, plot_bhf_fh_difference)
bhf_fh_difference_maps[[1]] <- bhf_fh_difference_maps[[1]] +
  ylab("Difference")

for (i in seq_len(length(bhf_fh_wave_years) - 1)) {
  bhf_fh_difference_maps[[i]] <- bhf_fh_difference_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(bhf_fh_difference_maps, ncol = 3) +
  plot_annotation(
    title = "BHF Minus Fay-Herriot",
    subtitle = "Positive values mean the BHF estimate is higher",
    theme = map_annotation_theme
  )
```

The Fay--Herriot model starts from weighted direct estimates and their sampling variances, so it remains closely tied to the design-based survey estimator. The BHF model is fitted directly to respondent records and does not fully account for survey weights or the complex survey design. If some groups are over-sampled, under-sampled, or unevenly distributed across areas, they can pull the fitted unit-level relationship away from the target population.

The model is also under-specified. It only includes no-car access, time, and spatial structure. Adding age, sex, ethnicity, employment status, or other respondent-level predictors would help account for sample composition, but it also raises the data requirement. For aggregation, we would no longer need only the MSOA mean of `no_car`; we would need a population table giving the joint distribution of all predictors used in the model.

That is the bridge to **multilevel regression and poststratification**. MRP fits a richer unit-level model, then averages the predicted outcomes over a poststratification table representing the target population. The next workbook takes that step.

```{r}
#| label: save-bhf-outputs
#| echo: false

write_csv(
  final_bhf_results,
  file.path(base_path, "freq_walker_estimates_bhf.csv")
)

write_csv(
  bhf_model_comparison,
  file.path(base_path, "freq_walker_bhf_model_summary.csv")
)
```
