---
title: "Spatio-Temporal Battese-Harter-Fuller Unit-Level Model"
subtitle: "Predict small area estimates from survey microdata using temporal and spatial smoothing"
---

```{r}
#| label: setup-bhf
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(spdep)
library(INLA)
library(tibble)
library(scales)

base_path <- here::here("data")
domain_frame <- read_csv(
  file.path(base_path, "clean", "msoa_year_domain_frame.csv"),
  show_col_types = FALSE
)

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = all_of("MSOA21CD")) |>
  filter(str_starts(.data$area_id, "W"))
```

```{r}
#| label: plotting-config
#| echo: false
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")
variance_label <- label_number(accuracy = 0.0001)
missing_fill <- "grey85"

map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

Unit-level SAE models respondent records directly. The Battese--Harter--Fuller model first estimates the individual outcome with a nested error regression using unit-level predictors form the survey together with spatial and temporal effects, and then aggregates the predictions for each *MSOA-year cell*.

::: {.callout-note title="Theory in brief"}
The unit-level route models respondent outcomes directly. For the area-time target $\mu_{jt}$, a simple nested-error form is:

$$
y_{ijt} = \alpha + \mathbf{x}_{ijt}^{T}\boldsymbol{\beta} + u_j + v_t + \epsilon_{ijt},
$$

where $\mathbf{x}_{ijt}$ are unit-level predictors, $u_j$ is an area effect, and $v_t$ is a time effect. The classical Battese--Harter--Fuller estimator can be applied to each area-time target:

$$
\hat{\mu}_{jt}^{\mathrm{EBLUP}} =
f_{jt} \bar{y}_{jtS}
+ \left(\bar{\mathbf{X}}_{jt} - f_{jt} \bar{\mathbf{x}}_{jtS}\right)^T
\hat{\boldsymbol{\beta}}
+ (1 - f_{jt})(\hat{u}_j + \hat{v}_t),
$$

where $f_{jt}=n_{jt}/N_{jt}$, $\bar{y}_{jtS}$ and $\bar{\mathbf{x}}_{jtS}$ are sample means in area $j$ and time period $t$, $\bar{\mathbf{X}}_{jt}$ is the corresponding population mean of the predictors, and $\hat{u}_j$ and $\hat{v}_t$ are fitted area and time effects. The area effect can also be replaced by a structured spatial effect.
:::

<div class="method-divider" aria-hidden="true"></div>

{{< include _includes/target-estimand.qmd >}}

## Prepare model and aggregation data

Before preparing the model inputs, we create a consistent numeric index for each area and survey year in the complete area-time domain. These indices are used by `INLA` to attach random effects to the correct area and period.

```{r}
#| label: define-model-index

area_levels <- sort(unique(domain_frame[["area_id"]]))
time_levels <- sort(unique(domain_frame[["time_id"]]))

area_index <- tibble(
  area_id = area_levels,
  area_inla_id = seq_along(area_levels)
)

time_index <- tibble(
  time_id = time_levels,
  time_inla_id = seq_along(time_levels)
)
```

To include a spatial random effect, we also need an adjacency graph `.adj` as input. We again use Queen contiguity.

```{r}
#| label: build-spatial-graph

msoa_boundaries_sorted <- msoa_boundaries |>
  inner_join(area_index, by = "area_id") |>
  arrange(.data$area_inla_id)

msoa_neighbours <- poly2nb(
  msoa_boundaries_sorted,
  queen = TRUE,
  row.names = msoa_boundaries_sorted$area_id
)

spatial_graph_file <- tempfile(fileext = ".adj")

nb2INLA(
  file = spatial_graph_file,
  nb = msoa_neighbours
)
```

Now, we assemble the data needed to fit the unit-level model. Unit-level SAE modelling has two separate data-preparation steps:

- prepare the respondent-level rows used to fit the model;
- prepare the area-level rows used to aggregate the fitted relationship to area-time estimates.

### Respondent model data

The respondent model data has one row per sampled person. For this constrained example we use:

- `indicator` as the binary response, our target;
- `no_car` as the unit-level auxiliary variable, our predictor;
- `time_id` as the survey wave year identifier;
- `area_id` as the area identifier;

```{r}
#| label: prepare-unit-sample

unit_predictors <- read_csv(
  file.path(base_path, "clean", "unit_predictors_daily_walking.csv"),
  show_col_types = FALSE
)

walking_indicator <- read_csv(
  file.path(base_path, "clean", "ind_weight_freq_walker.csv"),
  show_col_types = FALSE
) |>
  select(
    unit_id,
    time_id,
    area_id,
    indicator
  )

unit_sample <- unit_predictors |>
  left_join(
    walking_indicator,
    by = "unit_id"
  ) |>
  filter(
    !is.na(.data$indicator),
    !is.na(.data$area_id),
    !is.na(.data$time_id),
    !is.na(.data$no_car)
  )

head(unit_sample)
```

### Area aggregation data

The aggregation input contains the complete area-time grid we want to estimate, and the area-level mean of the unit-level predictor. Since `no_car` is binary, the area mean is simply the share of households with no car or van (a pre-processed area-level predictor).

```{r}
#| label: prepare-unit-aggregation-data

area_covariates <- read_csv(
  file.path(base_path, "clean", "area_covariates_daily_walking.csv"),
  show_col_types = FALSE
)

prediction_frame <- domain_frame |>
  select(area_id, msoa_name, time_id) |>
  left_join(
    area_covariates |>
      transmute(
        area_id,
        no_car = raw_share_no_car_households
      ),
    by = "area_id"
  )

head(prediction_frame)
```

::: {.callout-note}
The area mean for `no_car` comes from Census 2021 and is used as a time-stable MSOA predictor across the three survey waves. This is reasonable for auxiliary variables that are unlikely to change quickly.
:::

Finally, we prepare two datasets for `INLA`.

- `model_data` contains the sampled respondent records used to fit the model. `no_car` here represents individual responses.
- `aggregation_data` contains one row for each area-time cell we want to estimate. `no_car` here represents area means.

```{r}
#| label: prepare-model-data

model_data <- unit_sample |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  transmute(
    unit_id,
    area_id,
    time_id,
    indicator,
    no_car,
    time_inla_id,
    area_spatial_id = area_inla_id
  ) |>
  arrange(.data$time_id, .data$area_spatial_id)

aggregation_data <- prediction_frame |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  transmute(
    area_id,
    msoa_name,
    time_id,
    no_car,
    time_inla_id,
    area_spatial_id = area_inla_id
  ) |>
  arrange(.data$time_id, .data$area_spatial_id)

head(model_data)
head(aggregation_data)
```

## Fit and evaluate models

We fit linear mixed models to respondent-level records, then use them to predict area-time means. This follows the BHF nested-error idea, extended here with temporal and spatial random effects.

::: {.callout-note}
The response is binary, so this is a linear probability version of the BHF-style model.
:::

The two models are:

| Model | Name | Specifications |
|---|---|---|
| M1 | Temporal + spatial | Respondent outcome with a temporal random walk and BYM2 spatial MSOA effect |
| M2 | +No-car | M1 plus respondent-level no-car access, aggregated using the MSOA Census share of households with no car or van |

```{r}
#| label: define-bhf-models
#| code-fold: false

m1_formula <- indicator ~
  1 +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_spatial_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m2_formula <- indicator ~
  no_car +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_spatial_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

model_formulas <- list(
  "M1 Temporal + Spatial" = m1_formula,
  "M2 +No-car" = m2_formula
)

```

`INLA` does not use a separate `predict(newdata = ...)` step; instead, it returns fitted values for every row in the input data. We therefore add aggregation rows for the area-time cells with `indicator = NA` for `INLA` to return fitted values for them.

::: {.callout-note}
In this workflow, this feature allows area-time cells without survey data to also receive an estimate.
:::

```{r}
#| label: fit-bhf-models
#| code-fold: false

inla_input <- bind_rows(
  model_data |>
    transmute(
      row_type = "model",
      indicator,
      no_car,
      time_inla_id,
      area_spatial_id
    ),
  aggregation_data |>
    transmute(
      row_type = "aggregation",
      indicator = NA_real_,
      no_car,
      time_inla_id,
      area_spatial_id
    )
)

fit_bhf_model <- function(formula) {
  inla(
    formula = formula,
    family = "gaussian",
    data = inla_input,
    control.predictor = list(compute = TRUE),
    control.compute = list(waic = TRUE, dic = TRUE)
  )
}

bhf_fits <- lapply(model_formulas, fit_bhf_model)
```

WAIC compares the BHF fits to the sampled respondent outcomes. Here, the lower WAIC for M2 shows that adding the no-car predictor improves the temporal-spatial model enough to justify the extra term.

```{r}
#| label: bhf-waic-comparison

bhf_model_comparison <- bind_rows(lapply(names(bhf_fits), function(model_name) {
  tibble(
    model = model_name,
    waic = bhf_fits[[model_name]]$waic$waic,
    effective_parameters = bhf_fits[[model_name]]$waic$p.eff,
    dic = bhf_fits[[model_name]]$dic$dic
  )
})) |>
  arrange(.data$waic)

bhf_model_comparison
```

A look into the fitted coefficients for `no_car` sheds more light about its predictive power. The positive coefficient of `9.1pp` means respondents without access to a car or van are predicted to have a higher probability of walking for transport daily, after accounting for temporal and spatial structure.

```{r}
#| label: bhf-no-car-coefficient

bhf_fits[["M2 +No-car"]]$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  filter(.data$term == "no_car") |>
  transmute(
    term,
    mean = pp_label(.data$mean),
    lower = pp_label(.data[["0.025quant"]]),
    upper = pp_label(.data[["0.975quant"]])
  )
```

Finally, the **spatial and temporal random-effect** hyperparameters help show what the temporal and spatial terms are doing in the selected model. The temporal term has a precision parameter `tau`; the BYM2 spatial term has both `tau` and `phi`.

```{r}
#| label: bhf-random-effect-hyperparameters

bhf_fits[["M2 +No-car"]]$summary.hyperpar |>
  as_tibble(rownames = "parameter") |>
  filter(str_detect(.data$parameter, "time_inla_id|area_spatial_id|Phi")) |>
  select(parameter, mean, sd, `0.025quant`, `0.5quant`, `0.975quant`) |>
  mutate(
    across(where(is.numeric), ~ round(.x, 3))
  )
```
The temporal `tau` is high, suggesting that the survey-wave adjustments are relatively small. By contrast, the spatial `tau` is lower, indicating more remaining variation between MSOAs than between survey waves.

The BYM2 `phi` is above `0.5`, so in this simple BHF model specification much of the remaining between-MSOA variation is spatially structured. By contrast, the lower `phi` in the Fay--Herriot model may reflect its richer specification: with more area-level predictors included, less residual between-MSOA variation is left for the spatial random effect to explain.

## Report estimates and uncertainty

Using M2 as the selected model, we extract fitted values for all MSOA-year rows. The posterior mean is the final BHF estimate, and the posterior 95% interval represents its uncertainty.

```{r}
#| label: collect-bhf-estimates

aggregation_row_ids <- which(inla_input$row_type == "aggregation")

bhf_estimates <- as_tibble(
  bhf_fits[["M2 +No-car"]]$summary.fitted.values[
    aggregation_row_ids,
    ,
    drop = FALSE
  ]
) |>
  transmute(
    bhf_estimate = pmin(pmax(.data$mean, 0), 1),
    bhf_variance = .data$sd^2,
    bhf_lower = pmin(pmax(.data[["0.025quant"]], 0), 1),
    bhf_upper = pmin(pmax(.data[["0.975quant"]], 0), 1)
  ) |>
  mutate(
    bhf_ci_width = .data$bhf_upper - .data$bhf_lower
  ) |>
  bind_cols(
    aggregation_data |>
      select(
        area_id,
        msoa_name,
        time_id,
        no_car
      )
  ) |>
  relocate(area_id, msoa_name, time_id, no_car)

```
```{r}
#| label: final-bhf-estimate-map
#| echo: false
#| fig-width: 16
#| fig-height: 12
#| out-width: "100%"

final_bhf_results <- bhf_estimates
final_bhf_map_data <- msoa_boundaries |>
  left_join(final_bhf_results, by = "area_id")
bhf_estimate_limits <- range(final_bhf_results$bhf_estimate, na.rm = TRUE)
bhf_ci_width_limits <- c(0, max(final_bhf_results$bhf_ci_width, na.rm = TRUE))
bhf_wave_years <- sort(unique(final_bhf_results$time_id))

plot_bhf_estimate <- function(wave) {
  final_bhf_map_data |>
    filter(.data$time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = .data$bhf_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = bhf_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_bhf_ci_width <- function(wave) {
  final_bhf_map_data |>
    filter(.data$time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = .data$bhf_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = bhf_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(bhf_wave_years, plot_bhf_estimate)
uncertainty_maps <- lapply(bhf_wave_years, plot_bhf_ci_width)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(bhf_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = 3) +
  plot_annotation(
    title = "Final Estimates (BHF)",
    subtitle = "Posterior mean and 95% interval width; with temporal, spatial effects",
    theme = map_annotation_theme
  )
```
The scatterplot shows that BHF estimates are much smoother than the direct estimates and closer to the Fay--Herriot estimates, as expected for a model-based approach that borrows strength across area-time cells.
```{r}
#| label: load-comparison-estimates
direct_estimates <- read_csv(
  file.path(base_path, "freq_walker_estimates_direct.csv"),
  show_col_types = FALSE
) |>
  select(area_id, time_id, direct_estimate, n)

fh_final_estimates <- read_csv(
  file.path(base_path, "freq_walker_estimates_area-fh.csv"),
  show_col_types = FALSE
) |>
  transmute(
    area_id,
    time_id,
    fh_estimate
  )

bhf_estimates <- bhf_estimates |>
  left_join(direct_estimates, by = c("area_id", "time_id")) |>
  left_join(fh_final_estimates, by = c("area_id", "time_id")) |>
  relocate(direct_estimate, n, .after = time_id)
```


```{r}
#| label: bhf-comparison-scatter
#| echo: false
#| fig-height: 6
#| fig-width: 12

bhf_scatter_comparison <- bhf_estimates |>
  select(
    area_id,
    time_id,
    n,
    bhf_estimate,
    direct_estimate,
    fh_estimate
  ) |>
  pivot_longer(
    cols = c(direct_estimate, fh_estimate),
    names_to = "comparison",
    values_to = "comparison_estimate"
  ) |>
  mutate(
    comparison = recode(
      .data$comparison,
      direct_estimate = "Direct estimates",
      fh_estimate = "Fay-Herriot estimates"
    )
  ) |>
  filter(!is.na(.data$comparison_estimate))

bhf_scatter_x_limits <- c(
  0,
  max(bhf_scatter_comparison$comparison_estimate, na.rm = TRUE)
)
bhf_scatter_y_limits <- c(
  0,
  max(bhf_scatter_comparison$bhf_estimate, na.rm = TRUE)
)

bhf_scatter_comparison |>
  ggplot(aes(x = .data$comparison_estimate, y = .data$bhf_estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(aes(size = .data$n), alpha = 0.55, colour = "#1f4e5f") +
  facet_wrap(vars(.data$comparison), nrow = 1) +
  scale_x_continuous(labels = share_label) +
  scale_y_continuous(labels = share_label) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  coord_cartesian(
    xlim = bhf_scatter_x_limits,
    ylim = bhf_scatter_y_limits
  ) +
  labs(
    title = "BHF comparison",
    subtitle = "Dashed line shows equality; points are area-time cells",
    x = "Comparison estimate",
    y = "BHF estimate (M2)"
  ) +
  theme_minimal(base_size = 12) +
  chart_text_theme +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

The BHF and Fay--Herriot estimates differ because the two models use the survey information differently. This simple BHF example is fitted to unweighted respondent records and uses only one respondent-level predictor. The final Fay--Herriot model starts from weighted direct estimates, uses their sampling variances, and includes a richer set of area-level predictors and spatio-temporal effects.

Adding more unit-level predictors such as age, sex, ethnicity, or employment status can help; aggregation then requires a population table with the joint distribution of those predictors for every area-time cell. This requirement leads to **multilevel regression with poststratification**: MRP fits a richer unit-level model and averages predicted outcomes over a poststratification table representing the target population.

```{r}
#| label: save-bhf-outputs
#| echo: false

write_csv(
  final_bhf_results,
  file.path(base_path, "freq_walker_estimates_bhf.csv")
)
```
