---
title: "Comparative diagnostics"
subtitle: "Benefits and trade-offs"
---


```{r}
#| label: val-setup
#| include: false

library(dplyr)
library(readr)
library(stringr)
library(sf)
library(ggplot2)
library(patchwork)
library(scales)

base_path <- if (dir.exists("data")) {
  normalizePath("data")
} else {
  normalizePath(file.path("..", "data"))
}

method_levels <- c("Direct", "FH", "BHF", "MRP Shallow", "MRP Rich")
missing_fill <- "grey85"
underestimate_colour <- "#2166ac"
overestimate_colour <- "#b2182b"
year_colours <- c("#2166ac", "#b2182b", "#4d9221")

residual_label <- label_number(accuracy = 0.01)
ci_width_label <- label_number(accuracy = 1, scale = 100, suffix = " pp")
share_label <- label_percent(accuracy = 1)

safe_limit <- function(x, fallback = 0.01, symmetric = FALSE, p = NULL) {
  x <- x[is.finite(x)]
  if (length(x) == 0) {
    return(if (symmetric) c(-fallback, fallback) else c(0, fallback))
  }

  limit <- if (is.null(p)) {
    max(abs(x), na.rm = TRUE)
  } else if (symmetric) {
    quantile(abs(x), p, na.rm = TRUE, names = FALSE)
  } else {
    quantile(x, p, na.rm = TRUE, names = FALSE)
  }

  if (!is.finite(limit) || limit == 0) {
    limit <- fallback
  }

  if (symmetric) c(-limit, limit) else c(0, limit)
}

map_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey30"),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

annotation_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  plot.caption = element_text(size = 10, hjust = 0.5, colour = "grey30")
)

chart_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
    legend.position = "bottom"
  )

colourbar_guide <- guide_colourbar(
  title.position = "top",
  barwidth = grid::unit(5, "cm"),
  barheight = grid::unit(0.35, "cm")
)

method_map <- function(data, fill_col) {
  data |>
    ggplot() +
    geom_sf(aes(fill = .data[[fill_col]]), colour = "transparent") +
    facet_wrap(vars(method), nrow = 1) +
    theme_void() +
    map_text_theme +
    theme(
      strip.text = element_text(size = 12, face = "bold"),
      legend.position = "bottom"
    )
}

read_estimate_set <- function(path, method_name, estimate_col, ci_width_col = NULL, variance_col = NULL) {
  estimate_data <- read_csv(path, show_col_types = FALSE)

  ci_width <- if (!is.null(ci_width_col)) {
    estimate_data[[ci_width_col]]
  } else if (!is.null(variance_col)) {
    2 * qnorm(0.975) * sqrt(estimate_data[[variance_col]])
  } else {
    NA_real_
  }

  estimate_data |>
    transmute(
      method = method_name,
      area_id,
      time_id,
      estimate = .data[[estimate_col]],
      ci_width = ci_width
    )
}

obs <- read_csv(
  file.path(base_path, "clean", "validation_bicycle_to_work_share.csv"),
  show_col_types = FALSE
) |>
  transmute(
    area_id,
    time_id,
    msoa_name,
    obs = validation_bicycle_to_work_share
  )

estimate_sets <- bind_rows(
  read_estimate_set(
    file.path(base_path, "freqbike_estimates_direct.csv"),
    "Direct",
    "direct_estimate",
    variance_col = "variance"
  ),
  read_estimate_set(
    file.path(base_path, "freqbike_estimates_fh.csv"),
    "FH",
    "fh_estimate",
    ci_width_col = "fh_ci_width"
  ),
  read_estimate_set(
    file.path(base_path, "freqbike_estimates_bhf.csv"),
    "BHF",
    "bhf_estimate",
    ci_width_col = "bhf_ci_width"
  ),
  read_estimate_set(
    file.path(base_path, "freqbike_estimates_mrp_shallow.csv"),
    "MRP Shallow",
    "shallow_mrp_estimate",
    ci_width_col = "shallow_mrp_ci_width"
  ),
  read_estimate_set(
    file.path(base_path, "freqbike_estimates_mrp_rich.csv"),
    "MRP Rich",
    "mrp_estimate",
    ci_width_col = "mrp_ci_width"
  )
)

validation_residuals <- estimate_sets |>
  inner_join(obs, by = c("area_id", "time_id")) |>
  mutate(
    method = factor(method, levels = method_levels),
    residual = if_else(estimate > 0, (obs - estimate) / sqrt(estimate), NA_real_),
    abs_residual = abs(residual),
    residual_direction = case_when(
      residual > 0 ~ "Underestimate",
      residual < 0 ~ "Overestimate",
      TRUE ~ "Zero"
    ),
    residual_direction = factor(
      residual_direction,
      levels = c("Overestimate", "Zero", "Underestimate")
    )
  )

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = all_of("MSOA21CD")) |>
  filter(str_starts(area_id, "W"))

map_data <- msoa_boundaries |>
  left_join(validation_residuals, by = "area_id")

residual_colour_limits <- safe_limit(validation_residuals$residual, symmetric = TRUE, p = 0.95)
residual_axis_limits <- safe_limit(validation_residuals$residual, symmetric = TRUE)
ci_width_colour_limits <- safe_limit(validation_residuals$ci_width, p = 0.95)
ci_width_axis_limits <- safe_limit(validation_residuals$ci_width)

method_summary <- validation_residuals |>
  group_by(method) |>
  summarise(
    area_time_cells = n(),
    median_estimate_pp = median(estimate, na.rm = TRUE) * 100,
    estimate_iqr_pp = IQR(estimate, na.rm = TRUE) * 100,
    mean_abs_residual = mean(abs_residual, na.rm = TRUE),
    median_abs_residual = median(abs_residual, na.rm = TRUE),
    p90_abs_residual = quantile(abs_residual, 0.9, na.rm = TRUE, names = FALSE),
    mean_ci_width_pp = mean(ci_width, na.rm = TRUE) * 100,
    median_ci_width_pp = median(ci_width, na.rm = TRUE) * 100,
    p90_ci_width_pp = quantile(ci_width, 0.9, na.rm = TRUE, names = FALSE) * 100,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

This comparison focuses on the trade-offs between direct estimation, area-level SAE, unit-level SAE, and poststratified unit-level SAE. The diagnostics compare the scale and smoothness of the estimates, agreement between methods, external validation residuals, uncertainty, and local residual structure used by the models.

## Method Trade-Offs

| Method | Main input | Main trade-off |
|---|---|---|
| Direct | Survey-weighted area-time estimates | Preserves the survey design most directly, but sparse MSOA-year cells can be unstable or missing. |
| FH | Weighted direct estimates and sampling variances | Uses the survey design through the direct estimates, but inherits their sparsity and degenerate variance issues. |
| BHF | Respondent records with unit-level predictors | Uses individual outcomes directly, but the fitted likelihood does not use survey weights. |
| MRP Shallow | Respondent records plus age-sex population cells and contextual predictors | Adds population reweighting, but only over the available poststratification cells. |
| MRP Rich | Respondent records plus age-sex-car population cells | Gives the closest alignment between model predictors and population composition, but holds the 2021 poststratification structure across later waves. |

```{r}
#| label: val-method-summary

method_summary |>
  select(
    method,
    area_time_cells,
    median_estimate_pp,
    estimate_iqr_pp,
    median_abs_residual,
    median_interval_width_pp = median_ci_width_pp
  )
```

## Estimate Scale and Smoothing

The distribution and rank plots show how much each method changes the spread of MSOA-year estimates. Direct estimates retain the most cell-level jaggedness, while the model-based approaches borrow strength across areas, years, predictors, or poststratification cells.

```{r}
#| label: val-estimate-dist-rank
#| echo: false
#| fig-width: 14
#| fig-height: 20
#| out-width: "100%"

estimate_plot_data <- estimate_sets |>
  filter(!is.na(estimate)) |>
  mutate(
    method = factor(method, levels = method_levels),
    time_id = factor(time_id, levels = sort(unique(time_id)))
  )

rank_reference_year <- levels(estimate_plot_data$time_id)[1]

estimate_rank <- estimate_plot_data |>
  group_by(method) |>
  filter(time_id == rank_reference_year) |>
  arrange(estimate, .by_group = TRUE) |>
  mutate(msoa_rank = row_number()) |>
  ungroup() |>
  select(method, area_id, msoa_rank)

ranked_estimate_data <- estimate_plot_data |>
  inner_join(estimate_rank, by = c("method", "area_id")) |>
  mutate(method = factor(method, levels = method_levels))

method_plots <- unlist(lapply(method_levels, function(method_name) {
  show_year_legend <- method_name == method_levels[1]

  distribution_plot <- estimate_plot_data |>
    filter(method == method_name) |>
    ggplot(aes(x = estimate, colour = time_id, fill = time_id)) +
    geom_density(linewidth = 0.8, alpha = 0.2, show.legend = show_year_legend) +
    scale_x_continuous(labels = share_label, limits = c(0, 0.1)) +
    scale_colour_manual(values = year_colours) +
    scale_fill_manual(values = year_colours) +
    guides(fill = "none") +
    labs(
      title = method_name,
      x = "Estimate",
      y = "Density",
      colour = "Survey year"
    ) +
    chart_theme +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(size = 12, face = "bold", hjust = 0),
      legend.position = "bottom"
    )

  rank_plot <- ranked_estimate_data |>
    filter(method == method_name) |>
    ggplot(aes(x = msoa_rank, y = estimate, colour = time_id, group = time_id)) +
    geom_line(linewidth = 0.5, alpha = 0.9, show.legend = FALSE) +
    scale_y_continuous(labels = share_label, limits = c(0, 0.1)) +
    scale_colour_manual(values = year_colours) +
    labs(
      x = "Ranked MSOA",
      y = "Estimate",
      colour = "Survey year"
    ) +
    chart_theme +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_blank(),
      legend.position = "bottom"
    )

  list(distribution_plot, rank_plot)
}), recursive = FALSE)

wrap_plots(method_plots, ncol = 2, widths = c(1, 1.35), guides = "collect") +
  plot_annotation(
    title = "Estimate distributions and ranked MSOAs",
    subtitle = "Each row shows one method; MSOA ranks are ordered by the 2021 estimate within that method",
    theme = annotation_theme
  ) &
  theme(legend.position = "bottom")
```

## Method Agreement

MRP Rich is used here as a modelling reference because it uses the fullest poststratification structure available in the workflow. It is not an external benchmark; disagreement with it indicates how much the other methods differ from the richest unit-level specification.

```{r}
#| label: val-method-agreement
#| echo: false
#| fig-width: 18
#| fig-height: 5
#| out-width: "100%"

reference_method <- "MRP Rich"

method_agreement <- estimate_sets |>
  filter(method != reference_method) |>
  inner_join(
    estimate_sets |>
      filter(method == reference_method) |>
      select(area_id, time_id, reference_estimate = estimate),
    by = c("area_id", "time_id")
  ) |>
  mutate(
    method = factor(method, levels = setdiff(method_levels, reference_method)),
    time_id = factor(time_id, levels = sort(unique(time_id)))
  ) |>
  filter(!is.na(estimate), !is.na(reference_estimate))

agreement_limits <- safe_limit(
  c(method_agreement$estimate, method_agreement$reference_estimate),
  fallback = 0.1,
  p = 0.99
)

method_agreement |>
  ggplot(aes(x = estimate, y = reference_estimate, colour = time_id)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(alpha = 0.55, size = 1.2) +
  facet_wrap(vars(method), nrow = 1) +
  coord_equal(xlim = agreement_limits, ylim = agreement_limits) +
  scale_x_continuous(labels = share_label) +
  scale_y_continuous(labels = share_label) +
  scale_colour_manual(values = year_colours) +
  labs(
    title = "Agreement with MRP Rich",
    subtitle = "Dashed line shows equality; points are MSOA-year cells",
    x = "Comparison estimate",
    y = "MRP Rich estimate",
    colour = "Survey year"
  ) +
  chart_theme +
  theme(panel.grid.minor = element_blank())
```

## External Validation Residuals

The validation benchmark is the Census bicycle-to-work share. It is related to frequent cycling for transport but does not measure the same target outcome, so the residuals should be read as a directional comparison rather than a direct accuracy test.

$$
\text{obs} = \frac{\text{bicycle-to-work count}}{\text{total commuters}}
$$
$$
\text{residual} = \frac{\text{obs} - \text{estimate}}{\sqrt{estimate}}
$$
```{r}
#| label: val-resid-summary

method_summary |>
  select(
    method,
    area_time_cells,
    mean_abs_residual,
    median_abs_residual,
    p90_abs_residual
  )
```

```{r}
#| label: val-resid-map
#| echo: false
#| fig-width: 18
#| fig-height: 8
#| out-width: "100%"

method_map(map_data, "residual") +
  scale_fill_gradient2(
    low = overestimate_colour,
    mid = "white",
    high = underestimate_colour,
    midpoint = 0,
    limits = residual_colour_limits,
    breaks = pretty_breaks(n = 3),
    oob = squish,
    labels = residual_label,
    na.value = missing_fill,
    name = "Residual",
    guide = colourbar_guide
  ) +
  plot_annotation(
    title = "Validation residuals",
    theme = annotation_theme
  )
```

```{r}
#| label: val-resid-density
#| echo: false
#| fig-width: 18
#| fig-height: 4
#| out-width: "100%"

validation_residuals |>
  filter(!is.na(residual)) |>
  ggplot(aes(x = residual, fill = residual_direction)) +
  geom_histogram(
    bins = 32,
    colour = "white",
    linewidth = 0.15,
    boundary = 0
  ) +
  geom_vline(xintercept = 0, colour = "grey25", linewidth = 0.35) +
  facet_wrap(vars(method), nrow = 1) +
  coord_cartesian(xlim = residual_axis_limits) +
  scale_fill_manual(
    values = c(
      "Overestimate" = overestimate_colour,
      "Zero" = "grey70",
      "Underestimate" = underestimate_colour
    ),
    name = NULL
  ) +
  scale_x_continuous(labels = residual_label) +
  labs(
    x = "Residual",
    y = "MSOA-year cells"
  ) +
  chart_theme +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank()
  )
```
## Uncertainty Width

Interval width shows the precision gained or lost by each method. Direct-estimate widths reflect design-based sampling variance, while model-based widths reflect posterior uncertainty after borrowing strength across areas, years, predictors, or population cells.

```{r}
#| label: val-ci-summary
method_summary |>
  select(
    method,
    area_time_cells,
    mean_ci_width_pp,
    median_ci_width_pp,
    p90_ci_width_pp
  )
```

```{r}
#| label: val-ci-map
#| echo: false
#| fig-width: 18
#| fig-height: 8
#| out-width: "100%"

method_map(map_data, "ci_width") +
  scale_fill_distiller(
    palette = "YlGnBu",
    direction = 1,
    limits = ci_width_colour_limits,
    breaks = pretty_breaks(n = 3),
    oob = squish,
    labels = ci_width_label,
    na.value = missing_fill,
    name = "CI width",
    guide = colourbar_guide
  ) +
  plot_annotation(
    title = "CI widths",
    theme = annotation_theme
  )
```

```{r}
#| label: val-ci-density
#| echo: false
#| fig-width: 18
#| fig-height: 4
#| out-width: "100%"

validation_residuals |>
  filter(!is.na(ci_width)) |>
  ggplot(aes(x = ci_width)) +
  geom_histogram(
    bins = 32,
    fill = "#3b7ea1",
    colour = "white",
    linewidth = 0.15,
    boundary = 0
  ) +
  facet_wrap(vars(method), nrow = 1) +
  coord_cartesian(xlim = ci_width_axis_limits) +
  scale_x_continuous(labels = ci_width_label) +
  labs(
    x = "CI width",
    y = "MSOA-year cells"
  ) +
  chart_theme +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank()
  ) +
  theme(legend.position = "none")
```

## Local Residual Effects

The unstructured area-effect maps show where the fitted models allocate remaining area-level variation after the fixed effects and temporal structure.

```{r}
#| label: val-area-effect-map
#| echo: false
#| fig-width: 18
#| fig-height: 7
#| out-width: "100%"

area_effect_method_levels <- c("FH", "BHF", "MRP Shallow", "MRP Rich")
area_effect_files <- c(
  "freqbike_area_effects_fh.csv",
  "freqbike_area_effects_bhf.csv",
  "freqbike_area_effects_mrp_shallow.csv",
  "freqbike_area_effects_mrp_rich.csv"
)

area_effects <- area_effect_files |>
  lapply(\(file_name) read_csv(file.path(base_path, file_name), show_col_types = FALSE)) |>
  bind_rows() |>
  filter(component == "Unstructured") |>
  mutate(
    method = factor(method, levels = area_effect_method_levels)
  )

area_effect_limits <- safe_limit(
  area_effects$effect,
  fallback = 0.01,
  symmetric = TRUE,
  p = 0.98
)

area_effect_label_accuracy <- if (max(abs(area_effect_limits)) < 0.01) 0.0001 else 0.01

msoa_boundaries |>
  select(area_id) |>
  left_join(area_effects, by = "area_id") |>
  method_map("effect") +
  scale_fill_gradient2(
    low = underestimate_colour,
    mid = "white",
    high = overestimate_colour,
    midpoint = 0,
    limits = area_effect_limits,
    oob = squish,
    labels = label_number(accuracy = area_effect_label_accuracy),
    na.value = missing_fill,
    name = "Effect",
    guide = colourbar_guide
  ) +
  plot_annotation(
    title = "Unstructured area effects",
    subtitle = "Posterior mean local area effect by method",
    theme = annotation_theme
  )
```
