---
title: "Direct estimation"
subtitle: "Survey design-based estimates and uncertainty"
---

<!-- ::: {.callout-note title="Theory in brief"}
For an area-time cell $(j,t)$, the target is a population mean or proportion:

$$
\mu_{jt} = \frac{1}{N_{jt}}\sum_{i=1}^{N_{jt}} y_{ijt}.
$$

If sampled unit $i$ has inclusion probability $\pi_{ijt}$ and design weight $w_{ijt}=1/\pi_{ijt}$, the Horvitz--Thompson ratio estimator gives the design-weighted direct estimate:

$$
\hat{\mu}^{w}_{jt} =
\frac{\sum_{i \in S_{jt}} w_{ijt} y_{ijt}}
{\sum_{i \in S_{jt}} w_{ijt}}.
$$
::: -->

```{r}
#| label: direct-setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(survey)
library(tibble)
library(scales)

base_path <- here::here("data")

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = MSOA21CD) |>
  filter(str_starts(area_id, "W")) |>
  arrange(area_id)

domain_frame <- read_csv(
  file.path(base_path, "clean", "domain_msoayear.csv"),
  show_col_types = FALSE
)
```

```{r}
#| label: direct-plot-config
#| echo: false
share_label <- label_percent(accuracy = 1)
variance_label <- label_number(accuracy = 0.001)
missing_fill <- "grey85"

map_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey30"),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

Direct estimation is the starting point for the SAE workflow. Each area-time estimate, or *MSOA-year cell* in this example, is calculated only from the survey respondents observed in that area and wave. For each *MSOA-year cell*, we calculate a Horvitz--Thompson weighted proportion using the respondent cycling outcome, matched survey weight, survey stratum, MSOA assignment, and wave year.

The output is a transparent survey-based baseline: a direct estimate, sampling variance, and usable sample size for each observed *MSOA-year cell*. These estimates show where the survey evidence is strong, sparse, missing, or unstable, helping us decide whether model-based small area estimation is needed.

## Prepare survey data

We use a Horvitz–Thompson weighted proportion among usable adult respondents. The cleaned survey microdata input has one row per respondent-wave for the cycling estimand. After filtering, each retained row has a valid binary cycling response, matched weight, MSOA assignment, wave label, and survey stratum:

| Column | Role |
|---|---|
| `area_id` | Estimation area |
| `time_id` | Survey wave / year |
| `freq_cyclist` | Binary response:<br>`1` if the adult cycles for transport frequently,<br>`0` otherwise |
| `weight` | Matched weight |
| `strata` | Local Authority code |
```{r}
#| label: direct-load

survey_bike <- read_csv(
  file.path(base_path, "clean", "surveyind_freqcyc.csv"),
  show_col_types = FALSE
) |>
  drop_na(freq_cyclist, weight, area_id, strata)

head(survey_bike)
```

## Calculate direct estimates

`svyby()` computes the weighted share and uncertainty diagnostics for every area-time cell with usable observations. The completed area-by-time grid retains missing cells for later modelling.

```{r}
#| label: direct-estimate
#| code-fold: false

# Define survey design
bike_design <- svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_bike
)

# Calculate direct estimate and uncertainty
direct_bike_estimates <- svyby(
  formula = ~freq_cyclist,
  by = ~area_id + time_id,
  design = bike_design,
  FUN = svymean,
  vartype = "se",
  na.rm = TRUE,
  keep.names = FALSE
) |>
  as_tibble() |>
  transmute(
    area_id,
    time_id,
    direct_estimate = freq_cyclist,
    variance = se^2
  )

# Calculate sample size per area-time cell
direct_bike_counts <- survey_bike |>
  group_by(area_id, time_id) |>
  summarise(n = n(), .groups = "drop")

# Join together
direct_bike <- domain_frame |>
  left_join(direct_bike_estimates, by = c("area_id", "time_id")) |>
  left_join(direct_bike_counts, by = c("area_id", "time_id")) |>
  mutate(n = replace_na(n, 0L))

head(direct_bike)
```

## Assess precision and coverage

Direct estimation provides a useful starting point because it makes three limitations visible:

- **Uneven temporal and spatial coverage:** The number of usable observations varies across survey waves and MSOAs, leaving some area-time cells with much less direct evidence than others.
- **Unstable area estimates over time:** For some MSOAs, direct estimates change sharply between waves, which may reflect real behavioural change or sampling noise.
- **Zero within-area variance:** For binary indicators, small samples can contain only one response value, producing zero or unrealistically low variances.

```{r}
#| label: direct-map
#| echo: false
#| fig-width: 16
#| fig-height: 12
#| out-width: "100%"
#| fig-align: "center"

# Plotting functions
base_map <- function(wave) {
  wave_direct <- direct_bike[direct_bike[["time_id"]] == wave, ]

  wave_data <- msoa_boundaries |>
    left_join(wave_direct, by = "area_id")

  ggplot(wave_data) +
    theme_void() +
    map_text_theme
}

plot_estimate <- function(wave) {
  base_map(wave) +
    geom_sf(aes(fill = direct_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = c(0, 1),
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    )
}

plot_uncertainty <- function(wave) {
  base_map(wave) +
    geom_sf(aes(fill = variance), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      na.value = missing_fill,
      limits = c(0, max(direct_bike$variance, na.rm = TRUE)),
      labels = variance_label,
      name = "Variance"
    )
}

plot_sample <- function(wave) {
  base_map(wave) +
    geom_sf(aes(fill = n), colour = "transparent") +
    scale_fill_gradient(
      low = "#fff8e1",
      high = "#4f3b10",
      limits = c(0, max(direct_bike$n, na.rm = TRUE)),
      na.value = missing_fill,
      name = "Sample size"
    )
}


# Generate and assemble plots
e_21 <- plot_estimate("2021") +
  ggtitle("2021") +
  ylab("Direct estimate") +
  theme(legend.position = "none")
e_23 <- plot_estimate("2023") +
  ggtitle("2023") +
  theme(legend.position = "none")
e_25 <- plot_estimate("2025") + ggtitle("2025")

u_21 <- plot_uncertainty("2021") +
  ylab("Variance") +
  theme(legend.position = "none")
u_23 <- plot_uncertainty("2023") + theme(legend.position = "none")
u_25 <- plot_uncertainty("2025")

s_21 <- plot_sample("2021") +
  ylab("Sample size") +
  theme(legend.position = "none")
s_23 <- plot_sample("2023") + theme(legend.position = "none")
s_25 <- plot_sample("2025")

diagnostic_panel <- wrap_plots(
  e_21, e_23, e_25,
  u_21, u_23, u_25,
  s_21, s_23, s_25,
  ncol = 3, nrow = 3
) +
  plot_annotation(
    title = "Share of adults who cycle for transport frequently",
    subtitle = "Direct estimates, sampling variance, and usable sample size by wave",
    theme = map_annotation_theme
  )

diagnostic_panel
```
Sparse and uneven survey coverage limits how far direct estimates can be used on their own. At the target MSOA-year scale, many estimates are unstable or unavailable, so the next step is to use model-based SAE methods that borrow strength across areas, years, and auxiliary predictors.

Two model-based options follow:

- **Area-level SAE models** stabilise direct MSOA-year estimates using sampling variances to control how much weight each direct estimate receives.
- **Unit-level SAE models** estimate the individual outcome from survey microdata and aggregate predictions to target MSOA-years.


```{r}
#| label: direct-summary
direct_summary_by_wave <- direct_bike |>
  group_by(time_id) |>
  summarise(
    `Usable observations` = sum(n),
    `MSOAs without direct estimates` = sum(is.na(direct_estimate)),
    `MSOAs without variance` = sum(is.na(variance)),
    `MSOAs with zero variance` = sum(variance == 0, na.rm = TRUE),
    `Median sample size` = median(n),
    .groups = "drop"
  ) |>
  rename(Wave = time_id)

direct_summary_by_wave
```

The choice between area-level and unit-level modelling should depend on two things: the quality of the direct estimates as model inputs, and whether the survey microdata contain unit-level predictors that help explain individual variation in the indicator.

In this setting, the summary points to important limitations for area-level modelling. Because the outcome is binary, many MSOA-year cells contain limited direct survey information: small binary samples often produce all-zero or all-one responses. This motivates prioritising unit-level modelling, where respondent characteristics can be used to model the outcome before predictions are aggregated back to areas. We still fit an [area-level model](area-sae-fh.qmd) first to show how much stabilisation is possible from direct-estimate inputs before moving to [unit-level modelling](unit-sae-summ.qmd).

```{r}
#| label: direct-export
#| code-fold: true
write_csv(direct_bike, file.path(base_path, "freqcyc_estimates_direct.csv"))
```
