---
title: "Direct Estimation"
subtitle: "Survey design-based estimates and uncertainty"
---

```{r}
#| label: setup-direct-estimation
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(survey)
library(tibble)
library(scales)

base_path <- here::here("data")
msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = all_of("MSOA21CD")) |>
  filter(str_starts(.data$area_id, "W"))

domain_frame <- read_csv(
  file.path(base_path, "clean", "domain_msoayear.csv"),
  show_col_types = FALSE
)
```

```{r}
#| label: plotting-config
#| echo: false
share_label <- label_percent(accuracy = 1)
variance_label <- label_number(accuracy = 0.001)
missing_fill <- "grey85"

map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)
```

Direct estimation is the starting point for the SAE workflow. At this stage, each area-time estimate, or an *MSOA-year cell* in this example, is calculated only from the survey respondents observed in that area and wave, using the matched survey weight. This gives us a transparent baseline before any model-based smoothing is introduced and shows where direct estimates are unstable, missing, or based on very small samples, helping us decide whether model-based small area estimation is needed.

::: {.callout-note title="Theory in brief"}
For an area-time cell $(j,t)$, the target is the population mean or proportion:

$$
\mu_{jt} = \frac{1}{N_{jt}}\sum_{i=1}^{N_{jt}} y_{ijt}.
$$

If respondent $i$ has inclusion probability $\pi_{ijt}$ and survey weight $w_{ijt}=1/\pi_{ijt}$, the Horvitz–Thompson ratio estimator is:

$$
\hat{\mu}^{w}_{jt} =
\frac{\sum_{i \in S_{jt}} w_{ijt} y_{ijt}}
{\sum_{i \in S_{jt}} w_{ijt}}.
$$
:::

<div class="method-divider" aria-hidden="true"></div>

{{< include _includes/target-estimand.qmd >}}

## Prepare survey data

We use a Horvitz–Thompson weighted proportion among usable adult respondents. The cleaned survey microdata input has one row per respondent-wave for the walking estimand. After filtering, each retained row has a valid binary walking response, matched weight, MSOA assignment, wave label, and survey stratum:

| Column | Role |
|---|---|
| `area_id` | Estimation area |
| `time_id` | Survey wave / year |
| `indicator` | Binary response: `1` if the adult walks for transport daily, `0` otherwise |
| `weight` | Matched weight |
| `strata` | Local Authority code |
```{r}
#| label: load-survey-indicator
survey_walk <- read_csv(
  file.path(base_path, "clean", "surveyind_freqwalk.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(.data$indicator),
    !is.na(.data$weight),
    !is.na(.data$area_id),
    !is.na(.data$strata)
  )
head(survey_walk)
```

## Calculate direct estimates

We use `svyby()` from the `survey` package to compute the weighted share and its uncertainty diagnostics for every area-time cell with usable observations. It then completes the area-by-time grid so missing cells are retained for modelling.

```{r}
#| label: direct-estimates-walking
#| code-fold: false

# Define survey design
walk_design <- svydesign(
  ids = as.formula("~1"),
  strata = as.formula("~strata"),
  weights = as.formula("~weight"),
  data = survey_walk
)

# Calculate direct estimate and uncertainty
direct_walk_estimates <- svyby(
  formula = as.formula("~indicator"),
  by = as.formula("~area_id + time_id"),
  design = walk_design,
  FUN = svymean,
  vartype = "se",
  na.rm = TRUE,
  keep.names = FALSE
) |>
  as_tibble() |>
  transmute(
    area_id,
    time_id,
    direct_estimate = .data$indicator,
    variance = .data$se^2
  )

# Calculate sample size per area-time cell
direct_walk_counts <- survey_walk |>
  group_by(across(all_of(c("area_id", "time_id")))) |>
  summarise(n = n(), .groups = "drop")

# Join
direct_walk <- domain_frame |>
  left_join(direct_walk_estimates, by = c(
    "area_id", "time_id"
  )) |>
  left_join(direct_walk_counts, by = c(
    "area_id", "time_id"
  )) |>
  mutate(n = replace_na(.data$n, 0L))

head(direct_walk)
```

## Assess precision and coverage

Direct estimation provides a useful starting point because it makes three limitations visible:

- **Uneven temporal and spatial coverage:** The number of usable observations varies across survey waves and MSOAs, leaving some area-time cells with much less direct evidence than others.
- **Unstable area estimates over time:** For some MSOAs, direct estimates change sharply between waves, which may reflect real behavioural change or sampling noise.
- **Zero within-area variance:** For binary indicators, small samples can contain only one response value, producing zero or unrealistically low variances.

```{r}
#| label: direct-map-diagnostics
#| echo: false
#| fig-width: 16
#| fig-height: 12
#| out-width: "100%"
#| fig-align: "center"
#| 
# Plotting functions
base_map <- function(wave) {
  wave_direct <- direct_walk[direct_walk[["time_id"]] == wave, ]

  wave_data <- msoa_boundaries |>
    left_join(wave_direct, by = "area_id")

  ggplot(wave_data) +
    theme_void() +
    map_text_theme
}

plot_estimate <- function(wave) {
  base_map(wave) +
    geom_sf(aes(fill = .data$direct_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = c(0, 1),
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    )
}

plot_uncertainty <- function(wave) {
  base_map(wave) +
    geom_sf(aes(fill = .data$variance), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      na.value = missing_fill,
      limits = c(0, max(direct_walk$variance, na.rm = TRUE)),
      labels = variance_label,
      name = "Variance"
    )
}

plot_sample <- function(wave) {
  base_map(wave) +
    geom_sf(aes(fill = .data$n), colour = "transparent") +
    scale_fill_gradient(
      low = "#fff8e1",
      high = "#4f3b10",
      limits = c(0, max(direct_walk$n, na.rm = TRUE)),
      na.value = missing_fill,
      name = "Sample size"
    )
}


# Generate and assemble plots
e_21 <- plot_estimate("2021") +
  ggtitle("2021") +
  ylab("Direct estimate") +
  theme(legend.position = "none")
e_23 <- plot_estimate("2023") +
  ggtitle("2023") +
  theme(legend.position = "none")
e_25 <- plot_estimate("2025") + ggtitle("2025")

u_21 <- plot_uncertainty("2021") +
  ylab("Variance") +
  theme(legend.position = "none")
u_23 <- plot_uncertainty("2023") + theme(legend.position = "none")
u_25 <- plot_uncertainty("2025")

s_21 <- plot_sample("2021") +
  ylab("Sample size") +
  theme(legend.position = "none")
s_23 <- plot_sample("2023") + theme(legend.position = "none")
s_25 <- plot_sample("2025")

diagnostic_panel <- wrap_plots(
  e_21, e_23, e_25,
  u_21, u_23, u_25,
  s_21, s_23, s_25,
  ncol = 3, nrow = 3
) +
  plot_annotation(
    title = "Share of adults who walk for transport daily",
    subtitle = "Direct estimates, sampling variance, and usable sample size by wave",
    theme = map_annotation_theme
  )

diagnostic_panel
```



```{r}
#| label: summarise-direct-estimates
direct_summary_by_wave <- direct_walk |>
  group_by(.data$time_id) |>
  summarise(
    `Usable observations` = sum(.data$n),
    `MSOAs without direct estimates` = sum(is.na(.data$direct_estimate)),
    `MSOAs with zero variance` = sum(.data$variance == 0, na.rm = TRUE),
    `Median sample size` = median(.data$n),
    `Median direct estimate` = round(
      median(.data$direct_estimate, na.rm = TRUE),
      3
    ),
    `Median variance` = round(
      median(.data$variance, na.rm = TRUE),
      5
    ),
    .groups = "drop"
  ) |>
  rename(`Wave` = all_of("time_id"))

direct_summary_by_wave
```

In case of unstable or unavailable direct estimates at the target area-time scale as seen above in this example, we turn to model-based SAE methods which borrow strength across areas and years and auxiliary predictors. There are two large families of SAE methods: **Area-level SAE** models stabilise direct area estimates, while **unit-level SAE models** estimate the individual outcome from survey records and aggregate predictions to target area-years.

```{r}
#| label: save-direct-estimates
#| echo: false
write_csv(
  direct_walk,
  file.path(base_path, "freqwalk_estimates_direct.csv")
)
```
