---
title: "Unit-Level SAE Modelling with Poststratification (a.k.a MRP)"
subtitle: "A Bayesian spatio-temporal extension"
---

::: {.callout-note title="Theory in brief"}

Let $c$ index poststratification cells within area $j$ and time $t$. The cells are defined by whichever population variables are available for poststratification. In a unit-level MRP approach, a binary walking outcome can be modelled at the respondent level as

$$
\begin{aligned}
Y_{ijt} &\sim \mathrm{Bernoulli}(\pi_{ijt}) \\
\mathrm{logit}(\pi_{ijt}) &=
\alpha + z_{ijt}^{T}\beta + u_j + v_t .
\end{aligned}
$$

where $z_{ijt}$ is a vector of respondent-level predictors used to define or adjust the poststratification cells, $\beta$ contains their coefficients, $u_j$ is an area effect, and $v_t$ is a time effect. The fitted model gives predicted cell means for the binary frequent-walking indicator in each poststratification cell:

$$
\hat{\pi}_{jtc}
=
\widehat{\mathbb{E}}(Y \mid j,t,c).
$$

The poststratification step combines those cell predictions with population counts $N_{jtc}$ available or estimated for the same cells. Area-time estimates are then population-weighted averages of the cell predictions.
$$
\hat{\mu}_{jt}^{\mathrm{MRP}}
=
\frac{\sum_c N_{jtc}\hat{\pi}_{jtc}}
{\sum_c N_{jtc}}.
$$
:::

```{r}
#| label: setup-mrp
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(spdep)
library(INLA)
library(tibble)
library(purrr)
library(scales)

base_path <- here::here("data")
domain_frame <- read_csv(
  file.path(base_path, "clean", "domain_msoayear.csv"),
  show_col_types = FALSE
)
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = all_of("MSOA21CD")) |>
  filter(str_starts(area_id, "W"))
```

```{r}
#| label: plotting-config-mrp
#| echo: false
missing_fill <- "grey85"

map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

MRP extends the unit-level SAE workflow by combining a respondent-level outcome model with a poststratification table describing the target population. For our estimand, we model the binary frequent-walking indicator using `age_band` (to reflect the age composition in the poststratification table), `sex`, `no_car`, and spatial and temporal random effects. We then produce predicted cell means for the binary frequent-walking indicator for each `age_band` × `sex` × `no_car` population cell and aggregate them to MSOA-year estimates using population weights. Uncertainty is propagated by repeating this aggregation across posterior draws.

One data limitation is that the available ONS poststratification table only contains `age_band` × `sex` population counts. To include `no_car`, we first estimate no-car probabilities for each `age_band` × `sex` cell and use them to extend the poststratification table. This avoids applying the same MSOA-level no-car share to every age-sex cell, but it introduces additional uncertainty because part of the population table is modelled rather than directly observed. A more conservative alternative would be to model walking using only `age_band` and `sex`, and use MSOA-level no-car share separately as a contextual area-level predictor.

## Prepare modelling and poststratification data

As in previous methods, we first create consistent numeric indices for each MSOA and survey year, and a Queen-contiguity adjacency graph for the spatial effect.

```{r}
#| label: define-mrp-model-index
#| code-fold: false
#|
area_levels <- sort(unique(domain_frame[["area_id"]]))
time_levels <- sort(unique(domain_frame[["time_id"]]))

area_index <- tibble(
  area_id = area_levels,
  area_inla_id = seq_along(area_levels)
)

time_index <- tibble(
  time_id = time_levels,
  time_inla_id = seq_along(time_levels)
)
```

```{r}
#| label: build-mrp-spatial-graph
#| code-fold: false
#|
msoa_boundaries_sorted <- msoa_boundaries |>
  inner_join(area_index, by = "area_id") |>
  arrange(area_inla_id)

msoa_neighbours <- poly2nb(
  msoa_boundaries_sorted,
  queen = TRUE,
  row.names = msoa_boundaries_sorted$area_id
)

spatial_graph_file <- tempfile(fileext = ".adj")

nb2INLA(
  file = spatial_graph_file,
  nb = msoa_neighbours
)
```

The MRP workflow needs two inputs:

- `model_data`, the sampled respondent records used to fit the model;
- `poststrat_data`, the population cells over which predictions are averaged.

### Respondent model data

The survey file contains respondent-level age, sex, no-car access, and walking outcome fields.

It is essential for the unit-level predictors to be identical to the ones present in the poststratification table. Therefore, age is grouped into the same five-year bands and sex is also recoded as Male and Female.
```{r}
#| label: load-mrp-survey-indicator
walking_indicator <- read_csv(
  file.path(base_path, "clean", "surveyind_freqwalk.csv"),
  show_col_types = FALSE
)

age_band_levels <- c(
  "age_15_19", "age_20_24", "age_25_29", "age_30_34",
  "age_35_39", "age_40_44", "age_45_49", "age_50_54",
  "age_55_59", "age_60_64", "age_65_69", "age_70_74",
  "age_75_79", "age_80_84", "age_85_89", "age_90_plus"
)

walking_indicator_recoded <- walking_indicator |>
  mutate(
    age_band = cut(
      age,
      breaks = c(seq(15, 90, by = 5), Inf),
      labels = age_band_levels,
      right = FALSE
    ),
    age_band = factor(age_band, levels = age_band_levels),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))
  )

model_data <- walking_indicator_recoded |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  drop_na(indicator, area_id, time_id, age_band, sex, no_car, time_inla_id, area_inla_id) |>
  select(
    unit_id,
    area_id,
    time_id,
    indicator,
    age_band,
    sex,
    no_car,
    time_inla_id,
    area_inla_id
  )

head(model_data)
```

### Population poststratification data

First, we load the age-sex poststratification table and align its `age_band` and `sex` categories with the survey model data.

```{r}
#| label: load-poststratification-data

poststrat_age_sex <- read_csv(
  file.path(base_path, "clean", "poststrat_freqwalk.csv"),
  show_col_types = FALSE
) |>
  transmute(
    area_id,
    msoa_name,
    time_id,
    age_band = factor(age_band, levels = age_band_levels),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    pop_count
  ) |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  filter(pop_count > 0) |>
  drop_na(age_band, sex, pop_count, time_inla_id, area_inla_id)

head(poststrat_age_sex)
```

The poststratification table has age-sex population counts, but it does not contain the joint distribution of age, sex, and car access. We first model respondent no-car access using the variables that are available in the poststratification table. We then predict the probability of `no_car = 1` for each age-sex-area-time population cell and use those probabilities to expand the poststratification table.

```{r}
#| label: estimate-no-car-probability
no_car_extension_data <- bind_rows(
  model_data |>
    transmute(
      row_type = "model",
      no_car,
      age_band,
      sex,
      time_inla_id,
      area_inla_id
    ),
  poststrat_age_sex |>
    transmute(
      row_type = "poststrat",
      no_car = NA_real_,
      age_band,
      sex,
      time_inla_id,
      area_inla_id
    )
)

no_car_extension_formula <- no_car ~
  age_band +
  sex +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

no_car_extension_fit <- inla(
  formula = no_car_extension_formula,
  data = no_car_extension_data,
  family = "binomial",
  Ntrials = 1,
  control.predictor = list(compute = TRUE, link = 1)
)

poststrat_age_sex_extended <- poststrat_age_sex |>
  mutate(
    no_car_prob = no_car_extension_fit$summary.fitted.values$mean[
      no_car_extension_data$row_type == "poststrat"
    ]
  )

poststrat_age_sex_extended |>
  select(area_id, time_id, age_band, sex, no_car_prob) |>
  arrange(area_id, time_id, age_band, sex) |>
  head()
```
With the modelled `no_car_prob` in place for the original poststratification table, we calculate the population count for the extended table now with an additional categorical variable.
```{r}
#| label: prepare-poststratification-frame
poststrat_data <- bind_rows(
  poststrat_age_sex_extended |>
    mutate(no_car = 1, pop_count = pop_count * no_car_prob),
  poststrat_age_sex_extended |>
    mutate(no_car = 0, pop_count = pop_count * (1 - no_car_prob))
) |>
  select(
    area_id,
    msoa_name,
    time_id,
    age_band,
    sex,
    no_car,
    pop_count,
    time_inla_id,
    area_inla_id
  ) |>
  arrange(time_id, area_inla_id, age_band, sex, no_car)

head(poststrat_data)
```

Note that although we have the share of households with no car by MSOA as an area predictor this approach avoids assuming that the MSOA-level no-car share is constant across every age-sex cell. The trade-off is that the new poststratification table now contains a modelled variable instead of being based on the ground-truth that is observed joint Census cross-tabulation.

## Fit and evaluate models

We fit a sequence of binomial unit-level models. The outcome is the respondent walking indicator, and fitted values for the poststratification rows are returned as response-scale cell means.

| Model | Name | Specification |
|---|---|---|
| M0 | Baseline | Temporal RW1 plus BYM2 spatial MSOA effect |
| M1 | +Demographic | M0 plus age band and sex |
| M2 | +Nocar | M1 plus no-car |

```{r}
#| label: define-mrp-models
#| code-fold: false

m0_formula <- indicator ~
  1 +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m1_formula <- indicator ~
  age_band +
  sex +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m2_formula <- indicator ~
  age_band +
  sex +
  no_car +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

model_formulas <- list(
  "M0 Baseline" = m0_formula,
  "M1 +Demographic" = m1_formula,
  "M2 +Nocar" = m2_formula
)
```

The observed respondent rows and the poststratification rows are combined before fitting. Rows with `indicator = NA` are not used in the likelihood, but `INLA` still returns their fitted values, which are needed for poststratification.

```{r}
#| label: fit-mrp-models
#| code-fold: false

model_poststrat_data <- bind_rows(
  model_data |>
    transmute(
      row_type = "model",
      indicator,
      age_band,
      sex,
      no_car,
      time_inla_id,
      area_inla_id
    ),
  poststrat_data |>
    transmute(
      row_type = "poststrat",
      indicator = NA_real_,
      age_band,
      sex,
      no_car,
      time_inla_id,
      area_inla_id
    )
)

fit_mrp_model <- function(formula) {
  inla(
    formula = formula,
    family = "binomial",
    Ntrials = 1,
    data = model_poststrat_data,
    control.predictor = list(compute = TRUE, link = 1),
    control.compute = list(waic = TRUE, config = TRUE)
  )
}

mrp_fits <- lapply(model_formulas, fit_mrp_model)
```

As before, we use `WAIC` as an initial model check. M2 demonstrates the lowest WAIC, meaning the age, sex and no-car access predictors together best explain the variablity among the binary outcome being modelled.

```{r}
#| label: compare-mrp-waic

model_waic <- imap_dfr(
  mrp_fits,
  \(fit, model_name) {
    tibble(
      model = model_name,
      waic = fit$waic$waic,
      eff_params = fit$waic$p.eff
    )
  }
) |>
  arrange(waic)

model_waic
```

We can also expose the coefficients for all fixed effect terms for M2 to check whether the modelled relationships are logical. Here, we see that the older the respondent is the less likely they are to walk daily as main transport (reference category is `age_15_19`), whereas those without access to a car or a van are more likely.

```{r}
#| label: mrp-no-car-coefficient

mrp_fits[["M2 +Nocar"]]$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  transmute(
    term,
    mean = pp_label(mean),
    lower = pp_label(.data[["0.025quant"]]),
    upper = pp_label(.data[["0.975quant"]])
  )
```

At this point, having selected our final model, we use poststratification to convert its cell-level predictions into the target MSOA-year estimates. That means, each `age_band` x `sex` x `no_car` cell prediction is weighted by its population count, then aggregated within each MSOA-year. One the one hand, we can apply this calculation using the posterior mean cell values returned above, which gives a simple point-estimate for each cell:

```{r}
#| label: poststratify-mrp-point-estimate

mrp_final_fit <- mrp_fits[["M2 +Nocar"]]
poststrat_cell_rows <- which(model_poststrat_data$row_type == "poststrat")

cell_mean_predictions <- mrp_final_fit$summary.fitted.values |>
  as_tibble() |>
  slice(poststrat_cell_rows) |>
  transmute(cell_mean = mean)

cell_point_estimates <- bind_cols(poststrat_data, cell_mean_predictions)

area_year_point_estimates <- cell_point_estimates |>
  group_by(area_id, time_id) |>
  summarise(
    weighted_cell_sum = sum(cell_mean * pop_count),
    population_total = sum(pop_count),
    .groups = "drop"
  ) |>
  mutate(mrp_point_estimate = weighted_cell_sum / population_total) |>
  select(area_id, time_id, mrp_point_estimate)

head(area_year_point_estimates)
```

However, the final output for SAE also needs uncertainty estimates. To propagate uncertainty through poststratification, we repeat the same population-weighted aggregation across posterior draws. This means the credible intervals describe uncertainty in the aggregated MSOA-year estimates, not just uncertainty in the individual cell predictions. The posterior draws used for this calculation are shown below.

## Final estimates and uncertainty

The function `inla.posterior.sample()` gives plausible versions of the fitted model; for each draw, we keep the poststratification rows and convert their fitted values from the logit scale to the original scale of the outcome (0-1).

```{r}
#| label: sample-mrp-posterior
#| code-fold: false

# Draw plausible versions of the fitted model from the posterior.
set.seed(100)
posterior_draws <- inla.posterior.sample(n = 300, result = mrp_final_fit)

# Rows are poststrat cells; columns are posterior draws; values are predicted cell means.
estimate_draws <- matrix(NA_real_, nrow = nrow(poststrat_data), ncol = length(posterior_draws))

predictor_row_index <- startsWith(rownames(posterior_draws[[1]]$latent), "Predictor")

for (draw_id in seq_along(posterior_draws)) {
  draw <- posterior_draws[[draw_id]]
  row_prediction_logits <- as.numeric(draw$latent[predictor_row_index, 1])
  poststrat_logits <- row_prediction_logits[poststrat_cell_rows]
  poststrat_cell_means <- 1 / (1 + exp(-poststrat_logits))
  estimate_draws[, draw_id] <- poststrat_cell_means
}
```

The matrix `estimate_draws` contains the draw-level predicted cell means. Using it, we poststratify each posterior draw. That means, for every draw, the predictions are weighted by the corresponding population counts, summed within each MSOA-year, and divided by the total MSOA-year population. This gives one population-weighted estimate per MSOA-year for each posterior draw.

```{r}
#| label: poststratify-mrp-posterior
#| code-fold: false

area_year_id <- paste(poststrat_data$area_id, poststrat_data$time_id, sep = "__")

# Population-weight cell means and aggregate them to MSOA-year estimates.
area_year_weighted_sums <- rowsum(estimate_draws * poststrat_data$pop_count, group = area_year_id, reorder = FALSE)
area_year_populations <- rowsum(poststrat_data$pop_count, group = area_year_id, reorder = FALSE)

area_year_means <- area_year_weighted_sums / as.numeric(area_year_populations)
area_year_means[1:5, 1:10]
```

Since each row of `area_year_means` is the posterior distribution for one MSOA-year estimate, we summarise those into a posterior mean, posterior variance, and 95% credible interval.

```{r}
#| label: summarise-mrp-final-estimates
#| code-fold: false

area_year_summary <- tibble(
  area_year_id = rownames(area_year_means),
  mrp_estimate = rowMeans(area_year_means),
  mrp_variance = apply(area_year_means, 1, var),
  mrp_lower = apply(area_year_means, 1, quantile, probs = 0.025, names = FALSE),
  mrp_upper = apply(area_year_means, 1, quantile, probs = 0.975, names = FALSE)
) |>
  mutate(mrp_ci_width = mrp_upper - mrp_lower) |>
  separate(area_year_id, into = c("area_id", "time_id"), sep = "__", convert = TRUE)

final_mrp_results <- domain_frame |>
  select(area_id, msoa_name, time_id) |>
  left_join(area_year_summary, by = c("area_id", "time_id"))

head(final_mrp_results)
```

As with previous methods, we report the MRP estimates alongside the direct estimates and variance outputs in order to provide a full view of the direct survey evidence that accompanies the modelled estimates.

```{r}
#| label: extract-mrp-final-results

direct_estimates <- read_csv(
  file.path(base_path, "freqwalk_estimates_direct.csv"),
  show_col_types = FALSE
) |>
  select(area_id, time_id, n, direct_estimate, variance)

final_mrp_results <- final_mrp_results |>
  left_join(direct_estimates, by = c("area_id", "time_id")) |>
  mutate(
    direct_ci_width = 2 * qnorm(0.975) * sqrt(pmax(variance, 0))
  ) |>
  select(
    area_id,
    msoa_name,
    time_id,
    n,
    direct_estimate,
    variance,
    direct_ci_width,
    mrp_estimate,
    mrp_variance,
    mrp_lower,
    mrp_upper,
    mrp_ci_width
  )

head(final_mrp_results)
```

```{r}
#| label: map-mrp-final-results
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

final_mrp_map_data <- msoa_boundaries |>
  left_join(final_mrp_results, by = "area_id")
mrp_estimate_limits <- range(final_mrp_results$mrp_estimate, na.rm = TRUE)
mrp_ci_width_limits <- c(0, max(final_mrp_results$mrp_ci_width, na.rm = TRUE))
mrp_wave_years <- sort(unique(final_mrp_results$time_id))

plot_mrp_estimate <- function(wave) {
  final_mrp_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = mrp_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = mrp_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_mrp_ci_width <- function(wave) {
  final_mrp_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = mrp_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = mrp_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(mrp_wave_years, plot_mrp_estimate)
uncertainty_maps <- lapply(mrp_wave_years, plot_mrp_ci_width)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(mrp_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = 2) +
  plot_annotation(
    title = "Share of adults who walk for transport daily",
    subtitle = "MRP final model and 95% credible interval width",
    theme = map_annotation_theme
  )
```

### Comparison diagnostics

The comparison plot shows that the MRP estimates are very close to the BHF estimates, with points clustered around the equality line. Agreement with the direct estimates is much weaker as expected, reflecting the instability of sparse MSOA-year survey cells. The Fay--Herriot estimates tend to sit lower than MRP for many cells, suggesting stronger shrinkage toward lower area-year rates.

```{r}
#| label: add-mrp-comparison-estimates
bhf_final_estimates <- read_csv(
  file.path(base_path, "freqwalk_estimates_bhf.csv"),
  show_col_types = FALSE
) |>
  select(
    area_id,
    time_id,
    bhf_estimate,
    bhf_variance,
    bhf_ci_width
  )

fh_final_estimates <- read_csv(file.path(base_path, "freqwalk_estimates_fh.csv"), show_col_types = FALSE) |>
  select(area_id, time_id, fh_estimate, fh_variance, fh_ci_width)

final_mrp_results <- final_mrp_results |>
  left_join(bhf_final_estimates, by = c("area_id", "time_id")) |>
  left_join(fh_final_estimates, by = c("area_id", "time_id")) |>
  mutate(mrp_minus_bhf = mrp_estimate - bhf_estimate) |>
  select(
    area_id,
    msoa_name,
    time_id,
    n,
    direct_estimate,
    variance,
    direct_ci_width,
    bhf_estimate,
    bhf_variance,
    bhf_ci_width,
    fh_estimate,
    fh_variance,
    fh_ci_width,
    mrp_estimate,
    mrp_variance,
    mrp_lower,
    mrp_upper,
    mrp_ci_width,
    mrp_minus_bhf
  )

head(final_mrp_results)
```
```{r}
#| label: plot-mrp-comparison-scatter
#| echo: false
#| fig-height: 6
#| fig-width: 14

mrp_scatter_comparison <- final_mrp_results |>
  select(
    area_id,
    time_id,
    n,
    mrp_estimate,
    direct_estimate,
    bhf_estimate,
    fh_estimate
  ) |>
  pivot_longer(
    cols = c(direct_estimate, bhf_estimate, fh_estimate),
    names_to = "comparison",
    values_to = "comparison_estimate"
  ) |>
  mutate(
    comparison = recode(
      comparison,
      direct_estimate = "Direct estimates",
      fh_estimate = "Fay-Herriot estimates",
      bhf_estimate = "BHF estimates"
    )
  ) |>
  filter(!is.na(comparison_estimate))

mrp_scatter_limits <- c(0, 0.5)

mrp_scatter_comparison |>
  ggplot(aes(x = comparison_estimate, y = mrp_estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(aes(size = n), alpha = 0.55, colour = "#1f4e5f") +
  facet_wrap(vars(comparison), nrow = 1) +
  scale_x_continuous(labels = share_label, limits = mrp_scatter_limits, oob = squish) +
  scale_y_continuous(labels = share_label, limits = mrp_scatter_limits, oob = squish) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  coord_equal() +
  labs(
    title = "Estimate comparison",
    subtitle = "Dashed line shows equality",
    x = "Comparison estimate",
    y = "MRP estimate"
  ) +
  theme_minimal(base_size = 12) +
  chart_text_theme +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

A closer look at the difference between the unit-level BHF and MRP estimates reveals subtle differences in how population composition is handled. The BHF workflow aggregates using area-level summaries, so it cannot fully represent the internal age, sex, and car-access composition of each MSOA-year. MRP instead predicts over the `age_band` x `sex` x `no_car` poststratification cells and then weights those cells by their population counts.

MRP estimates should therefore be higher than BHF estimates where the poststratification table gives more weight to groups with higher predicted frequent-walking rates. In this scenario, those groups are mainly younger adults and people without car access. Where these groups make up a larger share of a local population than the BHF area-level summaries can represent, MRP will tend to shift the estimate upward.

```{r}
#| label: map-mrp-bhf-comparison
#| echo: false
#| fig-width: 16
#| fig-height: 6
#| out-width: "100%"

mrp_bhf_difference_limit <- quantile(abs(final_mrp_results$mrp_minus_bhf), 0.95, na.rm = TRUE, names = FALSE)
mrp_bhf_difference_limits <- c(-mrp_bhf_difference_limit, mrp_bhf_difference_limit)

final_mrp_comparison_map_data <- msoa_boundaries |>
  left_join(final_mrp_results, by = "area_id")

plot_mrp_bhf_difference <- function(wave) {
  final_mrp_comparison_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = mrp_minus_bhf), colour = "transparent") +
    scale_fill_gradient2(
      low = "#2166ac",
      mid = "white",
      high = "#b2182b",
      midpoint = 0,
      limits = mrp_bhf_difference_limits,
      oob = squish,
      labels = pp_label,
      na.value = missing_fill,
      name = "MRP - BHF"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

bhf_difference_maps <- lapply(mrp_wave_years, plot_mrp_bhf_difference)

bhf_difference_maps[[1]] <- bhf_difference_maps[[1]] + ylab("MRP - BHF")

for (i in seq_len(length(mrp_wave_years) - 1)) {
  bhf_difference_maps[[i]] <- bhf_difference_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(bhf_difference_maps, ncol = 3) +
  plot_annotation(
    title = "Poststratification diagnostics",
    subtitle = "Difference from BHF estimates",
    theme = map_annotation_theme
  )
```


```{r}
#| label: export-mrp-results
#| code-fold: true

write_csv(final_mrp_results, file.path(base_path, "freqwalk_estimates_mrp.csv"))
```
