---
title: "Unit-level SAE model with area summaries"
subtitle: "A Bayesian spatio-temporal extension of a unit-level GLMM with area-summary prediction"
---

<!--
::: {.callout-note title="Theory in brief"}
Unit-level SAE models the individual outcome before aggregating predictions to area-time estimates. For a continuous outcome, the unit-level model can be written as:

$$
Y_{ijt}
=
\alpha + x_{ijt}^{T}\beta + u_j + v_t + \epsilon_{ijt}.
$$

Here, $x_{ijt}$ contains unit-level predictors, $u_j$ is an area effect, $v_t$ is a time effect, and $\epsilon_{ijt}$ is residual error. For a binary outcome, the model predicts the probability that $Y_{ijt}=1$.

Population summaries of the predictors $\bar{x}_{jt}$ are substituted into the fitted unit-level model:

$$
\hat{\mu}_{jt}
=
\widehat{\mathbb{E}}(Y \mid j, t, \bar{x}_{jt}).
$$
::: -->


```{r}
#| label: bhf-setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(INLA)
library(tibble)
library(purrr)
library(scales)

base_path <- here::here("data")

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = MSOA21CD) |>
  filter(str_starts(area_id, "W")) |>
  arrange(area_id)

domain_frame <- read_csv(
  file.path(base_path, "clean", "domain_msoayear.csv"),
  show_col_types = FALSE
)

spatial_graph_file <- file.path(base_path, "clean", "msoa_queen.adj")
```

```{r}
#| label: bhf-plot-config
#| echo: false
variance_label <- label_number(accuracy = 0.0001)
missing_fill <- "grey85"
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")
map_text_theme <- theme(
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

Unit-level SAE models survey microdata directly instead of starting from pre-aggregated direct estimates. This is useful when direct estimates are too sparse or unstable to provide reliable area-level modelling inputs. Inspired by the Battese--Harter--Fuller model, this approach uses a generalised linear mixed model to model the individual outcome, whether continuous or binary, using predictors available in the survey microdata. Final small area estimates are then produced by applying the fitted model to area-level summaries of those predictors.

Here, the binary frequent-cycling indicator `freq_cyclist` is modelled from the survey microdata using temporal and spatial random effects, with unit-level `age`, `sex`, and `caraccess` added across model specifications. The predictor summaries are extracted from [ONS UK Census 2021](https://www.ons.gov.uk/census) to produce the final *MSOA-year* estimates. Note that aggregating using area summaries of predictors assumes the area summaries of `age`, `sex`, and `caraccess` adequately represent the target population composition in each *MSOA-year* cell.

## Prepare model and aggregation data

There are two inputs when performing unit-level SAE:

- `model_data` contains the survey microdata used to fit the model.
- `aggregation_data` contains one row for each area-time cell we want to estimate.

```{r}
#| label: bhf-model-data

survey_microdata <- read_csv(
  file.path(base_path, "clean", "surveyind_freqcyc.csv"),
  show_col_types = FALSE
)

model_data <- survey_microdata |>
  transmute(
    unit_id,
    area_id,
    time_id,
    freq_cyclist,
    age,
    female = as.integer(sex == 2),
    caraccess
  )

head(model_data)
```

```{r}
#| label: bhf-agg-data

area_predictors <- read_csv(
  file.path(base_path, "clean", "areapred_freqcyc.csv"),
  show_col_types = FALSE
) |>
  select(
    area_id,
    age = age_adult_mean,
    female = female_adult_share,
    caraccess = caraccess_adult_share
  )

aggregation_data <- domain_frame |>
  left_join(area_predictors, by = "area_id") |>
  select(area_id, msoa_name, time_id, age, female, caraccess) |>
  arrange(time_id, area_id)

head(aggregation_data)
```

::: {.callout-note}
The aggregation data contains the complete area-time grid we want to estimate, and the area-level mean of the unit-level predictors. These area means come from Census 2021 and are used as a time-stable MSOA predictor across the three survey waves. This is reasonable only if auxiliary variables are unlikely to change quickly.
:::

Before fitting the model, we add the `INLA` time and area indices to the prepared datasets.

```{r}
#| label: bhf-add-ids
#| code-fold: false

area_index <- domain_frame |>
  distinct(area_id, area_inla_id)

time_index <- domain_frame |>
  distinct(time_id, time_inla_id)

model_data <- model_data |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  arrange(time_inla_id, area_inla_id)

aggregation_data <- aggregation_data |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  arrange(time_inla_id, area_inla_id)
```

## Fit and evaluate models

We fit binomial unit-level models to the survey microdata, then use them to predict area-time means. We specify three unit-level models to evaluate at the next step:

| Model | Name | Specifications |
|---|---|---|
| M1 | +TempSpatial | Temporal RW1 plus BYM2 spatial MSOA effect  |
| M2 | +Demographic | M1 plus respondent age and sex |
| M3 | +CarAccess | M2 plus respondent-level car access |

```{r}
#| label: bhf-specs
#| code-fold: false

m1_formula <- freq_cyclist ~
  1 +
  f(time_inla_id, model = "rw1", constr = TRUE) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE,
    constr = TRUE
  )

m2_formula <- update(m1_formula, . ~ . +
  age +
  female
)

m3_formula <- update(m2_formula, . ~ . + caraccess)

model_formulas <- list(
  "M1 +TempSpatial" = m1_formula,
  "M2 +Demographic" = m2_formula,
  "M3 +CarAccess" = m3_formula
)
```

```{r}
#| label: bhf-fit
#| code-fold: false

fit_bhf_model <- function(formula) {
  inla(
    formula = formula,
    family = "binomial",
    data = model_data,
    control.compute = list(waic = TRUE, config = TRUE)
  )
}

bhf_fits <- lapply(model_formulas, fit_bhf_model)
```

:::{.callout-note}
When fitting unit-level SAE models with `INLA`, `control.compute = list(waic = TRUE, config = TRUE)` requests WAIC for comparing model fit and stores the configuration needed for posterior sampling.
:::

To compare the fitted models' overall performance, we again use `WAIC` as a first check. Here, the car-access specification has the strongest WAIC support among the model specifications.
```{r}
#| label: bhf-waic

model_waic <- imap_dfr(
  bhf_fits,
  \(fit, model_name) {
    tibble(
      model = model_name,
      waic = fit$waic$waic,
      eff_params = fit$waic$p.eff
    )
  }
) |>
  arrange(waic)

model_waic

bhf_best_model <- model_waic |>
  slice(1) |>
  pull(model)
bhf_selected_fit <- bhf_fits[[bhf_best_model]]
cat("Best fitted model spec: ", bhf_best_model)
```
The fitted model summary reports the estimated coefficients and their uncertainty intervals.
```{r}
#| label: bhf-fixed

bhf_fits[[bhf_best_model]]$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  transmute(
    term,
    mean,
    lower = `0.025quant`,
    upper = `0.975quant`
  )
```

Finally, the **spatial and temporal random-effect** hyperparameters `tau` and `phi` help show what the temporal and spatial terms are doing in the selected model.

```{r}
#| label: bhf-hyper

bhf_fits[[bhf_best_model]]$summary.hyperpar |>
  as_tibble(rownames = "parameter") |>
  filter(str_detect(parameter, "time_inla_id|area_inla_id|Phi")) |>
  select(parameter, mean, sd, `0.025quant`, `0.5quant`, `0.975quant`) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

The random-effect table reports the posterior summaries for the temporal and spatial components in the selected unit-level specification. Higher precision means less residual variation in that component, while the BYM2 `phi` indicates how much of the area effect is spatially structured.

## Estimation and aggregation with posterior draws

The aggregation step uses posterior draws from the selected `INLA` fit to produce unit-level SAE estimates for each MSOA-year cell. We first sample from the posterior and extract the fixed effects, spatial area effect, and temporal effect. We then rebuild the fitted model for every MSOA-year aggregation row and summarise the resulting draws.

:::{.callout-note}
Each posterior draw stores the fixed effects, spatial area effect, and temporal effect together in `INLA`'s latent vector. Fixed effects are labelled as `term:1`, while random effects are labelled as `effect:index`.
:::

```{r}
#| label: bhf-draws
#| code-fold: false

set.seed(100)
posterior_draws <- inla.posterior.sample(
  n = 300,
  result = bhf_selected_fit
)

# Fixed-effect draw matrix
fixed_terms <- rownames(bhf_selected_fit$summary.fixed)
fixed_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0(fixed_terms, ":1"), 1]),
    fixed_terms
  )
})

# Area-effect draw matrix
area_ids <- sort(area_index$area_inla_id)
area_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("area_inla_id:", area_ids), 1]),
    area_ids
  )
})

# Time-effect draw matrix
time_ids <- sort(time_index$time_inla_id)
time_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("time_inla_id:", time_ids), 1]),
    time_ids
  )
})
```

Each posterior draw is evaluated over the MSOA-year aggregation frame by plugging in the area-level predictor summaries and the corresponding area and time effects.

::: {.callout-note title="Area summaries and binary outcomes"}
For **continuous outcomes**, such as commute time in minutes, substituting area-level predictor summaries into a linear model gives the same result as averaging individual predictions:

$$
\bar{x}\beta
=
\frac{1}{N}\sum_{i=1}^{N} x_i\beta
$$

For **binary outcomes**, such as `freq_cyclist` (`1` = yes, `0` = no), the target is the population proportion in the area, or the average predicted probability across individuals. Plugging in area-level summaries instead gives the predicted probability for an “average” individual, so it is only an approximation to the target.

$$
\operatorname{logit}^{-1}(\bar{x}\beta)
\approx
\frac{1}{N}\sum_{i=1}^{N}
\operatorname{logit}^{-1}(x_i\beta)
$$
:::

```{r}
#| label: bhf-predict
#| code-fold: false

prediction_matrix <- model.matrix(
  ~ age + female + caraccess,
  data = aggregation_data
)[, fixed_terms, drop = FALSE]

prediction_draws <- sapply(seq_along(posterior_draws), \(draw_id) {
  predictions <- as.numeric(prediction_matrix %*% fixed_draws[, draw_id]) +
    area_draws[as.character(aggregation_data$area_inla_id), draw_id] +
    time_draws[as.character(aggregation_data$time_inla_id), draw_id]

  # Convert log-odds to probabilities.
  plogis(predictions)
})
```

The matrix `prediction_draws` now has one MSOA-year estimate per row and one posterior draw per column. Posterior means, variances, and 95% credible intervals are computed directly from this draw matrix.

```{r}
#| label: bhf-results
#| code-fold: false

draw_quantile <- \(p) {
  apply(prediction_draws, 1, quantile, probs = p, names = FALSE)
}

final_bhf_results <- aggregation_data |>
  select(area_id, time_id) |>
  bind_cols(
    tibble(
      bhf_estimate = rowMeans(prediction_draws),
      bhf_variance = apply(prediction_draws, 1, var),
      bhf_lower = draw_quantile(0.025),
      bhf_upper = draw_quantile(0.975)
    )
  ) |>
  mutate(
    bhf_ci_width = bhf_upper - bhf_lower
  )

head(final_bhf_results)
```

```{r}
#| label: bhf-map
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

final_bhf_map_data <- msoa_boundaries |>
  left_join(final_bhf_results, by = "area_id")
bhf_estimate_limits <- range(final_bhf_results$bhf_estimate, na.rm = TRUE)
bhf_ci_width_limits <- c(0, max(final_bhf_results$bhf_ci_width, na.rm = TRUE))
bhf_wave_years <- sort(unique(final_bhf_results$time_id))

plot_bhf_estimate <- function(wave) {
  final_bhf_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = bhf_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = bhf_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_bhf_ci_width <- function(wave) {
  final_bhf_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = bhf_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = bhf_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(bhf_wave_years, plot_bhf_estimate)
uncertainty_maps <- lapply(bhf_wave_years, plot_bhf_ci_width)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(bhf_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = length(bhf_wave_years)) +
  plot_annotation(
    title = "Share of adults who cycle for transport frequently",
    subtitle = paste0("Selected model (", bhf_best_model, ") and 95% CI width"),
    theme = map_annotation_theme
  )
```

Although unit-level modelling can stabilise estimates where direct estimates are weak, aggregation using area-level predictor summaries still relies on strong assumptions about population composition. Representing age, sex, and car access using mean adult age, the adult female share, and the adult share with car access can conceal important differences in their joint distribution. Two areas may have similar marginal summaries but very different age–sex–car compositions, leading to similar model-based estimates even when cycling behaviour differs across particular population groups.

This limitation is especially important for a binary outcome modelled using a logistic model (setting `family = "binomial"`). The target is the population-average predicted probability. Substituting area-level predictor summaries to estimate the probability at the average predictor values only is an approximation.

A poststratification table addresses this limitation by providing population counts for the joint combinations of the relevant predictors. This is the logic of **multilevel regression with poststratification (MRP)**: predict the outcome for each population cell, then aggregate the cell predictions using the corresponding population counts so that the final estimate reflects the composition of the target population.


```{r}
#| label: bhf-export
#| code-fold: true

write_csv(final_bhf_results, file.path(base_path, "freqcyc_estimates_bhf.csv"))
```

<!-- Besag + IID diagnostic exports are generated by scripts/fit_besag_effects.R. -->
