---
title: "Unit-level SAE model with poststratification"
subtitle: "A Bayesian spatio-temporal extension of Multilevel Regression with Poststratification (MRP)"
---
<!--
::: {.callout-note title="Theory in brief"}
Unit-level SAE models the individual outcome before aggregating predictions to area-time estimates. For a continuous outcome, the unit-level model can be written as:

$$
Y_{ijt}
=
\alpha + x_{ijt}^{T}\beta + u_j + v_t + \epsilon_{ijt}.
$$

Here, $x_{ijt}$ contains unit-level *categorical* predictors, $u_j$ is an area effect, $v_t$ is a time effect, and $\epsilon_{ijt}$ is residual error. For a binary outcome, the model predicts the probability that $Y_{ijt}=1$.

The fitted model is first evaluated for each population cell $c$, representing one combination of the categorical predictors in $x$.

$$
\hat{\mu}_{jtc}
=
\widehat{\mathbb{E}}(Y \mid j, t, x_{jtc}).
$$

The final area-time estimate is the population-weighted average of those cell predictions:

$$
\hat{\mu}_{jt}
=
\frac{\sum_c N_{jtc}\hat{\mu}_{jtc}}
{\sum_c N_{jtc}},
$$

where $N_{jtc}$ is the population count for cell $c$ in area $j$ and time $t$.
::: -->

```{r}
#| label: mrp-setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)
library(INLA)
library(tibble)
library(purrr)
library(scales)

base_path <- here::here("data")
output_path <- here::here("outputs")
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)

msoa_boundaries <- st_read(
  file.path(base_path, "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(area_id = MSOA21CD) |>
  filter(str_starts(area_id, "W")) |>
  arrange(area_id)

domain_frame <- read_csv(
  file.path(base_path, "clean", "domain_msoayear.csv"),
  show_col_types = FALSE
)

spatial_graph_file <- file.path(base_path, "clean", "msoa_queen.adj")
```

```{r}
#| label: mrp-plot-config
#| echo: false
missing_fill <- "grey85"
share_label <- label_percent(accuracy = 0.1)
pp_label <- label_number(accuracy = 0.1, scale = 100, suffix = " pp")
map_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey30"),
  axis.title.y = element_text(
    size = 12,
    face = "bold",
    angle = 90,
    margin = margin(r = 15)
  ),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 9)
)

map_annotation_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30")
)

chart_text_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5, colour = "grey30"),
  strip.text = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  legend.text = element_text(size = 12)
)
```

## Estimation approach {.unnumbered}

{{< include assets/target-estimand.qmd >}}

Unit-level SAE with poststratification combines a unit-level model fitted to survey microdata with a poststratification table describing the target population. This is the core structure of Multilevel Regression with Poststratification (MRP): model predictions are averaged over population cells.

For this estimand, we model the binary frequent-cycling indicator `freq_cyclist` using `age_band`, `sex`, `caraccess`, and spatial and temporal random effects using a generalised linear mixed model. We then produce predicted cell means for each `age_band` x `sex` x `caraccess` population cell and aggregate them to MSOA-year estimates using population weights. Uncertainty is propagated by repeating this aggregation across posterior draws. 

Here, the poststratification table is sourced from [ONS UK Census 2021](https://www.ons.gov.uk/census), giving the 2021 population count for each `age_band` x `sex` x `caraccess` cell in each MSOA. For the 2023 and 2025 estimands, the model can still estimate different area-time outcomes through the temporal and spatial terms, but the poststratification weights are held fixed. This assumes that local demographic and car-access composition does not change across the survey years. In other words, $N_{jtc} = N_{jc,2021}$ for all three waves.

The unit-level models do not use survey weights in the likelihood. They assume that survey inclusion is independent of frequent cycling after conditioning on the fixed and random effects. Poststratification adjusts only for variables represented in the population table.

## Prepare modelling and poststratification data

There are two inputs when performing unit-level SAE with poststratification:

- `poststrat_data`, the population cells over which predictions are averaged.
- `model_data`, the survey microdata used to fit the model.

It is essential for the unit-level predictors to be identical to the ones present in the poststratification table. Therefore, age is grouped into the same broad Census age bands and sex is also recoded as Male and Female.

```{r}
#| label: mrp-load-poststrat
age_band_levels <- c("16_24", "25_34", "35_49", "50_64", "65_plus")

poststrat_data <- read_csv(
  file.path(base_path, "clean", "poststrat_freqcyc_age_sex_car_21.csv"),
  show_col_types = FALSE
) |>
  transmute(
    area_id,
    msoa_name,
    time_id,
    age_band = factor(age_band, levels = age_band_levels),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    caraccess,
    pop_count
  ) |>
  filter(pop_count > 0) |>
  arrange(time_id, area_id, age_band, sex, caraccess)

head(poststrat_data)
```

```{r}
#| label: mrp-load-survey
survey_microdata <- read_csv(
  file.path(base_path, "clean", "surveyind_freqcyc.csv"),
  show_col_types = FALSE
)

survey_microdata_recoded <- survey_microdata |>
  mutate(
    age_band = cut(
      age,
      breaks = c(16, 25, 35, 50, 65, Inf),
      labels = age_band_levels,
      right = FALSE
    ),
    age_band = factor(age_band, levels = age_band_levels),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))
  )

model_data <- survey_microdata_recoded |>
  select(unit_id, area_id, time_id, freq_cyclist, age_band, sex, caraccess) |>
  drop_na(freq_cyclist, age_band, sex, caraccess)

head(model_data)
```
Before fitting the model, we add the `INLA` time and area indices to the prepared datasets.

```{r}
#| label: mrp-add-ids
#| code-fold: false

area_index <- domain_frame |>
  distinct(area_id, area_inla_id)

time_index <- domain_frame |>
  distinct(time_id, time_inla_id)

model_data <- model_data |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  arrange(time_inla_id, area_inla_id)

poststrat_data <- poststrat_data |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  arrange(time_inla_id, area_inla_id, age_band, sex, caraccess)
```

## Fit and evaluate models

We fit binomial unit-level models to the survey microdata, then use them to predict area-time means. We specify three unit-level models to evaluate at the next step:

| Model | Name | Specification |
|---|---|---|
| M1 | +TempSpatial | Temporal RW1 plus BYM2 spatial MSOA effect |
| M2 | +Demographic | M1 plus age band and sex |
| M3 | +CarAccess | M2 plus respondent car access |

```{r}
#| label: mrp-specs
#| code-fold: false

m1_formula <- freq_cyclist ~
  1 +
  f(
    time_inla_id, 
    model = "rw1", 
    scale.model = TRUE
    ) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE
  )

m2_formula <- update(m1_formula, . ~ . +
  age_band +
  sex
)

m3_formula <- update(m2_formula, . ~ . + caraccess)

model_formulas <- list(
  "M1 +TempSpatial" = m1_formula,
  "M2 +Demographic" = m2_formula,
  "M3 +CarAccess" = m3_formula
)
```

```{r}
#| label: mrp-fit
#| code-fold: false

fit_mrp_model <- function(formula) {
  inla(
    formula = formula,
    family = "binomial",
    data = model_data,
    control.compute = list(waic = TRUE, config = TRUE)
  )
}

mrp_fits <- lapply(model_formulas, fit_mrp_model)
```

Model fit is compared with `WAIC` as an initial check. The fixed-effect table reports the selected model's predictor terms and gives a directional check on whether the modelled relationships align with expected cycling patterns.

```{r}
#| label: mrp-waic

model_waic <- imap_dfr(
  mrp_fits,
  \(fit, model_name) {
    tibble(
      model = model_name,
      waic = fit$waic$waic,
      eff_params = fit$waic$p.eff
    )
  }
) |>
  arrange(waic)

model_waic

mrp_best_model <- model_waic |>
  slice(1) |>
  pull(model)

mrp_final_fit <- mrp_fits[[mrp_best_model]]
cat("Best fitted model spec: ", mrp_best_model)
```
The fitted model summary reports the estimated coefficients and their uncertainty intervals.

```{r}
#| label: mrp-fixed

mrp_final_fit$summary.fixed |>
  as.data.frame() |>
  rownames_to_column("term") |>
  transmute(
    term,
    mean,
    lower = `0.025quant`,
    upper = `0.975quant`
  )
```

## Poststratification with posterior draws

As before, posterior draws from the selected `INLA` fit produce MSOA-year estimates. Each draw provides fixed effects, spatial area effects, and temporal effects. The fitted model is evaluated over every `age_band` x `sex` x `caraccess` population cell, transformed to the response scale, weighted by `pop_count`, and summarised as MSOA-year draws.

:::{.callout-note}
Each posterior draw stores the fixed effects, spatial area effect, and temporal effect together in `INLA`'s latent vector. Fixed effects are labelled as `term:1`, while random effects are labelled as `effect:index`.
:::

```{r}
#| label: mrp-draws
#| code-fold: false

set.seed(100)
posterior_draws <- inla.posterior.sample(
  n = 300,
  result = mrp_final_fit
)

# Fixed-effect draw matrix
fixed_terms <- rownames(mrp_final_fit$summary.fixed)
fixed_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0(fixed_terms, ":1"), 1]),
    fixed_terms
  )
})

# Area-effect draw matrix
area_ids <- sort(area_index$area_inla_id)
area_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("area_inla_id:", area_ids), 1]),
    area_ids
  )
})

# Time-effect draw matrix
time_ids <- sort(time_index$time_inla_id)
time_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("time_inla_id:", time_ids), 1]),
    time_ids
  )
})
```

Each posterior draw is evaluated over every poststratification cell. The cell predictions are then weighted by population count and aggregated within each MSOA-year.

::: {.callout-note title="Why poststratification?"}
Unit-level SAE with poststratification improves on the previous area-summary approach in two ways.

- First, the unit-level model is fitted to survey microdata, but the survey sample may not match the target population composition (see below figure). Poststratification partially adjusts this by predicting the outcome for known population groups and weighting those predictions by the corresponding population counts. This helps align the final estimates with the population structure (but only for the variables included in the poststratification table).

```{r}
#| label: mrp-composition
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| out-width: "100%"

composition_levels <- c(
  paste0("Age: ", c("16-24", "25-34", "35-49", "50-64", "65+")),
  " ",
  "Male",
  "Female",
  "  ",
  "No car",
  "Have car"
)

bind_rows(
  model_data |> mutate(source = "Survey", weight = 1),
  poststrat_data |> mutate(source = "Census", weight = pop_count)
) |>
  mutate(across(c(age_band, sex, caraccess), as.character)) |>
  pivot_longer(c(age_band, sex, caraccess), names_to = "variable", values_to = "category") |>
  count(source, variable, category, wt = weight, name = "count") |>
  mutate(
    proportion = count / sum(count),
    category = recode(
      category,
      "16_24" = "Age: 16-24",
      "25_34" = "Age: 25-34",
      "35_49" = "Age: 35-49",
      "50_64" = "Age: 50-64",
      "65_plus" = "Age: 65+",
      "0" = "No car",
      "1" = "Have car"
    ),
    category = factor(category, levels = composition_levels),
    source = factor(source, levels = c("Survey", "Census")),
    .by = c(source, variable)
  ) |>
  ggplot(aes(x = category, y = proportion, fill = source)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_x_discrete(
    drop = FALSE,
    labels = \(x) if_else(str_trim(x) == "", "", x)
  ) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_fill_manual(
    values = c("Survey" = "#3b7ea1", "Census" = "#d95f02"),
    name = NULL
  ) +
  labs(
    title = "Survey sample vs. Census population",
    x = NULL,
    y = "Proportion"
  ) +
  theme_minimal(base_size = 12) +
  chart_text_theme +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

- Second, for binary outcomes such as `freq_cyclist` (`1` = yes, `0` = no), the target is the population proportion in each area. Poststratification estimates this adequately by averaging predicted probabilities across population cells, rather than plugging a single set of area-level predictor summaries into the model.
:::

```{r}
#| label: mrp-predict
#| code-fold: false

prediction_matrix <- model.matrix(
  ~ age_band + sex + caraccess,
  data = poststrat_data
)[, fixed_terms, drop = FALSE]

cell_draws <- sapply(seq_along(posterior_draws), \(draw_id) {
  cell_predictions <- as.numeric(prediction_matrix %*% fixed_draws[, draw_id]) +
    area_draws[as.character(poststrat_data$area_inla_id), draw_id] +
    time_draws[as.character(poststrat_data$time_inla_id), draw_id]

  # Convert log-odds to probabilities.
  plogis(cell_predictions)
})

# Aggregate and apply poststrat weight for each area-time
area_year_id <- paste(poststrat_data$area_id, poststrat_data$time_id, sep = "__")

aggregate_draws <- function(draw_matrix, weights, group) {
  weighted_sums <- rowsum(draw_matrix * weights, group = group, reorder = FALSE)
  population_totals <- rowsum(weights, group = group, reorder = FALSE)
  weighted_sums / as.numeric(population_totals)
}

prediction_draws <- aggregate_draws(
  draw_matrix = cell_draws,
  weights = poststrat_data$pop_count,
  group = area_year_id
)
```

The matrix `prediction_draws` now has one MSOA-year estimate per row and one posterior draw per column. Posterior means, variances, and 95% credible intervals are computed directly from this draw matrix.

```{r}
#| label: mrp-results
#| code-fold: false

draw_quantile <- \(p) {
  apply(prediction_draws, 1, quantile, probs = p, names = FALSE)
}

final_mrp_results <- tibble(area_year_id = rownames(prediction_draws)) |>
  separate(
    area_year_id,
    into = c("area_id", "time_id"),
    sep = "__",
    convert = TRUE
  ) |>
  bind_cols(
    tibble(
      mrp_estimate = rowMeans(prediction_draws),
      mrp_variance = apply(prediction_draws, 1, var),
      mrp_lower = draw_quantile(0.025),
      mrp_upper = draw_quantile(0.975)
    )
  ) |>
  mutate(
    mrp_ci_width = mrp_upper - mrp_lower
  )

head(final_mrp_results)
```

```{r}
#| label: mrp-map
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

final_mrp_map_data <- msoa_boundaries |>
  left_join(final_mrp_results, by = "area_id")
mrp_estimate_limits <- range(final_mrp_results$mrp_estimate, na.rm = TRUE)
mrp_ci_width_limits <- c(0, max(final_mrp_results$mrp_ci_width, na.rm = TRUE))
mrp_wave_years <- sort(unique(final_mrp_results$time_id))

plot_mrp_estimate <- function(wave) {
  final_mrp_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = mrp_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = mrp_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_mrp_ci_width <- function(wave) {
  final_mrp_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = mrp_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = mrp_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

estimate_maps <- lapply(mrp_wave_years, plot_mrp_estimate)
uncertainty_maps <- lapply(mrp_wave_years, plot_mrp_ci_width)

estimate_maps[[1]] <- estimate_maps[[1]] + ylab("Final estimate")
uncertainty_maps[[1]] <- uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(mrp_wave_years) - 1)) {
  estimate_maps[[i]] <- estimate_maps[[i]] +
    theme(legend.position = "none")
  uncertainty_maps[[i]] <- uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(estimate_maps, uncertainty_maps), ncol = length(mrp_wave_years)) +
  plot_annotation(
    title = "Share of adults who cycle for transport frequently",
    subtitle = "Selected model and 95% credible interval width",
    theme = map_annotation_theme
  )
```

These estimates reuse the 2021 age-sex-caraccess population composition for the 2023 and 2025 waves, while the modelled time effect captures changes in the outcome. This requires the relative `age_band` x `sex` x `caraccess` poststratification weights to stay stable across target periods.

## Appendix: Enhance shallow poststratification with contextual predictors

Sometimes the population table is available for every target year but only at a shallow cell grain. ONS mid-year population estimates can give MSOA-year counts by `age_band` and `sex`, but not by `caraccess`. In that setting, poststratification can still be used, but only for unit-level predictors represented in the population table. 

To recover some between-area variation in `caraccess`, we can add a contextual predictor representing average share of the adult population that has access to a car. However, they do not describe respondent composition within an MSOA, i.e., they cannot represent how cycling behaviour differs between adults with and without car access inside the same age-sex cell.

```{r}
#| label: mrp-load-shallow

shallow_poststrat_data <- read_csv(
  file.path(base_path, "clean", "poststrat_freqcyc_age_sex_212325.csv"),
  show_col_types = FALSE
) |>
  filter(pop_count > 0) |>
  mutate(
    age_band = factor(age_band, levels = age_band_levels),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))
  ) |>
  arrange(time_id, area_id, age_band, sex)

shallow_poststrat_summary <- shallow_poststrat_data |>
  group_by(time_id) |>
  summarise(
    poststrat_cells = n(),
    total_population = sum(pop_count),
    .groups = "drop"
  )

head(shallow_poststrat_data)
shallow_poststrat_summary
```

```{r}
#| label: mrp-shallow-data

shallow_model_data <- survey_microdata_recoded |>
  select(unit_id, area_id, time_id, freq_cyclist, age_band, sex)

head(shallow_model_data)
```

```{r}
#| label: mrp-add-context

area_context_predictors <- readr::read_csv(
  file.path(base_path, "clean", "areapred_freqcyc.csv"),
  show_col_types = FALSE
) |>
  dplyr::transmute(
    area_id,
    z_caraccess_adult_share = as.numeric(scale(caraccess_adult_share))
  )

shallow_model_data <- shallow_model_data |>
  left_join(area_context_predictors, by = "area_id") |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  drop_na(
    freq_cyclist, age_band, sex, z_caraccess_adult_share,
    area_inla_id, time_inla_id
  ) |>
  arrange(time_inla_id, area_inla_id)

shallow_poststrat_data <- shallow_poststrat_data |>
  left_join(area_context_predictors, by = "area_id") |>
  left_join(area_index, by = "area_id") |>
  left_join(time_index, by = "time_id") |>
  drop_na(
    age_band, sex, z_caraccess_adult_share,
    area_inla_id, time_inla_id
  ) |>
  arrange(time_inla_id, area_inla_id, age_band, sex)
```

In the **modelling** step, the model specification uses `age_band` and `sex` as the poststratification-cell predictors, and adds the contextual area predictor directly as fixed effect terms.

```{r}
#| label: mrp-fit-shallow
#| code-fold: false

shallow_formula <- freq_cyclist ~
  f(
    time_inla_id, 
    model = "rw1", 
    scale.model = TRUE
  ) +
  f(
    area_inla_id,
    model = "bym2",
    graph = spatial_graph_file,
    scale.model = TRUE
  ) +
  age_band +
  sex +
  z_caraccess_adult_share

shallow_context_fit <- inla(
  formula = shallow_formula,
  family = "binomial",
  data = shallow_model_data,
  control.predictor = list(compute = TRUE),
  control.compute = list(waic = TRUE, config = TRUE)
)
```

The WAIC comparison suggests that contextual predictors add less to model performance than the richer specification with the unit-level car-access predictor. When survey microdata and poststratification data are available at compatible predictor grains, the richer specification is preferred.

```{r}
#| label: mrp-shallow-waic
rich_shallow_waic <- bind_rows(
  model_waic |>
    filter(model == mrp_best_model) |>
    transmute(
      workflow = "Rich poststratification",
      model = "Age + Sex + CarAccess",
      waic,
      eff_params
    ),
  tibble(
    workflow = "Shallow poststratification",
    model = "Age + Sex (Context: Car access rate)",
    waic = shallow_context_fit$waic$waic,
    eff_params = shallow_context_fit$waic$p.eff
  )
) |>
  arrange(waic)

rich_shallow_waic
```

The shallow **poststratification** step uses the same posterior-draw workflow, but weights fitted cell predictions only by the age-sex population counts for each MSOA-year.

```{r}
#| label: shallow-results

set.seed(101)
shallow_draws <- inla.posterior.sample(
  n = 300,
  result = shallow_context_fit
)

shallow_fixed_terms <- rownames(shallow_context_fit$summary.fixed)
shallow_fixed_draws <- sapply(shallow_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0(shallow_fixed_terms, ":1"), 1]),
    shallow_fixed_terms
  )
})

shallow_area_draws <- sapply(shallow_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("area_inla_id:", area_ids), 1]),
    area_ids
  )
})

shallow_time_draws <- sapply(shallow_draws, \(draw) {
  setNames(
    as.numeric(draw$latent[paste0("time_inla_id:", time_ids), 1]),
    time_ids
  )
})

shallow_prediction_matrix <- model.matrix(
  ~ age_band + sex + z_caraccess_adult_share,
  data = shallow_poststrat_data
)[, shallow_fixed_terms, drop = FALSE]

shallow_cell_draws <- sapply(seq_along(shallow_draws), \(draw_id) {
  shallow_cell_predictions <- as.numeric(
    shallow_prediction_matrix %*% shallow_fixed_draws[, draw_id]
  ) +
    shallow_area_draws[as.character(shallow_poststrat_data$area_inla_id), draw_id] +
    shallow_time_draws[as.character(shallow_poststrat_data$time_inla_id), draw_id]

  # Convert log-odds to probabilities.
  plogis(shallow_cell_predictions)
})

shallow_area_year_id <- paste(
  shallow_poststrat_data$area_id,
  shallow_poststrat_data$time_id,
  sep = "__"
)

shallow_prediction_draws <- aggregate_draws(
  draw_matrix = shallow_cell_draws,
  weights = shallow_poststrat_data$pop_count,
  group = shallow_area_year_id
)

shallow_draw_quantile <- \(p) {
  apply(shallow_prediction_draws, 1, quantile, probs = p, names = FALSE)
}

shallow_mrp_results <- tibble(area_year_id = rownames(shallow_prediction_draws)) |>
  separate(
    area_year_id,
    into = c("area_id", "time_id"),
    sep = "__",
    convert = TRUE
  ) |>
  bind_cols(
    tibble(
      shallow_mrp_estimate = rowMeans(shallow_prediction_draws),
      shallow_mrp_variance = apply(shallow_prediction_draws, 1, var),
      shallow_mrp_lower = shallow_draw_quantile(0.025),
      shallow_mrp_upper = shallow_draw_quantile(0.975)
    )
  ) |>
  mutate(
    shallow_mrp_ci_width = shallow_mrp_upper - shallow_mrp_lower
  )

head(shallow_mrp_results)
```

```{r}
#| label: mrp-map-shallow
#| echo: false
#| fig-width: 16
#| fig-height: 16
#| out-width: "100%"

shallow_context_map_data <- msoa_boundaries |>
  left_join(shallow_mrp_results, by = "area_id")
shallow_estimate_limits <- range(shallow_mrp_results$shallow_mrp_estimate, na.rm = TRUE)
shallow_ci_width_limits <- c(0, max(shallow_mrp_results$shallow_mrp_ci_width, na.rm = TRUE))

plot_shallow_estimate <- function(wave) {
  shallow_context_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = shallow_mrp_estimate), colour = "transparent") +
    scale_fill_viridis_c(
      option = "magma",
      limits = shallow_estimate_limits,
      labels = share_label,
      na.value = missing_fill,
      name = "Share"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

plot_shallow_ci_width <- function(wave) {
  shallow_context_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = shallow_mrp_ci_width), colour = "transparent") +
    scale_fill_distiller(
      palette = "Oranges",
      direction = 1,
      limits = shallow_ci_width_limits,
      labels = pp_label,
      na.value = missing_fill,
      name = "95% CI width"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

shallow_estimate_maps <- lapply(mrp_wave_years, plot_shallow_estimate)
shallow_uncertainty_maps <- lapply(mrp_wave_years, plot_shallow_ci_width)

shallow_estimate_maps[[1]] <- shallow_estimate_maps[[1]] + ylab("Final estimate")
shallow_uncertainty_maps[[1]] <- shallow_uncertainty_maps[[1]] + ylab("95% CI width")

for (i in seq_len(length(mrp_wave_years) - 1)) {
  shallow_estimate_maps[[i]] <- shallow_estimate_maps[[i]] +
    theme(legend.position = "none")
  shallow_uncertainty_maps[[i]] <- shallow_uncertainty_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(c(shallow_estimate_maps, shallow_uncertainty_maps), ncol = length(mrp_wave_years)) +
  plot_annotation(
    title = "Shallow poststratification",
    subtitle = "Age-sex poststratification with MSOA car access rate as contextual",
    theme = map_annotation_theme
  )
```

Compared with the richer three-predictor poststratified model, the local differences show whether estimates are driven more by detailed population composition or broader area context. For example, the rich model may lower the estimate for an MSOA if its poststratification cells contain more residents with car access than the flat area-level contextual predictor implies, and vice versa.

```{r}
#| label: mrp-map-rich-shallow
#| echo: false
#| fig-width: 16
#| fig-height: 6
#| out-width: "100%"

rich_minus_shallow_results <- final_mrp_results |>
  select(
    area_id,
    time_id,
    rich_mrp_estimate = mrp_estimate
  ) |>
  left_join(
    shallow_mrp_results |>
      select(area_id, time_id, shallow_mrp_estimate),
    by = c("area_id", "time_id")
  ) |>
  mutate(rich_minus_shallow_mrp = rich_mrp_estimate - shallow_mrp_estimate)

rich_minus_shallow_limit <- quantile(
  abs(rich_minus_shallow_results$rich_minus_shallow_mrp),
  0.95,
  na.rm = TRUE,
  names = FALSE
)
rich_minus_shallow_limits <- c(-rich_minus_shallow_limit, rich_minus_shallow_limit)

rich_minus_shallow_map_data <- msoa_boundaries |>
  left_join(rich_minus_shallow_results, by = "area_id")

plot_rich_minus_shallow <- function(wave) {
  rich_minus_shallow_map_data |>
    filter(time_id == wave) |>
    ggplot() +
    geom_sf(aes(fill = rich_minus_shallow_mrp), colour = "transparent") +
    scale_fill_gradient2(
      low = "#2166ac",
      mid = "white",
      high = "#b2182b",
      midpoint = 0,
      limits = rich_minus_shallow_limits,
      oob = squish,
      labels = pp_label,
      na.value = missing_fill,
      name = "Rich - Shallow"
    ) +
    ggtitle(wave) +
    theme_void() +
    map_text_theme
}

rich_minus_shallow_maps <- lapply(mrp_wave_years, plot_rich_minus_shallow)
rich_minus_shallow_maps[[1]] <- rich_minus_shallow_maps[[1]] + ylab("Rich - shallow")

for (i in seq_len(length(mrp_wave_years) - 1)) {
  rich_minus_shallow_maps[[i]] <- rich_minus_shallow_maps[[i]] +
    theme(legend.position = "none")
}

wrap_plots(rich_minus_shallow_maps, ncol = 3) +
  plot_annotation(
    title = "Difference in estimates between models with rich and shallow poststratification",
    theme = map_annotation_theme
  )
```
```{r}
#| label: mrp-export
#| code-fold: true

posterior_hyperparameter_names <- names(posterior_draws[[1]]$hyperpar)
posterior_hyperparameter_draws <- sapply(posterior_draws, \(draw) {
  setNames(
    as.numeric(draw$hyperpar[posterior_hyperparameter_names]),
    posterior_hyperparameter_names
  )
})

saveRDS(
  list(
    schema_version = 1L,
    route = "Unit-level SAE with poststratification",
    model = mrp_best_model,
    seed = 100L,
    n_draws = length(posterior_draws),
    fixed = fixed_draws,
    area = area_draws,
    time = time_draws,
    hyperparameters = posterior_hyperparameter_draws,
    area_index = area_index,
    time_index = time_index
  ),
  file.path(output_path, "freqcyc_posterior_draws_unit_mrp.rds"),
  compress = "xz"
)

write_csv(final_mrp_results, file.path(output_path, "freqcyc_estimates_mrp_rich.csv"))
write_csv(shallow_mrp_results, file.path(output_path, "freqcyc_estimates_mrp_shallow.csv"))
```
