---
title: "Tutorial: Direct Estimation"
---

Direct estimation is the process of deriving area-level estimates using only the survey data. The direct estimates and their sampling variances are also key inputs for the Fay-Herriot area-level estimation in the next step.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Scenario | Estimator | Formula | When to use |
|---|---|---|---|
| Known domain sizes; has grossing weights | Horvitz-Thompson estimator | $\hat{\theta}_d^{DIR}=\frac{\sum_{i\in s_d}w^G_{di}y_{di}}{N_d}$ | Anchors estimates to the true population size of each estimation area; required for counts/totals and useful for population rates. |
| Unknown domain sizes; has survey weights | Hájek estimator | $\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}w_{di}^Ay_{di}}{\sum_{i \in s_d}w^A_{di}}$ | Used for means or proportions when analysis weights are available but reliable area population sizes are not. |
| Unknown domain sizes; no survey weights | Unweighted sample mean | $\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}y_{di}}{n_d}$ | The simplest direct estimator, but likely biased if survey sampling is not random and imprecise when $n_d$ is too low. |

Here, the direct estimate $\hat{\theta}_d^{DIR}$ is for area $d$ with sample size $n_d$ and domain size $N_d$ if known.

Direct estimates are paired with **sampling variances**, which describe how much uncertainty comes from observing only a sample within each estimation area. The variance depends on the estimator, the weight scale, and the available survey design information.

For easier interpretation, we usually convert sampling variance into:

- **Coefficient of Variation (CV):** the standard error divided by the estimate. Values above 1 indicate that the typical standard error is larger than the estimate.
- **Confidence Interval Width:** the approximate width of the 95% confidence interval, expressed in the original units.

## Evaluate Direct Estimates

Use direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Too many areas have no or very few samples; variances are missing or unreasonably large | Revisit the estimand definition, target population, or estimation geography. |
| All areas have direct estimates with acceptable sample sizes and uncertainty | Direct estimates may be enough for reporting. |
| Most areas have direct estimates, but many are imprecise | Fit an area-level model to borrow strength from covariates and neighbouring areas. |

## Practical tutorial

In this workbook, we calculate direct estimates for **three example estimands** at MSOA level:

- mean commute distance among commuters,
- share of recent bus riders who are dissatisfied with bus services, and
- number of frequent walkers per 1,000 residents.

The examples use the `survey` package which can deal with survey microdata that come with weights and strata.

```{r}
# Load libraries
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)

# Load analysis-ready data into a temp folder
zip_tmp <- tempfile(fileext = ".zip")
url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-clean-data.zip"
download.file(url, zip_tmp, mode = "wb")
unzip(zip_tmp, exdir = tempdir())
base_path <- file.path(tempdir())

# Load shapefiles for plotting
boundary_zip <- tempfile(fileext = ".zip")
boundary_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-boundaries.zip"
download.file(boundary_url, boundary_zip, mode = "wb")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

msoa_domains <- msoa_boundaries |>
  st_drop_geometry() |>
  select(domain_id)
```

### Mean commute distance among commuters by MSOA
The indicator `commute_distance` is continuous, so the direct estimate is simply the weighted mean commute distance among available commuter respondents.

#### Direct Estimation

```{r}
#| code-fold: false
# Load the indicator and weight
survey_indicator_commute <- read_csv(
  file.path(base_path, "ind_weight_commute_distance.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator),
    !is.na(weight),
    !is.na(domain_id),
    !is.na(strata)
  )

# Set up the survey design
design_commute <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_commute
)

# Estimate the weighted mean in each MSOA
commute_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_commute,
  FUN = survey::svymean,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Output with uncertainty measures
dir_est_commute <- commute_mean |>
  transmute(
    domain_id,
    direct = indicator,
    variance = se^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_commute)
```
#### Evaluation

```{r}
commute_map_data <- msoa_boundaries |>
  left_join(
    dir_est_commute,
    by = "domain_id"
  )

commute_estimate_map <- ggplot(commute_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.05) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey85") +
  labs(title = "Direct estimate", fill = "Estimate") +
  theme_void()

commute_ci_map <- ggplot(commute_map_data) +
  geom_sf(aes(fill = ci_width), colour = "white", linewidth = 0.05) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey85") +
  labs(title = "CI width", fill = "CI width") +
  theme_void()

commute_estimate_map + commute_ci_map +
  plot_annotation(
    title = "Mean commute distance among commuters by MSOA"
  )
```

```{r}
commute_sample_n <- survey_indicator_commute |>
  count(domain_id, name = "n")

msoa_domains |>
  left_join(commute_sample_n, by = "domain_id") |>
  left_join(dir_est_commute, by = "domain_id") |>
  mutate(n = replace_na(n, 0)) |>
  summarise(
    areas_total = n(),
    areas_n_0 = sum(n == 0),
    areas_n_1 = sum(n == 1),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

The direct estimates are relatively stable: most MSOAs have at least some usable commuter observations. Even so, the confidence intervals remain fairly wide in practical terms, so the direct estimates should still be treated as noisy area-level summaries.

### Dissatisfied bus users among recent adult bus riders by MSOA

The indicator `bus_nosat` is binary. It is defined only among adults who rode a bus in the last 12 months: `1` means dissatisfied, `0` means not dissatisfied, and non-riders are left as nulls. The direct estimate is the weighted proportion among available recent bus riders.

#### Direct Estimation

```{r}
#| code-fold: false
# Load the indicator and weight
survey_indicator_bus <- read_csv(
  file.path(base_path, "ind_weight_bus_nosat.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator),
    !is.na(weight),
    !is.na(domain_id),
    !is.na(strata)
  )

# Set up the survey design
design_bus <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_bus
)

# Estimate the weighted share in each MSOA
bus_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_bus,
  FUN = survey::svymean, # when paired with binary indicator, it represents share
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Output with uncertainty measures
dir_est_bus <- bus_mean |>
  transmute(
    domain_id,
    direct = indicator,
    variance = se^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_bus)
```
#### Evaluation
```{r}
bus_map_data <- msoa_boundaries |>
  left_join(
    dir_est_bus,
    by = "domain_id"
  )

bus_estimate_map <- ggplot(bus_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.05) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey85") +
  labs(title = "Direct estimate", fill = "Estimate") +
  theme_void()

bus_ci_map <- ggplot(bus_map_data) +
  geom_sf(aes(fill = ci_width), colour = "white", linewidth = 0.05) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey85") +
  labs(title = "CI width", fill = "CI width") +
  theme_void()

bus_estimate_map + bus_ci_map +
  plot_annotation(
    title = "Dissatisfied recent adult bus riders by MSOA"
  )
```

```{r}
#| echo: false
bus_sample_n <- survey_indicator_bus |>
  count(domain_id, name = "n")

msoa_domains |>
  left_join(bus_sample_n, by = "domain_id") |>
  left_join(dir_est_bus, by = "domain_id") |>
  mutate(n = replace_na(n, 0)) |>
  summarise(
    areas_total = n(),
    areas_n_0 = sum(n == 0),
    areas_n_1 = sum(n == 1),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

The direct estimates have reasonable geographic coverage, but they remain imprecise because the estimand is restricted to adults who rode a bus in the last 12 months. This smaller eligible group reduces the effective sample size within many MSOAs, which is reflected in the CI-width map and summary diagnostics. Model-based estimation is likely to produce more stable local estimates.

### Frequent walkers per 1,000 residents by MSOA

`freq_walker` is a binary survey indicator for whether a respondent walks for transport daily. The direct estimate for each area is therefore to be calculated as a weighted total of this indicator (i.e., total number of frequent walkers in the MSOA), divided by a domain size (i.e., total population by MSOA), and multiple by 1,000.

For a standard **Horvitz-Thompson rate estimator**, we would usually need grossing weights $w^G$ and known domain sizes $N_d$.

$$
\hat{\theta}_d^{DIR}=\frac{\sum_iw^G_{di}y_{di}}{N_d}
$$

 Here, we can use a simpler equivalent approach. `freq_walker` uses `SampleTravelWeight`, an analysis weight designed for estimating total population proportions. If a grossing weight is just this analysis weight rescaled so that they sum to the domain population, or $w^G_{di} = w^A_{di} \times \frac{N_d}{\sum_{i \in d} w^A_{di}}$, then the weighted total divided by the domain size is equal to the analysis-weighted mean:
$$
\hat{\theta}_d^{DIR}
=
\frac{\sum_i w^G_{di}y_{di}}{N_d}
=
\frac{\sum_i w^A_{di}y_{di}}{\sum_i w^A_{di}}.
$$

:::{.callout-warning}
This approach only works when the analysis weight $w^A$ and domain size $N_d$ refer to the same target population; here, that is the total population by MSOA.

If they do not, for example when adult-population domain sizes are needed, use the survey-provided adult-population grossing weights $w^G$ and the corresponding adult population by MSOA as the domain sizes $N_d$.
:::

#### Direct Estimation

```{r}
#| code-fold: false
# Load the active-travel indicator with its analysis weight
survey_indicator_walk <- read_csv(
  file.path(base_path, "ind_weight_freq_walker.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator),
    !is.na(weight),
    !is.na(domain_id),
    !is.na(strata)
  )

# Set up the survey design
design_walk <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_walk
)

# Estimate the weighted mean in each MSOA
walk_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_walk,
  FUN = survey::svymean,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Convert the weighted proportion to a rate per 1,000 residents
dir_est_walk <- walk_mean |>
  transmute(
    domain_id,
    direct = indicator * 1000,
    variance = (se  * 1000)^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_walk)
```

#### Evaluation

```{r}
# Plot estimate and CI width
walk_map_data <- msoa_boundaries |>
  left_join(dir_est_walk, by = "domain_id")

walk_estimate_map <- ggplot(walk_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.05) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey85") +
  labs(title = "Direct estimate", fill = "Estimate") +
  theme_void()

walk_ci_map <- ggplot(walk_map_data) +
  geom_sf(aes(fill = ci_width), colour = "white", linewidth = 0.05) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey85") +
  labs(title = "CI width", fill = "CI width") +
  theme_void()

walk_estimate_map + walk_ci_map +
  plot_annotation(
    title = "Residents walking daily as main transport mode per 1,000 residents by MSOA"
  )
```

```{r}
# Uncertainty diagnostics summary
walk_sample_n <- survey_indicator_walk |>
  count(domain_id, name = "n")

msoa_domains |>
  left_join(walk_sample_n, by = "domain_id") |>
  left_join(dir_est_walk, by = "domain_id") |>
  mutate(n = replace_na(n, 0)) |>
  summarise(
    areas_total = n(),
    areas_n_0 = sum(n == 0),
    areas_n_1 = sum(n == 1),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

The direct estimates appear weak, with many areas showing abnormally high estimates with wide confidence intervals. The diagnostics confirm that this is driven by sparse subsampled active-travel module data: several MSOAs have no usable observations, more have fewer than two observations, and the median CI width is high. This makes it a strong candidate for model-based estimation, where covariates and spatial structure can help stabilise estimates for areas with little direct survey evidence.

### Final thoughts
Direct estimation is the necessary statistical summary to stress-test the raw survey evidence. It shows how much information we have in each area, how uncertain the estimates are, and whether direct reporting is defensible.

The **direct estimate calculation** should follow the measure and the estimand:

- For the estimand based on `commute_distance`, the estimate is the **mean** commute distance among commuter respondents.
- For the estimand based on `bus_nosat`, the estimate is the **share** of recent adult bus riders who are dissatisfied with bus services.
- For the estimand based on `freq_walker`, the estimate is a **population rate**, i.e., the number of frequent walkers per 1,000 residents.

The **uncertainty diagnostics** above show different levels of uncertainty across the three examples. Direct estimates may be enough when sample sizes are adequate and confidence intervals are narrow enough for the intended use. 
