---
title: "Data Prep for Auxiliary Data"
---

Auxiliary data are the non-survey inputs that let small-area models borrow strength and project survey relationships into the full population. In this tutorial, we prepare three reusable auxiliary inputs:

- area-level covariates for Fay-Herriot models;
- population cells for multilevel regression and poststratification;
- a lightweight synthetic population for unit-level prediction examples.

The same discipline used for survey microdata also applies here: the geography, target population, reference period, and variable definitions must align with the estimand.

## Load Census and Boundary Data

The tutorial uses bundled Census 2021 aggregate tables and MSOA boundaries. The Census files are stored as topic tables, so we define one helper for reading a table and keeping Welsh MSOAs only.

```{r}
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)

# Load Census files into a temp folder
census_zip_tmp <- tempfile(fileext = ".zip")
census_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-census-data.zip"
download.file(census_url, census_zip_tmp, mode = "wb")
unzip(census_zip_tmp, exdir = tempdir())
census_path <- file.path(tempdir())

# Load shapefiles for checking domain coverage
boundary_zip <- tempfile(fileext = ".zip")
boundary_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-boundaries.zip"
download.file(boundary_url, boundary_zip, mode = "wb")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

msoa_domains <- msoa_boundaries |>
  st_drop_geometry() |>
  select(domain_id) |>
  arrange(domain_id)

read_census <- function(topic_code) {
  file_name <- paste0("census2021-", topic_code, "-msoa.csv")
  read_csv(file.path(census_path, file_name), show_col_types = FALSE) |>
    filter(str_starts(`geography code`, "W")) |>
    rename(domain_id = `geography code`)
}
```

## Area-Level Covariates

For `commute_distance`, useful area-level covariates should capture urban-rural structure, commuting mode, car access, housing form, and the local employment mix. [The Census 2021](https://www.nomisweb.co.uk/sources/census_2021_bulk) gives us several candidate covariates at MSOA level.

| Covariate | Census table | Why it may explain commute distance |
|---|---|---|
| `cov_pop_density` | TS006 | Dense urban areas tend to have shorter trips and more nearby jobs/services. |
| `cov_share_work_home` | TS061 | Areas with more home-working may have a different commuter composition. |
| `cov_share_hh_withcar` | TS045 | Household car access indicates ability to travel longer distances. |
| `cov_share_detached` | TS044 | Detached housing is a proxy for lower-density suburban/rural settlement. |
| `cov_share_higher_managerial` | TS062 | Occupational structure may reflect access to different labour markets. |

:::{.callout-note}
We do not use Census commute-distance table `TS058` as a covariate because it is reserved later as an external validation benchmark for this tutorial.
:::

It is also good practice to standardise covariates so the model coefficients are on a more comparable scale.

```{r}
# Population density (TS006)
msoa_density <- read_census("ts006") |>
  transmute(
    domain_id,
    cov_pop_density = `Population Density: Persons per square kilometre; measures: Value`
  )

# Travel to work (TS061)
msoa_travel <- read_census("ts061") |>
  transmute(
    domain_id,
    commute_total = `Method of travel to workplace: Total: All usual residents aged 16 years and over in employment the week before the census`,
    cov_share_work_home = if_else(
      commute_total > 0,
      `Method of travel to workplace: Work mainly at or from home` / commute_total,
      NA_real_
    )
  ) |>
  select(-commute_total)

# Car availability (TS045)
msoa_cars <- read_census("ts045") |>
  transmute(
    domain_id,
    household_total = `Number of cars or vans: Total: All households`,
    cov_share_hh_withcar = if_else(
      household_total > 0,
      (household_total - `Number of cars or vans: No cars or vans in household`) / household_total,
      NA_real_
    )
  ) |>
  select(-household_total)

# Housing type (TS044)
msoa_housing <- read_census("ts044") |>
  transmute(
    domain_id,
    household_total = `Accommodation type: Total: All households`,
    cov_share_detached = if_else(
      household_total > 0,
      `Accommodation type: Detached` / household_total,
      NA_real_
    )
  ) |>
  select(-household_total)

# Occupation structure (TS062)
msoa_nssec <- read_census("ts062") |>
  transmute(
    domain_id,
    nssec_total = `National Statistics Socio-economic Classification (NS-SEC): Total: All usual residents aged 16 years and over`,
    cov_share_higher_managerial = if_else(
      nssec_total > 0,
      `National Statistics Socio-economic Classification (NS-SEC): L1, L2 and L3 Higher managerial, administrative and professional occupations` / nssec_total,
      NA_real_
    )
  ) |>
  select(-nssec_total)

area_covariates_commute <- msoa_domains |>
  left_join(msoa_density, by = "domain_id") |>
  left_join(msoa_travel, by = "domain_id") |>
  left_join(msoa_cars, by = "domain_id") |>
  left_join(msoa_housing, by = "domain_id") |>
  left_join(msoa_nssec, by = "domain_id")

area_covariates_commute_scaled <- area_covariates_commute |>
  transmute(
    domain_id,
    cov_pop_density = as.numeric(scale(cov_pop_density)),
    cov_share_work_home = as.numeric(scale(cov_share_work_home)),
    cov_share_hh_withcar = as.numeric(scale(cov_share_hh_withcar)),
    cov_share_detached = as.numeric(scale(cov_share_detached)),
    cov_share_higher_managerial = as.numeric(scale(cov_share_higher_managerial))
  )

head(area_covariates_commute_scaled)
```

## Synthetic Population Cells

Population cells are cross-tabulated counts for combinations of predictors. They are the projection data used in multilevel regression and poststratification: the model predicts an outcome for each cell, then cell predictions are weighted by `pop_count` and aggregated to the estimation area.

Here we prepare a compact MSOA-by-age-by-sex cell table from Census 2021 marginal tables.

:::{.callout-note}
Because only marginal tables are bundled, this example uses iterative-proportional-style assumptions within each MSOA rather than a full Census cross-tabulation. In a production workflow, use an observed cross-tabulated table where available.
:::

```{r}
age_cells <- read_census("ts007a") |>
  transmute(
    domain_id,
    age_0_15 =
      `Age: Aged 4 years and under` +
      `Age: Aged 5 to 9 years` +
      `Age: Aged 10 to 14 years`,
    age_16_24 =
      `Age: Aged 15 to 19 years` +
      `Age: Aged 20 to 24 years`,
    age_25_44 =
      `Age: Aged 25 to 29 years` +
      `Age: Aged 30 to 34 years` +
      `Age: Aged 35 to 39 years` +
      `Age: Aged 40 to 44 years`,
    age_45_64 =
      `Age: Aged 45 to 49 years` +
      `Age: Aged 50 to 54 years` +
      `Age: Aged 55 to 59 years` +
      `Age: Aged 60 to 64 years`,
    age_65_plus =
      `Age: Aged 65 to 69 years` +
      `Age: Aged 70 to 74 years` +
      `Age: Aged 75 to 79 years` +
      `Age: Aged 80 to 84 years` +
      `Age: Aged 85 years and over`
  ) |>
  pivot_longer(
    cols = starts_with("age_"),
    names_to = "age_group",
    values_to = "age_count"
  )

sex_cells <- read_census("ts008") |>
  transmute(
    domain_id,
    sex_female = `Sex: Female; measures: Value`,
    sex_male = `Sex: Male; measures: Value`
  ) |>
  pivot_longer(
    cols = starts_with("sex_"),
    names_to = "sex",
    values_to = "sex_count"
  )

domain_sizes <- read_census("ts001") |>
  transmute(
    domain_id,
    domain_size = `Residence type: Total; measures: Value`
  )

population_cells <- age_cells |>
  left_join(sex_cells, by = "domain_id") |>
  left_join(domain_sizes, by = "domain_id") |>
  mutate(
    pop_count = round(age_count * sex_count / domain_size),
    cell_id = paste(domain_id, age_group, sex, sep = "_")
  ) |>
  select(cell_id, domain_id, age_group, sex, pop_count) |>
  filter(pop_count > 0)

head(population_cells)
```

## Synthetic Population

Some unit-level models need one row per target population unit rather than one row per cell. A synthetic population expands the cell table into person-like records that preserve the cell counts and carry the same covariates used by the unit-level model.

[wip]

```{r}
#| echo: false
output_dir <- file.path("..", "data")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  area_covariates_commute_scaled,
  file.path(output_dir, "bayesian_area_sae_commute_covariates.csv")
)

write_csv(
  domain_sizes,
  file.path(output_dir, "domain_sizes_msoa.csv")
)

write_csv(
  population_cells,
  file.path(output_dir, "population_cells_age_sex.csv")
)

# write_csv(
#   synthetic_population,
#   file.path(output_dir, "synthetic_population_age_sex.csv")
# )

auxiliary_data_files <- file.path(
  output_dir,
  c(
    "bayesian_area_sae_commute_covariates.csv",
    "domain_sizes_msoa.csv",
    "population_cells_age_sex.csv"
    # "synthetic_population_age_sex.csv"
  )
)

zip(
  zipfile = file.path(output_dir, "tutorial-auxiliary-data.zip"),
  files = auxiliary_data_files,
  flags = "-j"
)
```
