---
title: "[WIP] Frequentist Approach to SAE"
---

Frequentist small area estimation is a model-based approach where  estimation is done using purely the observed data, without incroporating priors and other assumptions seen in the Bayesian approach. In it, unknown model quantities are estimated as fixed but unknown parameters, as opposed to a probability distribution.

## Pros

Frequentist SAE is often preferred when the priority is a transparent, repeatable production workflow.

- **Familiar reporting:** outputs such as point estimates, standard errors, RMSEs, and confidence intervals are familiar.
- **Production-friendly:** frequentist models can often be fitted quickly and repeatedly across many indicators.
- **Lower computational burden:** many frequentist SAE models are lighter to run than fully Bayesian models.

## Cons

Frequentist SAE can be less suitable when uncertainty communication or prediction for no-sample areas is central.

- **Less intuitive uncertainty:** confidence intervals and RMSEs can be harder to explain than credible intervals or posterior probabilities.
- **Out-of-sample areas need caution:** areas with no direct estimate may require separate synthetic prediction or a model that explicitly handles missing direct estimates.

## Practical Tutorial

Here, we fit a simple frequentist Fay-Herriot model for `commute_distance`, then compare the resulting EBLUPs with the direct estimates and the Bayesian area-level estimates.

```{r}
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(survey)
library(sae)

clean_zip <- file.path("..", "data", "tutorial-clean-data.zip")
unzip(clean_zip, exdir = tempdir())
base_path <- file.path(tempdir())

boundary_zip <- file.path("..", "data", "tutorial-boundaries.zip")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

# Load Bayesian area-level estimates exported from the previous workbook
bayesian_est_path <- file.path("..", "data", "commute_estimates_fh_bayesian.csv")

bayesian_area_estimates <- read_csv(
  bayesian_est_path,
  show_col_types = FALSE
)

# Load the standardised covariates used in the Bayesian FH model
covariate_path <- file.path("..", "data", "aux_area_covar_commute.csv")

commute_covariates <- read_csv(
  covariate_path,
  show_col_types = FALSE
)
```

First, recreate the **direct estimates** and sampling variances.

```{r}
survey_indicator_commute <- read_csv(
  file.path(base_path, "ind_weight_commute_distance.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator),
    !is.na(weight),
    !is.na(domain_id),
    !is.na(strata)
  ) |>
  dplyr::select(
    unit_id,
    domain_id,
    strata,
    indicator,
    weight
  )

design_commute <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_commute
)

commute_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_commute,
  FUN = survey::svymean,
  vartype = "se",
  na.rm = TRUE
) |>
  tibble::as_tibble()

direct_est_commute <- commute_mean |>
  transmute(
    domain_id,
    direct_estimate = indicator,
    variance = se^2,
    direct_ci_lower = direct_estimate - 1.96 * sqrt(variance),
    direct_ci_upper = direct_estimate + 1.96 * sqrt(variance),
    direct_ci_width = direct_ci_upper - direct_ci_lower
  )
```

Next, fit a covariate-based Fay-Herriot model using the `sae` package. The covariates are the same standardised MSOA-level covariates used in the Bayesian FH model.

```{r}
fh_input <- direct_est_commute |>
  left_join(commute_covariates, by = "domain_id") |>
  filter(!is.na(direct_estimate), !is.na(variance), variance > 0) |>
  as.data.frame()

freq_fh_result <- sae::mseFH(
  direct_estimate ~ cov_pop_density +
    cov_share_work_home +
    cov_share_hh_withcar +
    cov_share_detached +
    cov_share_higher_managerial,
  vardir = variance,
  method = "REML",
  data = fh_input
)

freq_fh_estimates <- fh_input |>
  transmute(
    domain_id,
    freq_fh_estimate = as.numeric(freq_fh_result$est$eblup),
    freq_fh_mse = as.numeric(freq_fh_result$mse),
    freq_fh_ci_lower = freq_fh_estimate - 1.96 * sqrt(freq_fh_mse),
    freq_fh_ci_upper = freq_fh_estimate + 1.96 * sqrt(freq_fh_mse),
    freq_fh_ci_width = freq_fh_ci_upper - freq_fh_ci_lower
  )
```

Compare the direct estimate, frequentist FH EBLUP, and Bayesian FH estimate. MSOAs are ranked by the direct estimate.

```{r}
commute_compare <- bayesian_area_estimates |>
  dplyr::select(
    domain_id,
    direct_estimate,
    direct_ci_lower,
    direct_ci_upper,
    fh_estimate,
    fh_ci_lower,
    fh_ci_upper
  ) |>
  left_join(freq_fh_estimates, by = "domain_id")

ranked_commute <- commute_compare |>
  filter(!is.na(direct_estimate)) |>
  arrange(freq_fh_estimate, domain_id) |>
  mutate(direct_rank = row_number())

ggplot(ranked_commute, aes(x = direct_rank)) +
  geom_ribbon(aes(ymin = freq_fh_ci_lower, ymax = freq_fh_ci_upper), fill = "red", alpha = 0.12) +
  geom_ribbon(aes(ymin = fh_ci_lower, ymax = fh_ci_upper), fill = "blue", alpha = 0.12) +
  geom_line(aes(y = freq_fh_estimate, colour = "Frequentist FH"), linewidth = 0.45, na.rm = TRUE) +
  geom_line(aes(y = fh_estimate, colour = "Bayesian FH"), linewidth = 0.45, na.rm = TRUE) +
  scale_colour_manual(values = c(
    "Frequentist FH" = "red",
    "Bayesian FH" = "blue"
  )) +
  labs(
    title = "Frequentist FH, and Bayesian FH estimates",
    x = "MSOAs (ranked by FH estimate)",
    y = "Mean commute distance (miles)",
    colour = NULL
  ) +
  theme_minimal()
```

Map the same three estimates to compare their spatial patterns.

```{r}
commute_compare_map <- msoa_boundaries |>
  left_join(commute_compare, by = "domain_id") |>
  dplyr::select(freq_fh_estimate, fh_estimate, geometry) |>
  rename(
    `Frequentist FH` = freq_fh_estimate,
    `Bayesian FH` = fh_estimate
  ) |>
  pivot_longer(
    cols = c(`Frequentist FH`, `Bayesian FH`),
    names_to = "method",
    values_to = "estimate"
  )

ggplot(commute_compare_map) +
  geom_sf(aes(fill = estimate), colour = "white", linewidth = 0.05) +
  facet_wrap(~method) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey85") +
  labs(
    title = "Mean commute distance by MSOA",
    fill = "Miles"
  ) +
  theme_void()
```
